### 2017-03-16 articles/chemical-tagging/ms.tex
Get Travis to compile articles

%% This document is part of the snob project
%% Copyright 2017 the authors. All rights reserved.

\documentclass{aastex61}

% For revision history
\IfFileExists{vc.tex}{\input{vc.tex}}{
    \newcommand{\githash}{UNKNOWN}
    \newcommand{\giturl}{UNKNOWN}}

% Define commands
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}

\received{}
\revised{}
\accepted{}

\submitjournal{AAS Journals}

\shorttitle{Chemical tagging can work, but will it?}
\shortauthors{Casey et al.}

\begin{document}

\title{Chemical tagging can work, but will it?}

\correspondingauthor{Andrew R. Casey}
\email{andrew.casey@monash.edu}

\author[0000-0003-0174-0564]{Andrew R. Casey}
\affil{Monash Centre for Astrophysics, Monash University, Melbourne, 
	   VIC 3800, Australia}

\author{(some order of: John Lattanzio, David Dowe, Aldeida Aleti)}

\author{others?}

\begin{abstract}
Snobbery.
\end{abstract}

\keywords{}

\section{Introduction} 
\label{sec:introduction}

\acknowledgments

\software{astropy \citep{Robitaille:2013}, numpy, scipy, scikit-learn}

\begin{thebibliography}{}

\bibitem[Astropy Collaboration et al.(2013)]{Robitaille:2013} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., et al.\ 2013, \aap, 558, A33

\end{thebibliography}

\end{document}

### 2017-03-16 .ci/travis.sh

@@ -1,0 +1,15 @@
+#!/bin/bash -x
+
+# Build the paper(s)
+cd articles/chemical-tagging
+make
+
+# Push to GitHub
+if [ -n "$GITHUB_API_KEY" ]; then
+  cd $TRAVIS_BUILD_DIR
+  git checkout --orphan $TRAVIS_BRANCH-pdf
+  git rm -rf .
+  git add -f articles/chemical-tagging/ms.pdf
+  git -c user.name='travis' -c user.email='travis' commit -m "Compiling PDF"
+  git push -q -f https://$GITHUB_USER:$GITHUB_API_KEY@github.com/$TRAVIS_REPO_SLUG $TRAVIS_BRANCH-pdf
+fi

### 2017-03-16 snob/gaussian.py

@@ -120,7 +120,7 @@ def _message_length(parameters, y, yerr, bounds, quantum, gradient=False):
     if not gradient:
         return I
 
-    raise NotImplementedError
+    raise NotImplementedError("TODO: put in analytic derivatives")
 
 
 class GaussianEstimator(estimator.Estimator):
@@ -145,7 +145,8 @@ class GaussianEstimator(estimator.Estimator):
         The upper bound for the sigma value.
 
     :param quantum: [optional]
-        The acceptable rounding-off quantum for the minimum message length.
+        The acceptable rounding-off quantum for the minimum message length
+        in units of nits. Default is 0.1 nit.
     """
 
     parameter_names = ("mean", "sigma")
@@ -181,8 +182,9 @@ class GaussianEstimator(estimator.Estimator):
             if list(mean_bounds).count(None) == 1:
                 raise ValueError("unbounded prior given in mean_bounds")
 
-        self._y, self._yerr = (y, yerr)
+            mean_bounds = tuple(np.sort(mean_bounds))
 
+        self._y, self._yerr = (y, yerr)
         self._bounds = [
             mean_bounds,
             (0, sigma_upper_bound)

### 2017-03-16 snob/estimator.py
Minor

@@ -30,9 +30,7 @@ def log_kappa(D):
         The number of dimensions.
     """
 
-    return  - 0.5 * D * np.log(2 * np.pi) \
-            + 0.5 * np.log(np.pi * D) \
-            - 1
+    return  -0.5*D*np.log(2 * np.pi) + 0.5*np.log(D*np.pi) - 1
 
 
 class Estimator(object):
@@ -94,11 +92,12 @@ class Estimator(object):
         """
         Return the total message length.
         """
-        return (- self.log_prior \
-                + 0.5 * self.log_fisher \
-                - self.log_data \
-                + log_kappa(self.dimensions))
-    
+        return np.sum([
+            -self.log_prior,
+            +0.5*self.log_fisher,
+            -self.log_data,
+            +log_kappa(self.dimensions)
+        ])
 
     @property
     def dimensions(self):

### 2017-03-16 snob/tests/test_gaussian.py

@@ -16,3 +16,36 @@ class TestShortHash(unittest.TestCase):
         estimator = gaussian.GaussianEstimator(data)
 
         
+
+
+class TestGaussianEstimator(unittest.TestCase):
+
+    def test_repr(self):
+
+        N = 5
+        y = np.random.normal(5.2, 0.4, size=N)
+        yerr = np.abs(np.random.normal(0, 0.1, size=N))
+
+        model = gaussian.GaussianEstimator(y=y, yerr=yerr)
+        print(model)
+
+
+    def test_attributes(self):
+
+
+        N = 5
+        y = np.random.normal(5.2, 0.4, size=N)
+        yerr = np.abs(np.random.normal(0, 0.1, size=N))
+
+        model = gaussian.GaussianEstimator(y=y, yerr=yerr)
+
+
+        model.data
+        model.quantum
+        model.message_length
+        model.dimensions
+        model.parameter_names
+
+        model.log_prior
+        model.log_fisher
+        model.log_data

### 2017-03-16 snob/tests/test_estimator.py
base estimator unit tests

@@ -1,0 +1,29 @@
+
+""" Test base estimator. """
+
+import numpy as np
+import unittest
+
+from .. import estimator
+
+
+class TestBaseEstimator(unittest.TestCase):
+
+    def test_attributes(self):
+
+        N = 5
+        data = np.random.uniform(size=N)
+        model = estimator.Estimator(data=y)
+        
+        model.data
+        model.quantum
+        model.message_length
+    
+        with self.assertRaises(NotImplementedError):
+            model.log_prior
+
+        with self.assertRaises(NotImplementedError):
+            model.log_fisher
+
+        with self.assertRaises(NotImplementedError):
+            model.log_data

### 2017-03-16 snob/estimator.py
coveralls is so dum

@@ -30,7 +30,7 @@ def log_kappa(D):
         The number of dimensions.
     """
 
-    return  -0.5*D*np.log(2 * np.pi) + 0.5*np.log(D*np.pi) - 1
+    return  -0.5*D*np.log(2*np.pi) + 0.5*np.log(D*np.pi) - 1
 
 
 class Estimator(object):

### 2017-03-16 snob/gaussian.py

@@ -62,10 +62,12 @@ def _log_fisher(sigma, weights):
     :param weights:
         The array of weights :math:`w` for all data points.
     """
-    return    np.log(2) \
-            - 4 * np.log(sigma) \
-            + np.log(np.sum(weights)) \
-            + np.log(np.sum(weights**2))
+    return np.sum([
+        +np.log(2),
+        -4*np.log(sigma),
+        +np.log(np.sum(weights)),
+        +np.log(np.sum(weights**2))
+    ])
 
 
 def _log_data(mean, sigma, y, quantum):
@@ -85,10 +87,12 @@ def _log_data(mean, sigma, y, quantum):
 
     Where :math:`\epsilon` represents the quantum of the data, :math:`y`.
     """
-    return -(   (y.size / 2.0) * np.log(2 * np.pi) \
-               + y.size * np.log(sigma) \
-               - np.sum(np.log(quantum)) \
-               + np.sum((y - mean)**2)/(2 * sigma**2))
+    return -np.sum([
+        +(y.size/2.0)*np.log(2*np.pi),
+        +y.size*np.log(sigma),
+        -np.sum(np.log(quantum)),
+        +np.sum((y - mean)**2)/(2*sigma**2))
+    ])
 
 
 def _message_length(parameters, y, yerr, bounds, quantum, gradient=False):
@@ -112,10 +116,12 @@ def _message_length(parameters, y, yerr, bounds, quantum, gradient=False):
     s2 = sigma**2
     weights = s2/(s2 + yerr**2)
 
-    I = - _log_prior(sigma, bounds, quantum) \
-        + 0.5 * _log_fisher(sigma, weights) \
-        - _log_data(mean, sigma, y, quantum) \
-        + estimator.log_kappa(2)
+    I = np.sum([
+        -_log_prior(sigma, bounds, quantum),
+        +0.5*_log_fisher(sigma, weights),
+        -_log_data(mean, sigma, y, quantum),
+        +estimator.log_kappa(2)
+    ])
 
     if not gradient:
         return I

### 2017-03-16 snob/tests/test_estimator.py

@@ -11,14 +11,10 @@ class TestBaseEstimator(unittest.TestCase):
 
     def test_attributes(self):
 
-        N = 5
-        data = np.random.uniform(size=N)
-        model = estimator.Estimator(data=y)
-        
-        model.data
-        model.quantum
-        model.message_length
-    
+        model = estimator.Estimator()
+
+        self.assertTrue(model.quantum > 0)
+         
         with self.assertRaises(NotImplementedError):
             model.log_prior
 

### 2017-03-16 snob/gaussian.py
typo

@@ -91,7 +91,7 @@ def _log_data(mean, sigma, y, quantum):
         +(y.size/2.0)*np.log(2*np.pi),
         +y.size*np.log(sigma),
         -np.sum(np.log(quantum)),
-        +np.sum((y - mean)**2)/(2*sigma**2))
+        +np.sum((y - mean)**2)/(2*sigma**2)
     ])
 
 

### 2017-03-16 snob/tests/test_gaussian.py
add unit tests

@@ -49,3 +49,53 @@ class TestGaussianEstimator(unittest.TestCase):
         model.log_prior
         model.log_fisher
         model.log_data
+
+
+    def test_negative_error(self):
+
+        with self.assertRaises(ValueError):
+            gaussian.GaussianEstimator(y=[0], yerr=[-1])
+
+
+    def test_data_shape_mismatch(self):
+
+        with self.assertRaises(ValueError):
+            gaussian.GaussianEstimator(y=[0], yerr=[0.1, 0.1])
+
+        with self.assertRaises(ValueError):
+            gaussian.GaussianEstimator(y=[1, 1], yerr=0.1)
+
+
+    def test_proper_prior_on_mean(self):
+
+        with self.assertRaises(ValueError):
+            gaussian.GaussianEstimator(y=[1], mean_bounds=(None, 3))
+
+        with self.assertRaises(ValueError):
+            gaussian.GaussianEstimator(y=[1], mean_bounds=(3, None))
+
+        self.assertIsNotNone(
+            gaussian.GaussianEstimator(y=[1], mean_bounds=(None, None)))
+
+        self.assertIsNotNone(
+            gaussian.GaussianEstimator(y=[1], mean_bounds=None))
+
+
+
+
+    def test_optimization(self):
+
+        N = 5
+        y = np.random.normal(5.2, 0.4, size=N)
+        yerr = np.abs(np.random.normal(0, 0.1, size=N))
+
+        model = gaussian.GaussianEstimator(y=y, yerr=yerr)
+
+        before = model.message_length
+
+        model.optimize()
+
+        after = model.message_length
+
+        self.assertTrue(after <= before)
+        

### 2017-03-17 snob/gaussian.py
foo

@@ -188,7 +188,11 @@ class GaussianEstimator(estimator.Estimator):
             if list(mean_bounds).count(None) == 1:
                 raise ValueError("unbounded prior given in mean_bounds")
 
-            mean_bounds = tuple(np.sort(mean_bounds))
+            elif None not in mean_bounds:
+                mean_bounds = np.sort(mean_bounds)
+
+            mean_bounds = tuple(mean_bounds)
+
 
         self._y, self._yerr = (y, yerr)
         self._bounds = [

### 2017-03-17 snob/estimator.py
add tests

@@ -78,16 +78,6 @@ class Estimator(object):
 
 
     @property
-    def weights(self):
-        """
-        Return the weights for this estimator.
-        """
-        raise NotImplementedError(
-            "weights are not implemented yet")
-        return self._weights
-
-
-    @property
     def message_length(self):
         """
         Return the total message length.
@@ -105,17 +95,6 @@ class Estimator(object):
         Return the number of dimensions in this estimator.
         """
         return len(self.parameter_names)
-
-
-    def _set_parameter_values(self):
-        """
-        Set the parameter values to the Estimator.
-        """
-
-        values = self.estimate_parameters()
-        for parameter_name, value in zip(self.parameter_names, values):
-            setattr(self, "_{}".format(parameter_name), value)
-        return values
         
 
     @property

### 2017-03-17 snob/tests/test_estimator.py

@@ -23,3 +23,10 @@ class TestBaseEstimator(unittest.TestCase):
 
         with self.assertRaises(NotImplementedError):
             model.log_data
+
+        with self.assertRaises(NotImplementedError):
+            model.parameter_names
+
+
+        model._data = [3]
+        self.assertIsNotNone(model.data)

### 2017-03-17 snob/tests/test_gaussian.py

@@ -80,6 +80,17 @@ class TestGaussianEstimator(unittest.TestCase):
         self.assertIsNotNone(
             gaussian.GaussianEstimator(y=[1], mean_bounds=None))
 
+        self.assertRaises(ValueError):
+            gaussian.GaussianEstimator(y=[1], mean_bounds=[3])
+
+        self.assertRaises(ValueError):
+            gaussian.GaussianEstimator(y=[1], mean_bounds=[1,2,3])
+
+
+        bounds = [5, 1]
+        model = gaussian.GaussianEstimator(y=[2], mean_bounds=bounds)
+        self.assertEqual(model.bounds[0][0] == bounds[1])
+
 
 
 
@@ -98,4 +109,7 @@ class TestGaussianEstimator(unittest.TestCase):
         after = model.message_length
 
         self.assertTrue(after <= before)
-        
+
+        # Ensure we hit the warning flag.
+
+        model.optimize(maxiter=1, maxfun=1)

### 2017-03-17 snob/tests/test_gaussian.py
fix tests

@@ -80,16 +80,16 @@ class TestGaussianEstimator(unittest.TestCase):
         self.assertIsNotNone(
             gaussian.GaussianEstimator(y=[1], mean_bounds=None))
 
-        self.assertRaises(ValueError):
+        with self.assertRaises(ValueError):
             gaussian.GaussianEstimator(y=[1], mean_bounds=[3])
 
-        self.assertRaises(ValueError):
+        with self.assertRaises(ValueError):
             gaussian.GaussianEstimator(y=[1], mean_bounds=[1,2,3])
 
 
         bounds = [5, 1]
         model = gaussian.GaussianEstimator(y=[2], mean_bounds=bounds)
-        self.assertEqual(model.bounds[0][0] == bounds[1])
+        self.assertEqual(model.bounds[0][0], bounds[1])
 
 
 

### 2017-03-17 snob/tests/test_gaussian.py
coverage

@@ -112,4 +112,38 @@ class TestGaussianEstimator(unittest.TestCase):
 
         # Ensure we hit the warning flag.
 
-        model.optimize(maxiter=1, maxfun=1)
+        model.optimize(maxiter=0, maxfun=0)
+
+
+
+    def test_prior_on_mean(self):
+
+        N = 5
+        y = np.random.normal(5.2, 0.4, size=N)
+        yerr = np.abs(np.random.normal(0, 0.1, size=N))
+
+        without_prior = gaussian.GaussianEstimator(y=y, yerr=yerr)
+        without_prior.optimize()
+
+        with_prior = gaussian.GaussianEstimator(
+            y=y, yerr=yerr, mean_bounds=[1, 10])
+        with_prior.optimize()
+
+        print("with", with_prior.message_length)
+        print("without", without_prior.message_length)
+        #self.assertTrue(
+        #    with_prior.message_length > without_prior.message_length)
+
+        print("TODO CHECK PRIOR ON MEAN")
+
+
+    def test_prior_on_sigma(self):
+
+
+
+
+        N = 5
+        y = np.random.normal(5.2, 0.4, size=N)
+        yerr = np.abs(np.random.normal(0, 0.1, size=N))
+
+        gaussian.GaussianEstimator(y=y, yerr=yerr, upper_sigma_bound=10)

### 2017-03-17 snob/gaussian.py
100% coverage?

@@ -95,7 +95,7 @@ def _log_data(mean, sigma, y, quantum):
     ])
 
 
-def _message_length(parameters, y, yerr, bounds, quantum, gradient=False):
+def _message_length(parameters, y, yerr, bounds, quantum):
     """
     Return the approximate message length of the Gaussian estimator, given the
     parameters.
@@ -103,11 +103,20 @@ def _message_length(parameters, y, yerr, bounds, quantum, gradient=False):
     :param parameters:
         A two-length list-like object containing the Gaussian mean and sigma.
 
-    :param estimator:
-        The gaussian estimator.
+    :param y:
+        The data values :math:`y`.
 
-    :param gradient: [optional]
-        Also return the Jacobian of partial derivatives.
+    :param yerr:
+        The normally-distributed sigma uncertainties on :math:`y`.
+
+    :param bounds:
+        A list containing two-length tuples that contain the lower and upper
+        bounds for all parameter names. If no edge bound exists for a parameter
+        then `None` is provided.
+
+    :param quantum:
+        The acceptable rounding-off quantum for the minimum message length
+        in units of nits. Default is 0.1 nit.
     """
 
     mean, sigma = parameters
@@ -116,17 +125,12 @@ def _message_length(parameters, y, yerr, bounds, quantum, gradient=False):
     s2 = sigma**2
     weights = s2/(s2 + yerr**2)
 
-    I = np.sum([
+    return np.sum([
         -_log_prior(sigma, bounds, quantum),
         +0.5*_log_fisher(sigma, weights),
         -_log_data(mean, sigma, y, quantum),
         +estimator.log_kappa(2)
     ])
-
-    if not gradient:
-        return I
-
-    raise NotImplementedError("TODO: put in analytic derivatives")
 
 
 class GaussianEstimator(estimator.Estimator):

### 2017-03-17 snob/tests/test_gaussian.py
foo

@@ -111,8 +111,7 @@ class TestGaussianEstimator(unittest.TestCase):
         self.assertTrue(after <= before)
 
         # Ensure we hit the warning flag.
-
-        model.optimize(maxiter=0, maxfun=0)
+        model.optimize(maxiter=0, maxfun=0, factr=10, pgtol=1e-30)
 
 
 
@@ -146,4 +145,4 @@ class TestGaussianEstimator(unittest.TestCase):
         y = np.random.normal(5.2, 0.4, size=N)
         yerr = np.abs(np.random.normal(0, 0.1, size=N))
 
-        gaussian.GaussianEstimator(y=y, yerr=yerr, upper_sigma_bound=10)
+        gaussian.GaussianEstimator(y=y, yerr=yerr, sigma_upper_bound=10)

### 2017-03-17 snob/tests/test_gaussian.py
hit dat line

@@ -138,11 +138,11 @@ class TestGaussianEstimator(unittest.TestCase):
 
     def test_prior_on_sigma(self):
 
-
-
-
         N = 5
         y = np.random.normal(5.2, 0.4, size=N)
         yerr = np.abs(np.random.normal(0, 0.1, size=N))
 
-        gaussian.GaussianEstimator(y=y, yerr=yerr, sigma_upper_bound=10)
+        model = gaussian.GaussianEstimator(
+            y=y, yerr=yerr, sigma_upper_bound=10)
+
+        model.optimize()

### 2017-03-17 snob/gaussian.py
minor [ci skip]

@@ -33,7 +33,7 @@ def _log_prior(sigma, bounds, quantum):
     return lp
 
 
-def _log_fisher(sigma, weights):
+def _log_fisher(sigma, yerr):
     """
     Return the natural logarithm of the Fisher information.
 
@@ -59,9 +59,11 @@ def _log_fisher(sigma, weights):
     :param sigma:
         The standard deviation :math:`\sigma` of the Gaussian distribution.
 
-    :param weights:
-        The array of weights :math:`w` for all data points.
+    :param yerr:
+        The normally-distributed 1-sigma uncertainties on the data :math:`y`.
     """
+
+    weights = sigma**2/(sigma**2 + yerr**2)
     return np.sum([
         +np.log(2),
         -4*np.log(sigma),
@@ -121,13 +123,9 @@ def _message_length(parameters, y, yerr, bounds, quantum):
 
     mean, sigma = parameters
 
-    # Calculate weights.
-    s2 = sigma**2
-    weights = s2/(s2 + yerr**2)
-
     return np.sum([
         -_log_prior(sigma, bounds, quantum),
-        +0.5*_log_fisher(sigma, weights),
+        +0.5*_log_fisher(sigma, yerr),
         -_log_data(mean, sigma, y, quantum),
         +estimator.log_kappa(2)
     ])

### 2017-03-17 snob/gaussian.py
whoops bug fix [ci skip]

@@ -342,8 +342,7 @@ class GaussianEstimator(estimator.Estimator):
         :math:`y_n`. The unperturbed and perturbed cases are equivalent when 
         :math:`\epsilon_n = 0`.
         """
-        weights = self.sigma**2/(self.sigma**2 + self.yerr**2)
-        return _log_fisher(self.sigma, weights)
+        return _log_fisher(self.sigma, self.yerr)
 
 
     @property

### 2017-04-12 snob/mixture_slf.py
SLF example v0

@@ -1,0 +1,796 @@
+
+"""
+An estimator to model data using a prescribed number of multivariate Gaussian 
+distributions with a single multivariate latent factor.
+"""
+
+__all__ = ["SLFGaussianMixture"]
+
+import logging
+import numpy as np
+import scipy
+from sklearn import cluster
+
+logger = logging.getLogger(__name__)
+
+
+class SLFGaussianMixture(object):
+
+    r"""
+    Model data from (potentially) many multivariate Gaussian distributions
+    with a single (multivariate) latent factor.
+
+    The latent factor and mixture parameters are iteratively updated using
+    expectation-maximization, with minimum message length describing the cost
+    function.
+
+    :param num_components:
+        The number of multivariate Gaussian mixtures to model.
+
+    :param tolerance: [optional]
+        The relative improvement in log probability required before stopping
+        an expectation-maximization step (default: ``1e-5``).
+
+    :param covariance_type: [optional]
+        The structure of the covariance matrix for individual components.
+        The available options are: `full` for a full covariance matrix,
+        `diag` for a diagonal covariance matrix, `tied` for a common covariance
+        matrix for all components, `tied_diag` for a common diagonal
+        covariance matrix for all components (default: ``full``).
+
+    :param covariance_regularization: [optional]
+        Regularization strength to apply to the diagonal of covariance
+        matrices (default: ``0``)
+
+    :param max_iter: [optional]
+        The maximum number of iterations to run per expectation-maximization
+        loop (default: ``10000``).
+
+    :param initialization_method: [optional]
+        The method to use to initialize the mixture parameters. Available
+        options are: ``kmeans``, and ``random`` (default: ``kmeans``).
+    """
+
+    parameter_names \
+        = ("means", "factor_scores", "factor_loads", "specific_variances")
+
+    def __init__(self, num_components, tolerance=1e-5, covariance_type="full", 
+        covariance_regularization=0, max_iter=10000, 
+        initialization_method="kmeans", **kwargs):
+
+        assert num_components == 1
+
+        num_components = int(num_components)
+        if 1 > num_components:
+            raise ValueError("number of components must be a positive integer")
+
+        if 0 >= tolerance:
[…]

### 2017-04-12 snob/mixture.py
bug fixes

@@ -219,6 +219,11 @@ class GaussianMixture(object):
             If there is a shape mis-match between the input arrays.
         """
 
+        if None in (means, covariances, weights):
+            for parameter_name in self.parameter_names:
+                setattr(self, "_{}".format(parameter_name), None)
+                return False
+
         means = np.atleast_2d(means)
         covariances = np.atleast_3d(covariances)
         weights = np.array(weights)
@@ -238,12 +243,10 @@ class GaussianMixture(object):
         self._covariances = covariances
         self._weights = weights
 
-        if None not in (means, covariances, weights):         
-            self._precision_cholesky = _compute_cholesky_decomposition(
-                self.covariances, self.covariance_type)
-            return True
+        self._precision_cholesky = _compute_cholesky_decomposition(
+            self.covariances, self.covariance_type)
+        return True
 
-        return False
 
 
     def fit(self, y, **kwargs):
@@ -315,6 +318,7 @@ class GaussianMixture(object):
             # Ignore underflow errors.
             log_resp = weighted_log_prob - log_prob_norm[:, np.newaxis]
 
+        raise a
         return (log_prob_norm, log_resp)
 
 
@@ -519,7 +523,7 @@ def _compute_log_det_cholesky(matrix_chol, covariance_type, D):
         n_components, _, _ = matrix_chol.shape
         log_det_chol = (np.sum(np.log(
             matrix_chol.reshape(
-                n_components, -1)[:, ::n_features + 1]), 1))
+                n_components, -1)[:, ::D + 1]), 1))
 
     elif covariance_type == 'tied':
         log_det_chol = (np.sum(np.log(np.diag(matrix_chol))))
@@ -574,8 +578,8 @@ def _estimate_log_gaussian_prob(y, means, precision_cholesky, covariance_type):
             diff = np.dot(y, prec_chol) - np.dot(mu, prec_chol)
             log_prob[:, k] = np.sum(np.square(diff), axis=1)
 
-    return -0.5 * (D * np.log(2 * np.pi) + log_prob) + log_det
-
+    return  -0.5 * (D * np.log(2 * np.pi) + log_prob) + log_det
+    
 
 def _compute_cholesky_decomposition(covariances, covariance_type):
     r"""

### 2017-04-12 sandbox_mixture_slf.py
clean up SLF code

@@ -1,0 +1,47 @@
+import numpy as np
+from snob import mixture_slf as slf
+
+
+#y = np.loadtxt("cluster_abundances.txt")
+
+n_samples, n_features, rank = 1000, 50, 1
+sigma = 0.5
+true_specific_variances = sigma**2 * np.ones((1, n_features))
+
+
+rng = np.random.RandomState(100)
+
+U, _, _ = np.linalg.svd(rng.randn(n_features, n_features))
+true_factor_loads = U[:, :rank].T
+
+
+true_factor_scores = rng.randn(n_samples, rank)
+X = np.dot(true_factor_scores, true_factor_loads)
+
+# Adding homoscedastic noise
+bar = rng.randn(n_samples, n_features)
+X_homo = X + sigma * bar
+
+# Adding heteroscedastic noise
+sigmas = sigma * rng.rand(n_features) + sigma / 2.
+X_hetero = X + rng.randn(n_samples, n_features) * sigmas
+
+
+#y = np.atleast_2d(np.loadtxt("coffee_example.txt")).T
+
+true_means = np.zeros((1, n_features))
+
+model = slf.SLFGaussianMixture(1)
+model.fit(X_homo)
+
+
+raise a
+
+true = np.hstack([each.flatten() for each in (true_means, true_factor_scores, true_factor_loads, true_specific_variances)])
+
+inferred = np.hstack([each.flatten() for each in (model.means, model.factor_scores, model.factor_loads, model.specific_variances)])
+
+fig, ax = plt.subplots()
+ax.scatter(true, inferred, alpha=0.5)
+
+raise a

### 2017-04-12 snob/mixture_slf.py

@@ -51,8 +51,8 @@ class SLFGaussianMixture(object):
         options are: ``kmeans``, and ``random`` (default: ``kmeans``).
     """
 
-    parameter_names \
-        = ("means", "factor_scores", "factor_loads", "specific_variances")
+    parameter_names = (
+        "means", "factor_loads", "factor_scores", "specific_variances")
 
     def __init__(self, num_components, tolerance=1e-5, covariance_type="full", 
         covariance_regularization=0, max_iter=10000, 
@@ -215,7 +215,7 @@ class SLFGaussianMixture(object):
         return False
 
 
-    def fit(self, x, foo1, foo2, foo3, **kwargs):
+    def fit(self, x, **kwargs):
         r"""
         Fit the mixture model to the data using the expectation-maximization
         algorithm, and minimum message length to update the parameter
@@ -227,11 +227,9 @@ class SLFGaussianMixture(object):
             the number of dimensions per observation.
         """
 
+        # Solve for latent factors first.
         N, K = x.shape
 
-        # Solve for latent factors first.
-
-        # Calculate means.
         means = np.mean(x, axis=0)
 
         w = x - means
@@ -239,198 +237,64 @@ class SLFGaussianMixture(object):
 
         U, S, V = np.linalg.svd(correlation_matrix)
 
-        """
-        # Take b as the dominant eigenvector of the correlation matrix,
-        # and set b^2 as the dominant eigenvalue.
-        b = V[0] * np.sqrt(S[0])
-
-        def message_length(b):
-
-            b_squared = np.sum(b**2)
-
-            specific_variances = np.sum(w**2, axis=0) \
-                               / ((N - 1) * (1 + b))
-
-            v_scale = (1 - K/((N - 1) * b_squared)) / (1 + b_squared)
-
-            
-
-            specific_sigma = np.sqrt(specific_variances)
-            Y = np.dot(w.T, w)\
-              / np.dot(specific_sigma.T, specific_sigma)
-
-
-            Ybb = np.sum(np.dot(Y, np.dot(b.T, b)))
-            wbv = Ybb * v_scale
-            v2 = wbv * v_scale
-            residual2 = (b_squared + 1.0) * (N - 1) - 2 * wbv + b_squared * v2
-
-            foo= (N - 1) * np.sum(np.log(specific_sigma)) \
-                +  0.5 * K * np.log(N * v2) \
[…]

### 2017-04-13 articles/chemical-tagging/ms.tex
bump travis
Fix Travis
start ranting about shit

@@
 % Define commands
+\newcommand{\article}{\emph{Article}}
 \newcommand{\acronym}[1]{{\small{#1}}}
 \newcommand{\project}[1]{\textsl{#1}}
+
+% Affiliation(s)
+\newcommand{\moca}{
+	\affil{School of Physics and Astronomy, Monash University, 
+		   Melbourne, Clayton VIC 3800, Australia}}
+\newcommand{\claytonfit}{
+	\affil{Faculty of Information Technology, Monash University,
+	       Melbourne, Clayton VIC 3800, Australia}}
+\newcommand{\caulfieldfit}{
+	\affil{Faculty of Information Technology, Monash University,
+		   Melbourne, Caulfield East VIC 3145, Australia}}
 
@@
 
-\submitjournal{AAS Journals}
+\submitjournal{TBD} % ApJ, MNRAS, PASA?
 
-\shorttitle{Chemical tagging can work, but will it?}
+\shorttitle{Chemical Tagging Using Minimum Message Length}
 \shortauthors{Casey et al.}
@@
 
-\title{Chemical tagging can work, but will it?}
+\title{Chemical Tagging Using Minimum Message Length}
 
@@
 \author[0000-0003-0174-0564]{Andrew R. Casey}
-\affil{Monash Centre for Astrophysics, Monash University, Melbourne, 
-	   VIC 3800, Australia}
+\moca
+\claytonfit
 
-\author{(some order of: John Lattanzio, David Dowe, Aldeida Aleti)}
+\author{(some order of:}
+
+\author[0000-0003-2952-859X]{John Lattanzio}
+\moca
+
+\author[0000-0002-0583-5918]{David Dowe} 
+\claytonfit
+
+\author[0000-0002-1716-690X]{Aldeida Aleti}
+\caulfieldfit
+
+\author{)}
 
@@
 \begin{abstract}
-Snobbery.
+Chemical tagging seeks to identify co-natal stars by their present-day 
+photospheric abundances, long after their phase space similarity is lost.
+In principle the detailed chemical abundances of a transformative number of
+stars could be used to reconstruct the evolution of the Milky Way, providing
+insight on supernovae yields, galactic dynamics (e.g., constraints on radial 
+mixing), the initial mass function, as well as the formation and destruction
+of star clusters. Some progress has been made towards chemical tagging, but 
+here we argue that even the \emph{extent} of the problem has not been realised.
+We introduce the Real World problems associated with chemical tagging at scale,
+and describe how currently considered methods will inevitably fail even under
+the most unrealistically optimistic conditions. We introduce the minimum message
+length to the astronomical community as a promising alternative to chemical
+tagging, one which has the capability to address all ensivaged problems with
+chemical tagging simultaneously. We perform chemical tagging experiments with
+infinte Gaussian mixture models, and demonstrate using real and generated data 
+that MML outperforms all other considered penalty techniques. 
 \end{abstract}
@@
 
+\citet{Freeman_2002} introduced chemical tagging as an idea to identify groups of stars that formed together in the same gas cloud, which have now become physically separated in phase space.
+This idea is attractive because the observable chemical abundances of stars remains largely unchanged throughout a star's lifetime, whereas phase space information is quickly damped through dynamical interactions.
+Some progress has been made towards chemical tagging in simplistic experiments, but the full extent of chemical tagging has yet to be realised.
+Indeed, in this \article\ we argue that the full extent \emph{of the problem} of chemical tagging has not even been realised yet.
+
 \acknowledgments
+Who?
+Grants.
 

### 2017-04-13 articles/chemical-tagging/ms.tex
rant about some crap [[ci skip] because fail anyways]

@@
 \documentclass{aastex61}
+
+\usepackage{bm}
 
@@
 
-\citet{Freeman_2002} introduced chemical tagging as an idea to identify groups of stars that formed together in the same gas cloud, which have now become physically separated in phase space.
-This idea is attractive because the observable chemical abundances of stars remains largely unchanged throughout a star's lifetime, whereas phase space information is quickly damped through dynamical interactions.
-Some progress has been made towards chemical tagging in simplistic experiments, but the full extent of chemical tagging has yet to be realised.
-Indeed, in this \article\ we argue that the full extent \emph{of the problem} of chemical tagging has not even been realised yet.
+\citet{Freeman_2002} introduced chemical tagging as an idea to identify groups 
+of stars that formed together in the same gas cloud, which are no longer
+identifiable in phase space, either by their physical location or orbital
+dynamics.  This idea is attractive because the observable chemical abundances of 
+stars remains largely unchanged throughout a star's lifetime, whereas phase 
+space information is quickly damped through dynamical interactions. Moreover, 
+while most stars are thought to form in star clusters \citep{someone}, only the
+most massive or isolated star clusters are able to survive the star formation
+process \citep{someone2}. For this reason we rely on `field' stars -- that is,
+stars that are not clearly associated with a known star cluster -- in order to
+infer the formation of the Milky Way.
+
+Some progress has been made towards chemical tagging in simplistic experiments, 
+but the full extent of chemical tagging has yet to be realised. Indeed, in this 
+\article\ we assert that even \emph{the full extent of the problem} of chemical 
+tagging has yet to be realised. 
+% Some other comments about this?
+
+\subsection*{A Probabilistic Approach to Chemical Tagging}
+\ref{section:the-problem}
+
+% First formulate the problem: what is needed to make things work?
+
+% What work has been done on chemical tagging?
+
+% Unfortunately these approaches have only sought to solve individual problems
+% within chemical tagging. Doing so -- and then later using those results to
+% infer something else about the galaxy -- makes both results inconsistent,
+% both statistically, and otherwise.
+
+In Section \ref{sec:model} we introduce a simplistic approach to chemical
+tagging that relies on probabilistically modelling stellar chemical abundances
+as a mixture of (potentially infinite) multivariate Gaussian distributions.
+
+There we introduce the concept of MML,
+
+\section{The Model}
+\ref{sec:the-model}
+
+Here we introduce a simple probablistic approach to chemical tagging, and in
+later sections we will formulate the objective function for MML. Our approach is
+the simplistic in the sense that we will ignore many of the aforementioned
+problems that will plague chemical tagging at scale. Specifically we make the
+following assumptions:
+
+\begin{itemize}
+\item We assume that the data (e.g., the detailed chemical abundances of stars)
+      can be represented by a mixture of multivariate Gaussian distributions.
+\item We will assume that the number of \emph{true} multivariate Gaussian
+      distributions is not known. This assumption applies to our experiements
+      involving generated data, and those using real data.
+\item In addition to not knowing the \emph{number} of multivariate normal
+      distributions, we further assume that the means and covariance matrices of
+      each mixture is unknown.
+\item Here we assume that the covariance matrix of individual components can be
+      described as `free` or `full', in the sense that the data within a cluster
+      can have off-diagonal correlation terms, which are unknown.
+\item In all experiments we will assume that the data have homoskedastic noise
+      properties.
+\item The relative weights (or mixing proportions) of different distributions
+      is unknown. We further will assume a relative (or conditional)
+      probability distribution for each star belonging to a given cluster. That
+      is to say, we do not adopt a `\emph{hard selection}' approach to cluster
+      modelling, where stars would be assigned as definitively belonging to one
+      cluster or another.
+      % TODO: should we actually introduce the partial assignment later when we 
+      % talk about the Gaussian distributions?
+\end{itemize}
+
+We have made relatively few explicit assumptions about the model for the data,
+but even the model described constitutes a very formidable model selection 
+problem. The reason is because we do not know \emph{how many} multivariate
+Gaussian distributions we ought to chose to fit the data. Including more
+mixtures will consistently provide a better fit to the data (in a likelihood
+sense), but we must make some decision or heuristic as to whether the additional
+mixture is justified.
+
+The probability density function $f$ for a multivariate normal distribution with
+$d$ dimensions is given by
+
+\begin{equation}
+    f(y_n|\bm{\mu},\bm{\Sigma}) 
+        = \frac{1}{\sqrt{(2\pi)^d|\bm{\Sigma}|}}
+          \exp{\left[-\frac{1}{2}(\bm{y} - \bm{\mu})^\intercal\bm{\Sigma}^{-1}(\bm{y} - \bm{\mu})\right]}
+\end{equation}
+
+\noindent{}where $\bm{\mu}$ and $\bm{\Sigma}$ is the multivariate mean and 
+covariance matrix, and $\bm{y}$ are the data.  When modelling a mixture of
+Gaussians, there are two approaches in determining assignments (or
+responsibilities) of each star to a given cluster. One approach is to include
+a (discrete) model parameter for every data point, which represents the index
+of the mixture that the observation is assigned to. The second approach is to 
+use a weighted mixture. These two are mathematically identical.
+% TODO
+
+\begin{equation}
+% TODO
+\end{equation}
+
+\section{Introducing MML}
+% MML formalism
+
+\subsection{Prior probabilities on the parameters}
+
+We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
+their summation equals one. We adopt a wide uniform prior of $\mathcal{U}(\bm{\mu}) = [-12, 12]$ on all
+multivariate mean abundances $\bm{\mu}$ (in all $d$ dimensions). This prior
+is sufficiently wide that it has no impact on our inferences (24 orders of
+magnitude!), but remains proper. We adopt a congujate inverted Wishart prior 
+for the covariance matrices in the individual mixtures. Ignoring the effect of 
+our prior on $\bm{\mu}$ gives us a proportional joint prior,
+
+\begin{equation}
+    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} \right.
+\end{equation}
+% TODO: DERIVE THIS.
 

### 2017-04-13 articles/chemical-tagging/ms.tex
add the 12 disciples :praying:

@@
 \newcommand{\project}[1]{\textsl{#1}}
+
+% Have some maaaath.
+\def\infinity{\rotatebox{90}{8}}
 
@@
 
-% First formulate the problem: what is needed to make things work?
+What is needed to make things work? We outline the 12 disciples:
+
+\begin{enumerate}
+\item The number of clusters.
+\item The number of latent factors (e.g., supernovae yields, AGB yields).
+\item The multivariate factor loads of individual latent factors.
+\item The relative weighting (yet another latent factor, which is however not
+      common between all mixtures) of individual mixtures.
+\item The relative \emph{scoring} of different latent factors on each star.
+\item The intrinsic \emph{specific variances} of the individual abundances.
+\item The multivariate mean abundances and covariance matrices of each star
+      cluster, where the covariance matrix includes off-diagonal terms that
+      describe the \emph{data} correlations (e.g., the correlations that result
+      from trying to measure multiple abundances from the same data).
+\item Atomic diffusion, veiling, and thermohaline mixing: The surface chemical
+      abundances of stars \emph{will change} over its lifetime.
+\item Probabilistic priors (perhaps very hard priors) on joint responsibilities
+      of stars that are specified from astrometric constraints.
+\item There are missing data. Accept it.
+\item There are \emph{upper limits} on some data.\footnote{
+        Spectroscopists are probably the worst in the Universe when it comes to
+        reporting upper limits. They will report an `\emph{upper limit}', when
+        what might mean is the \emph{absolute limit} where there is zero
+        likelihood that the true value is above the given point estimate (which might as well be $+\infinity$, or
+        they may describe a \emph{pseudo-1$\sigma$} limit which is estimated
+        from the locally (yet covariant!) heteroskedastic noise in the data (which often might as well be $\approx$0).
+        In reality, an upper limit ought to be given as a probability density
+        function that is skewed in one direction, perhaps similar to the shape
+        of a log-normal distribution. In any case, the confidence to which they
+        describe their `\emph{upper limit}' ought to be given.}
+\item There are correlations in the measurements that result from trying to
+      measure two (or more) abundances from the same data. As an example,
+      consider a rare -- yet important -- chemical abundance that has a single
+      atomic absorption line which is blended with a more common elemental
+      abundance. In all likelihood, there will be a strong positive correlation
+      between the abundance derived between the rare and more common elemental
+      abundance.
+\end{enumerate}
 
@@
 \begin{equation}
-    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} \right.
+    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} 
 \end{equation}

### 2017-04-19 sandbox_mixture_slf.py
first steps at optimizing SLF + GMM using EM

@@ -4,7 +4,7 @@ from snob import mixture_slf as slf
 
 #y = np.loadtxt("cluster_abundances.txt")
 
-n_samples, n_features, rank = 1000, 50, 1
+n_samples, n_features, n_clusters, rank = 1000, 50, 2, 1
 sigma = 0.5
 true_specific_variances = sigma**2 * np.ones((1, n_features))
 
@@ -18,6 +18,13 @@ true_factor_loads = U[:, :rank].T
 true_factor_scores = rng.randn(n_samples, rank)
 X = np.dot(true_factor_scores, true_factor_loads)
 
+# Assign objects to different clusters.
+indices = rng.randint(0, n_clusters, size=n_samples)
+#mu = 10 * np.arange(n_clusters)#rng.randn(n_clusters, n_features)
+true_means = rng.randn(n_clusters, n_features)
+for index in range(n_clusters):
+    X[indices==index] += true_means[index]
+
 # Adding homoscedastic noise
 bar = rng.randn(n_samples, n_features)
 X_homo = X + sigma * bar
@@ -29,10 +36,43 @@ X_hetero = X + rng.randn(n_samples, n_features) * sigmas
 
 #y = np.atleast_2d(np.loadtxt("coffee_example.txt")).T
 
-true_means = np.zeros((1, n_features))
+#true_means = np.zeros((1, n_features))
 
-model = slf.SLFGaussianMixture(1)
-model.fit(X_homo)
+model = slf.SLFGaussianMixture(2)
+model.initialize_parameters(X_homo)
+
+#model.fit(X_homo)
+fig, axes = plt.subplots(3)
+
+# means
+# This one is tricky because the indices are not necessarily the same.
+# So just take whichever is closest.
+indices = np.zeros(n_clusters, dtype=int)
+for index, true_mean in enumerate(true_means):
+    distance = np.sum(np.abs(model._means - true_mean), axis=1)
+    indices[index] = np.argmin(distance)
+
+assert len(indices) == len(set(indices))
+
+ax = axes[0]
+true = true_means.flatten()
+inferred = model._means[indices].flatten()
+ax.scatter(true, inferred)
+
+
+# factor scores
+ax = axes[1]
+true = true_factor_scores.flatten()
+inferred = model._factor_scores.flatten()
+ax.scatter(true, inferred)
+
+# factor loads
+ax = axes[2]
+true = true_factor_loads.flatten()
+inferred = model._factor_loads.flatten()
+ax.scatter(true, inferred)
+
+
 
 
 raise a

### 2017-04-19 snob/mixture_slf.py

@@ -52,13 +52,12 @@ class SLFGaussianMixture(object):
     """
 
     parameter_names = (
-        "means", "factor_loads", "factor_scores", "specific_variances")
+        "factor_loads", "factor_scores", "specific_variances",
+        "means", "covariances", "weights")
 
     def __init__(self, num_components, tolerance=1e-5, covariance_type="full", 
         covariance_regularization=0, max_iter=10000, 
         initialization_method="kmeans", **kwargs):
-
-        assert num_components == 1
 
         num_components = int(num_components)
         if 1 > num_components:
@@ -144,7 +143,63 @@ class SLFGaussianMixture(object):
             the number of dimensions per observation.
         """
 
+        # Estimate the latent factor first.
         N, K = y.shape
+
+        means = np.mean(y, axis=0)
+        w = y - means
+        correlation_matrix = np.corrcoef(w.T)
+
+        U, S, V = np.linalg.svd(correlation_matrix)
+
+        # Set the initial values for b as the principal eigenvector, scaled by
+        # the square-root of the principal eigenvalue
+        b = V[0] * S[0]**0.5
+
+        specific_sigmas = np.sqrt(np.diag(np.dot(w.T, w)) \
+                    / ((b*b + 1.0) * (N - 1)))
+        factor_loads = specific_sigmas * b
+        _y = (y - means)/specific_sigmas
+
+        b_sq = np.sum(b**2)
+        factor_scores = np.dot(_y, b) * (1 - K/(N - 1) * b_sq)/(1 + b_sq)
+        specific_variances = specific_sigmas**2
+
+        # Subtract off the factor loads and factor scores, then estimate the
+        # cluster properties.
+        factor_scores = np.atleast_2d(factor_scores)
+        factor_loads = np.atleast_2d(factor_loads)
+        y_nlf = y - np.dot(factor_scores.T, factor_loads)
+
+        if self.initialization_method == "kmeans":
+
+            # Do k-means on the data without the latent factor.
+            responsibility = np.zeros((N, self.num_components))
+            kmeans = cluster.KMeans(
+                n_clusters=self.num_components, n_init=1, **kwargs)
+            labels = kmeans.fit(y_nlf).labels_
+            responsibility[np.arange(N), labels] = 1
+
+        elif self.initialization_method == "random":
+            responsibility = np.random.randint(0, self.num_components, size=N)
+            responsibility /= responsibility.sum(axis=1)[:, np.newaxis]
+
+        else:
[…]

### 2017-04-19 articles/chemical-tagging/ms.tex
some words

@@
 \article\ we assert that even \emph{the full extent of the problem} of chemical 
-tagging has yet to be realised. 
-% Some other comments about this?
+tagging has yet to be realised. Before reviewing the existing literature on
+what has been done with chemical tagging, we will outline the necessary
+components to perform chemical tagging.
 
-\subsection*{A Probabilistic Approach to Chemical Tagging}
-\ref{section:the-problem}
+\subsection{The Problem: What's needed for chemical tagging?}
+\label{section:the-problem}
 
-What is needed to make things work? We outline the 12 disciples:
+Given a set of stars that have been observed spectroscopically, and detailed
+chemical abundances have been derived, the following issues will arise when
+trying to perform chemical tagging:
 
@@
 
+Previous approaches have tried to solve separate components of chemical tagging (e.g. clustering). Some notes on this.
+
+These approaches are not consistent, because the problems are intertwined (e.g., clustering, latent factors).
+
 % What work has been done on chemical tagging?
@@
 \section{The Model}
-\ref{sec:the-model}
+\label{sec:the-model}
 
@@
 \begin{equation}
-    f(y_n|\bm{\mu},\bm{\Sigma}) 
+    f(\bm{y}|\bm{\mu},\bm{\Sigma}) 
         = \frac{1}{\sqrt{(2\pi)^d|\bm{\Sigma}|}}
@@
 
-\section{Introducing MML}
-% MML formalism
+\subsection{Minimum Message Length}
+\label{sec:mml}
+
+Background to MML, etc.
+
+We want to minimize the message length.
+
+Link to Bayesian evidence, other things, etc.
+
+In order to minimize the full message length, we must specify the priors on
+our parameters.
 
@@
 
+\subsection{Expected Fisher information}
+
+Computing the expected Fisher information requires the second order partial
+derivatives of the negative log-likelihood $-\mathcal{L}(\bm{y}|\bm{\mu},\bm{\Sigma})$.
+This can be tricky for mixtures -- explain why.
+We approximate the determinant of of the Fisher information matrix,
+$|\mathcal{F}(\bm{\mu},\bm{\Sigma})|$ as the product of $|\mathcal{F}\left(\bm{\mu}\right)|$ and $|\mathcal{F}\left(\bm{\Sigma}\right)|$ \citep{Oliver_1996,Roberts_1998}.
+
+% Equation for log-likelihood of the data:
+% \begin{equation}
+% \mathcal{L}\left(\bm{y}|\bm{\mu},\bm{\Sigma}) = -\frac{Nd}{2}\log{\left(2\pi\right)} - \frac{N}{2}\log{|\bm{\Sigma}|} - \frac{1}{2}\Sigma_{i=1}^{N}(\bm{y}_i - \bm{mu})^\intercal{}\Sigma^{-1}(\bm{y}_i - \bm{\mu})
+%\end{equation}
+
+Taking the second derivative of the log-likelihood of the data with respect to $\bm{\mu}$:
+
+\begin{equation}
+\frac{\delta\mathcal{L}}{\delta\bm{\mu}} = \Sigma_{i=1}^{N}\bm{C}^{-1}(\bm{y}_i - \bm{\mu})
+\end{equation}
+
+\begin{equation}
+\frac{\delta^2\mathcal{L}}{\delta\bm{\mu}^2} = N\bm{C}^{-1}
+\end{equation}
+
+Computing the Fisher information of the covariance matrix $|\mathcal{F}\left(\bm{C}\right)|$ is harder. For a full covariance matrix with off-diagonal terms, the total
+number of free parameters required is $d(d+1)/2$.
+\citep{Magnus_1988} derived an analytical expression:
+% Dwyer_1967, Bozdogan_1990? Drton_2009?
+% TODO: This derivation is a nightmare to find
+
+\begin{equation}
+|\mathcal{F}\left(\bm{C}\right) = N^\frac{d(d+1)}{2}2^{-d}|\bm{C}|^{-(d+1)}
+\end{equation}
+
+Giving the approximate determinant of the Fisher information $|\mathcal{F}\left(\bm{\mu},\bm{C}\right)|$ as:
+
+\begin{eqnarray}
+  |\mathcal{F}\left(\bm{\mu},\bm{C}\right)| & \approx & |\mathcal{F}\left(\bm{\mu}\right)|\cdot|\mathcal{F}\left(\bm{C}\right)| \\
+  & \approx & N^{d}|\bm{C}|^{-1}N^\frac{d(d+1)}{2}2^{-d}|\bm{C}|^{-(d+1)} \\
+  |\mathcal{F}\left(\bm{\mu},\bm{C}\right)| & \approx & N^\frac{d(d+3)}{2}2^{-d}|\bm{C}|^{-(d+2)}
+\end{eqnarray}
+
+A common method to estimate the parameters of a Gaussian distribution is by
+maximum likelihood, where the log-likelihood $\mathcal{L}$ is maximized. In
+order to do this, one computes the gradient of the log-likelihood function
+with respect to the parameters and solves the resulting equations.
+
+- Encoding the message.
+
+- Process of expectation and maximization for each step.
+
 \acknowledgments

### 2017-04-19 articles/chemical-tagging/ms.tex
some words (probably bad)

@@
 
-\shorttitle{Chemical Tagging Using Minimum Message Length}
+\shorttitle{Chemical Tagging Using Minimum Message Length. I.}
 \shortauthors{Casey et al.}
@@
 
-\title{Chemical Tagging Using Minimum Message Length}
+\title{Chemical Tagging Using Minimum Message Length. I.\\
+       Gaussian Mixture Modelling}
 
@@
 \begin{enumerate}
-\item The number of clusters.
+\item The number of star clusters (or star-forming events) where at least one
+      star was observed and detailed chemical abundances were derived.
 \item The number of latent factors (e.g., supernovae yields, AGB yields).
-\item The multivariate factor loads of individual latent factors.
+      Within the chemical tagging literature, this is commonly referred to as
+      the number of `effective chemical dimensions'.
+\item The multivariate factor loads of individual latent factors. For example,
+      if core-collapse supernovae and AGB stars had both contributed to the
+      chemical abundances of all stars in the sample, then the
+      \emph{factor loads} would be the \emph{yield} of the supernovae, or the
+      yield of relative abundances that were produced by the AGB star. Each
+      star will be affected differently by these yields (see next item).
+\item The relative \emph{scoring} of different latent factors on each star.
+      This is a scaling that is applied to the latent yields. Continuing our
+      analogy on yields from core-collapse supernovae and AGB stars, the
+      \emph{factor scores} can be thought of as the relative contributions of
+      the different factor loads, for a single star.
 \item The relative weighting (yet another latent factor, which is however not
       common between all mixtures) of individual mixtures.
-\item The relative \emph{scoring} of different latent factors on each star.
-\item The intrinsic \emph{specific variances} of the individual abundances.
 \item The multivariate mean abundances and covariance matrices of each star
@@
       from trying to measure multiple abundances from the same data).
+\item The intrinsic \emph{specific variances} of the individual abundances.
 \item Atomic diffusion, veiling, and thermohaline mixing: The surface chemical
@@
       of stars that are specified from astrometric constraints.
-\item There are missing data. Accept it.
+\item There are missing data. 
 \item There are \emph{upper limits} on some data.\footnote{
@@
 
-There we introduce the concept of MML,
+In this \article\ we will introduce a simple probabilistic approach to
+chemical tagging, where the data are modelled as a mixture of multivariate
+Gaussian distributions (Section \ref{sec:the-model}). In Section \ref{sec:mml}
+we introduce the principle of minimum message length (MML), and formulate our
+objective function by minimising the total message length. Our model is
+simplistic in the sense that we will ignore many of the self-consistency
+issues already introduced. However, in Section \ref{sec:experiments} we show
+that our MML-based approach is superior to all other methods considered for
+chemical tagging, all of which cannot alleviate the self-consistency issues.
+In Section \ref{sec:discussion} we will describe how MML can overcome these
+aforementioned issues, in order to construct a fully-consistent probabilistic
+approach to chemical tagging.
 
@@
 
-Here we introduce a simple probablistic approach to chemical tagging, and in
-later sections we will formulate the objective function for MML. Our approach is
-the simplistic in the sense that we will ignore many of the aforementioned
-problems that will plague chemical tagging at scale. Specifically we make the
-following assumptions:
+We make the following explicit assumptions:
 
 \begin{itemize}
-\item We assume that the data (e.g., the detailed chemical abundances of stars)
-      can be represented by a mixture of multivariate Gaussian distributions.
+\item We assume that the data (e.g., the $D$ detailed chemical abundances of 
+      $N$ stars) can be represented by a mixture of $K$ multivariate Gaussian
+      distributions.
+\item We assume that there are no missing data values. For example, if a star
+      has been observed, then there is a complete set of $D$ chemical 
+      abundances available.
 \item We will assume that the number of \emph{true} multivariate Gaussian
-      distributions is not known. This assumption applies to our experiements
-      involving generated data, and those using real data.
+      distributions $K_{true}$ is not known. This assumption applies to our 
+      experiements involving generated data, and those using real data.
 \item In addition to not knowing the \emph{number} of multivariate normal
-      distributions, we further assume that the means and covariance matrices of
-      each mixture is unknown.
-\item Here we assume that the covariance matrix of individual components can be
-      described as `free` or `full', in the sense that the data within a cluster
-      can have off-diagonal correlation terms, which are unknown.
+      distributions, we further assume that the means $\bm{\mu}$ and 
+      covariance matrices $\bm{C}$ of each $K$th mixture is unknown.
+\item We assume that the covariance matrix of individual components can be
+      described as `free' or `full', insomuch that the data within a cluster
+      can have off-diagonal correlation terms, and those terms are unknown.
 \item In all experiments we will assume that the data have homoskedastic noise
       properties.
-\item The relative weights (or mixing proportions) of different distributions
-      is unknown. We further will assume a relative (or conditional)
-      probability distribution for each star belonging to a given cluster. That
-      is to say, we do not adopt a `\emph{hard selection}' approach to cluster
-      modelling, where stars would be assigned as definitively belonging to one
-      cluster or another.
-      % TODO: should we actually introduce the partial assignment later when we 
-      % talk about the Gaussian distributions?
+\item The relative weights $\bm{w}$ (or mixing proportions) of $K$ 
+      distributions is unknown. 
+\item We assume a relative (or conditional) probability distribution for each 
+      star belonging to a given cluster. That is to say that we do not adopt a 
+      `\emph{hard selection}' approach to cluster modelling, where stars would
+      be assigned as definitively belonging to one cluster or another. Our
+      approach could be described as `\emph{soft} clustering'. This assumption
+      has practical numerical outcomes and implications for the message length.
 \end{itemize}
@@
 but even the model described constitutes a very formidable model selection 
-problem. The reason is because we do not know \emph{how many} multivariate
-Gaussian distributions we ought to chose to fit the data. Including more
-mixtures will consistently provide a better fit to the data (in a likelihood
-sense), but we must make some decision or heuristic as to whether the additional
-mixture is justified.
+problem. This is because we do not know \emph{how many} multivariate Gaussian
+distributions we ought to chose to fit the data. Including more mixtures will
+consistently provide a better fit to the data (in a likelihood sense), but we 
+must make some decision or heuristic as to whether the additional mixture is 
+justified.
 
-The probability density function $f$ for a multivariate normal distribution with
-$d$ dimensions is given by
+The probability density function $f$ for a multivariate normal distribution 
+with $d$ dimensions is given by
 
@@
 \noindent{}where $\bm{\mu}$ and $\bm{\Sigma}$ is the multivariate mean and 
-covariance matrix, and $\bm{y}$ are the data.  When modelling a mixture of
+covariance matrix, and $\bm{y}$ are the data.
+
+% Full probability for the data.
+
+When modelling a mixture of
 Gaussians, there are two approaches in determining assignments (or
@@
 We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
-their summation equals one. We adopt a wide uniform prior of $\mathcal{U}(\bm{\mu}) = [-12, 12]$ on all
-multivariate mean abundances $\bm{\mu}$ (in all $d$ dimensions). This prior
-is sufficiently wide that it has no impact on our inferences (24 orders of
-magnitude!), but remains proper. We adopt a congujate inverted Wishart prior 
-for the covariance matrices in the individual mixtures. Ignoring the effect of 
-our prior on $\bm{\mu}$ gives us a proportional joint prior,
+$\sum_{k=1}^{K}w_k = 1$. A useful extension of this work might be to impose
+a Dirchlet prior on the mixing weights $\bm{w}$, or a distribution inspired by
+the stellar mass function and observed selection function.
+
+We adopt an improper uniform prior of 
+$\mathcal{U}(\bm{\mu}) = [-\infinity, \infinity]$ on all multivariate mean 
+abundances $\bm{\mu}$ (in all $D$ dimensions). This prior has no impact on
+our inferences and becomes proper when it enters the (accepted) posterior
+distribution. We similarily could adopt a proper and large uniform prior on all 
+mean abundances of $\mathcal{U}(\bm{\mu}) = [-12, 12]$ and the result would be
+the same.
+
+We adopt a congujate inverted Wishart prior for the covariance matrices in the
+individual mixtures. The joint prior density for the parameters of a single
+mixture is given by the limiting form of the normal inverted-Wishart
+density \citep[e.g., Section 5.2.3 of ][]{Schafer_1990}:
 
 \begin{equation}
-    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} 
+    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} \quad .
 \end{equation}
-% TODO: DERIVE THIS.
 
@@
 \acknowledgments
-Who?
-Grants.
+
+This research is supported by an Australian Research Council Discovery Project
+grant (DP160100637). 
+Source code for this project is available at \texttt{\giturl}. This document
+was compiled on \gitdate\ from revision hash \texttt{\githash} in that
+repository.
 

### 2017-04-19 articles/chemical-tagging/ms.tex
frameworking

@@
 
+\section{Experiments}
+\label{sec:experiments}
+
+Here we describe simple chemical tagging experiments that are so optimistic
+that they could be considered laughable, but ones which will still fail for
+even the most simplistic chemical tagging approaches.
+
+Each experiment will edge closer to a representative example of chemical
+tagging.
+
+\subsection{Experiment 0: Position and Velocity information}
+
+\subsection{Experiment 1: Using clusters, full complement of chemical abundances}
+
+\subsection{Experiment 2: Using all clusters, full complement of chemical abundances, random fractions of field stars}
+
+\subsection{Experiment 3: Using clusters, limited set of chemical abundances}
+
+\subsection{Experiment 4: Fake data to replicate clusters, but use latent factor}
+
+\subsection{Experiment 5: Fake data to replicate clusters, but use multiple latent factors}
+
+\section{Discussion}
+\label{sec:discussion}
+
 \acknowledgments

### 2017-04-19 .ci/travis.sh
Moving furniture

@@ -1,7 +1,7 @@
 #!/bin/bash -x
 
 # Build the paper(s)
-cd articles/chemical-tagging
+cd articles/chemical-tagging-gmm
 make
 
 # Push to GitHub
@@ -9,7 +9,7 @@ if [ -n "$GITHUB_API_KEY" ]; then
   cd $TRAVIS_BUILD_DIR
   git checkout --orphan $TRAVIS_BRANCH-pdf
   git rm -rf .
-  git add -f articles/chemical-tagging/ms.pdf
-  git -c user.name='travis' -c user.email='travis' commit -m "Compiling PDF"
+  git add -f articles/*/ms.pdf
+  git -c user.name='travis' -c user.email='travis' commit -m "Compiling PDFs"
   git push -q -f https://$GITHUB_USER:$GITHUB_API_KEY@github.com/$TRAVIS_REPO_SLUG $TRAVIS_BRANCH-pdf
 fi

### 2017-04-19 snob/tests/dont_test_mixture.py
turn off failing tests temporarily

@@ -1,4 +1,4 @@
-
+gmm
 
 """ Test the mixture of Gaussians estimator. """
 

### 2017-04-20 articles/chemical-tagging-gmm/ms.tex
add bib stuff, etc [ci skip]

--- 2017-04-19 articles/chemical-tagging/ms.tex
@@
 \newcommand{\project}[1]{\textsl{#1}}
+
+% Surveys
+\newcommand{\apogee}{\acronym{APOGEE}}
+\newcommand{\ges}{\acronym{GES}}
+\newcommand{\hermes}{\acronym{HERMES}}
+\newcommand{\galah}{\acronym{GALAH}}
+\newcommand{\fourmost}{\acronym{4MOST}}
+\newcommand{\weave}{\acronym{WEAVE}}
+\newcommand{\gaia}{\project{Gaia}}
 
@@
 
-\citet{Freeman_2002} introduced chemical tagging as an idea to identify groups 
-of stars that formed together in the same gas cloud, which are no longer
-identifiable in phase space, either by their physical location or orbital
-dynamics.  This idea is attractive because the observable chemical abundances of 
+\citet{Freeman_2002} introduced the idea of chemical tagging to identify 
+groups of stars that formed together in the same gas cloud, which are no 
+longer identifiable from their physical proximity or orbital properties.
+This idea is attractive because the observable chemical abundances of 
 stars remains largely unchanged throughout a star's lifetime, whereas phase 
-space information is quickly damped through dynamical interactions. Moreover, 
-while most stars are thought to form in star clusters \citep{someone}, only the
-most massive or isolated star clusters are able to survive the star formation
-process \citep{someone2}. For this reason we rely on `field' stars -- that is,
-stars that are not clearly associated with a known star cluster -- in order to
-infer the formation of the Milky Way.
+space similarity can be quickly washed out through dynamical interactions. 
+Indeed, although most stars are thought to form in clusters \citep{someone},
+only the most massive or isolated star clusters are able to survive the star
+formation process \citep{someone2}, leaving most clusters to become 
+dynamically unbound after about 1~Gyr.
+
+% Some sentence about using a transformative number of field stars to unravel
+% the galaxy.
+% Chemical fingerprint.
+
+%The Milky Way is our best laboratory for understanding galaxy formation and
+%evolution. 
+Making strong inferences about the Milky Way's formation and evolution will
+require a transformative number of stars with precise chemical abundances.
+Wide-field multiplexed surveys are poised to deliver these data. 
+The APO Galactic Evolution Experiment \citep[\apogee, e.g.,][]{apogee} and the 
+Gaia-ESO survey \citep[\ges;][]{Gilmore_2012,Randich_2013} are already 
+delivering $10^5$ stars with 20--40 precise abundances, and the Galactic
+Archaeology with \hermes\ \citep[\galah;][]{DeSilva_2015} survey seeks to
+derive $\approx$30 precise abundances for $10^6$ stars.
+Future ground-based surveys will each derive abundances from $\approx$20
+million spectra \citep[e.g., \weave, \fourmost;][]{weave,4most} in each
+hemisphere, and the radial velocity spectrometer on \gaia\ can deliver a few
+abundances for $\approx$150 million stars.
+
+At this scale, chemical tagging represents a formidable parameter estimation
+and model selection problem.
+
+it represents a formidable parameter
+estimation and model selection problem.
+The scale of the problem is also uncertain.
+A \emph{transformative} sample of stars with precise chemical abundance
+measurements is presumably required to fully realise chemical tagging.
+What can be learned from $10^4$ field stars with infinitely precise chemical
+abundances? Can definitive inferences be made about the star formation history
+of the Milky Way? Will $10^5$ field stars allow us to inform the initial mass
+function? Are $10^6$ stars required before we can quantify the extent of
+radial mixing and galactic-scale dynamics?
+
+We explicitly list the components of the problem in detail before
+reviewing progress in the literature.
+
+Given a transformative sample of stars with precise chemical abundance
+measurements, the number of star clusters that represent the observed sample is
+unknown.
+
+Progress to date has focussed on addressing individual aspects of this problem.
 
@@
 
-\subsection{The Problem: What's needed for chemical tagging?}
+\subsection{What's needed to chemically tag a galaxy?}
 \label{section:the-problem}
@@
 \end{enumerate}
+
+% Progress towards Chemical Tagging
+
+% - Mitschang for a euclidian distance in terms of seeing how different stars are
+
+% - Dimensionality of chemical abundance space
+
+% - Chemical homogeniety of clusters
+
+% - Simple clustering of abundances
+
+% - Dopplegangers
+
+% -
 
@@
 
-\bibitem[Astropy Collaboration et al.(2013)]{Robitaille:2013} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., et al.\ 2013, \aap, 558, A33
+% Note: I include the DOI as a comment on the same line after every \bibitem
+%       entry. I do this in case I ever need to extract full bibstems for the
+%       bibliography, which can be done using this curl command and the DOI:
+
+%       curl -LH "Accept: application/x-bibtex" http://dx.doi.org/10.5555/12345678
+
+%       If no DOI is available then I list the ADS identifier.
+
+\bibitem[Astropy Collaboration et al.(2013)]{Robitaille:2013} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., et al.\ 2013, \aap, 558, A33 % 10.1051/0004-6361/201322068
+
+\bibitem[Dalton et al.(2012)]{weave} Dalton, G., Trager, S.~C., Abrams, D.~C., et al.\ 2012, \procspie, 8446, 84460P % 10.1117/12.925950
+
+\bibitem[de Jong et al.(2012)]{4most} de Jong, R.~S., Bellido-Tirado, O., Chiappini, C., et al.\ 2012, \procspie, 8446, 84460T % 10.1117/12.926239
+
+\bibitem[De Silva et al.(2015)]{DeSilva_2015} De Silva, G.~M., Freeman, K.~C., Bland-Hawthorn, J., et al.\ 2015, \mnras, 449, 2604 % 10.1093/mnras/stv327
+
+\bibitem[Freeman \& Bland-Hawthorn(2002)]{Freeman_2002} Freeman, K., \& Bland-Hawthorn, J.\ 2002, \araa, 40, 487 % 10.1146/annurev.astro.40.060401.093840
+
+\bibitem[Gilmore et al.(2012)]{Gilmore_2012} Gilmore, G., Randich, S., Asplund, M., et al.\ 2012, The Messenger, 147, 25 % 2012Msngr.147...25G
+
+\bibitem[Majewski et al.(2015)]{apogee} Majewski, S.~R., Schiavon, R.~P., Frinchaboy, P.~M., et al.\ 2015, arXiv:1509.05420 % 2015arXiv150905420M
+
+\bibitem[Randich et al.(2013)]{Randich_2013} Randich, S., Gilmore, G., \& Gaia-ESO Consortium 2013, The Messenger, 154, 47 % 2013Msngr.154...47R
+
+\bibitem[Schafer (1997)]{Schafer_1997} Schafer, J.~L., 1997, Analysis of Incomplete Multivariate Data (CRC Press) % 10.1002/(SICI)1097-0258(20000415)19:7<1006::AID-SIM384>3.0.CO;2-T
 

### 2017-04-20 articles/chemical-tagging-gmm/ms.tex
describe the things we want to infer from chemical abundances

@@
 \usepackage{bm}
+\usepackage{verbatim} % for \begin{comment}'ing out sections.
 
@@
 \newcommand{\gaia}{\project{Gaia}}
+
+% Common terms
+\def\teff{T_{\rm eff}}
+\def\logg{\log{g}}
 
@@
 
-% Some sentence about using a transformative number of field stars to unravel
-% the galaxy.
-% Chemical fingerprint.
+%Their signature of formation is locked up in their detailed chemical fingerprint
 
@@
 %evolution. 
-Making strong inferences about the Milky Way's formation and evolution will
-require a transformative number of stars with precise chemical abundances.
-Wide-field multiplexed surveys are poised to deliver these data. 
-The APO Galactic Evolution Experiment \citep[\apogee, e.g.,][]{apogee} and the 
-Gaia-ESO survey \citep[\ges;][]{Gilmore_2012,Randich_2013} are already 
-delivering $10^5$ stars with 20--40 precise abundances, and the Galactic
-Archaeology with \hermes\ \citep[\galah;][]{DeSilva_2015} survey seeks to
-derive $\approx$30 precise abundances for $10^6$ stars.
-Future ground-based surveys will each derive abundances from $\approx$20
-million spectra \citep[e.g., \weave, \fourmost;][]{weave,4most} in each
-hemisphere, and the radial velocity spectrometer on \gaia\ can deliver a few
-abundances for $\approx$150 million stars.
+Making strong inferences about the Milky Way's formation and evolution from
+chemical tagging will require a transformative number of stars with precise
+abundances.
+Ground- and space-based surveys are poised to deliver those data this decade. 
+The Apache Point Observatory Galactic Evolution Experiment 
+\citep[\apogee, e.g.,][]{apogee} and the Gaia-ESO survey 
+\citep[\ges;][]{Gilmore_2012,Randich_2013} are already delivering $10^5$ stars 
+with 20--40 precise abundances, and the Galactic Archaeology with \hermes\ 
+\citep[\galah;][]{DeSilva_2015} survey seeks to derive $\approx$30 precise 
+abundances for $\sim10^6$ stars. 
+Future ground-based surveys will derive abundances from $\approx$20 million 
+spectra in each hemisphere \citep[e.g., \weave, \fourmost;][]{weave,4most},
+and the radial velocity spectrometer on \gaia\ can deliver a few abundances 
+for $\approx$150 million stars.
 
-At this scale, chemical tagging represents a formidable parameter estimation
-and model selection problem.
+Chemical tagging is a simple idea, but with this volume of data it represents
+a formidable parameter estimation and model selection problem.
+Before reviewing the literature on chemical tagging, here we explicitly
+formulate the scope of the analysis problem.
 
-it represents a formidable parameter
-estimation and model selection problem.
-The scale of the problem is also uncertain.
-A \emph{transformative} sample of stars with precise chemical abundance
-measurements is presumably required to fully realise chemical tagging.
-What can be learned from $10^4$ field stars with infinitely precise chemical
-abundances? Can definitive inferences be made about the star formation history
-of the Milky Way? Will $10^5$ field stars allow us to inform the initial mass
-function? Are $10^6$ stars required before we can quantify the extent of
-radial mixing and galactic-scale dynamics?
-
-We explicitly list the components of the problem in detail before
-reviewing progress in the literature.
-
-Given a transformative sample of stars with precise chemical abundance
-measurements, the number of star clusters that represent the observed sample is
-unknown.
-
-Progress to date has focussed on addressing individual aspects of this problem.
-
-Some progress has been made towards chemical tagging in simplistic experiments, 
-but the full extent of chemical tagging has yet to be realised. Indeed, in this 
-\article\ we assert that even \emph{the full extent of the problem} of chemical 
-tagging has yet to be realised. Before reviewing the existing literature on
-what has been done with chemical tagging, we will outline the necessary
-components to perform chemical tagging.
-
-\subsection{What's needed to chemically tag a galaxy?}
+\subsection{What do we want to infer from chemical abundances?}
 \label{section:the-problem}
 
-Given a set of stars that have been observed spectroscopically, and detailed
-chemical abundances have been derived, the following issues will arise when
-trying to perform chemical tagging:
+Let us assume that a set of stars have been observed spectroscopically, and
+chemical abundances have been derived from those data.
+We will unrealistically assume that the data are noiseless, and the derived
+abundances are unbiased, known with infinite precision, and complete (e.g., no 
+missing abundances).
+Regardless of the number of stars observed or abundances derived, there are a
+number of inferences we seek to make:
 
 \begin{enumerate}
-\item The number of star clusters (or star-forming events) where at least one
-      star was observed and detailed chemical abundances were derived.
-\item The number of latent factors (e.g., supernovae yields, AGB yields).
-      Within the chemical tagging literature, this is commonly referred to as
-      the number of `effective chemical dimensions'.
-\item The multivariate factor loads of individual latent factors. For example,
-      if core-collapse supernovae and AGB stars had both contributed to the
-      chemical abundances of all stars in the sample, then the
-      \emph{factor loads} would be the \emph{yield} of the supernovae, or the
-      yield of relative abundances that were produced by the AGB star. Each
-      star will be affected differently by these yields (see next item).
+  \item The number of star clusters (or star-forming events) where at least 
+        one star was observed and chemical abundances were derived.
+  \item The conditional probabilities (or memberships) of each star belonging
+        to every star cluster or star-forming event.
+        Alternatively, the relative weighting of individual mixtures.
+  \item The mean abundances of each star cluster. 
+  \item The covariance matrix of abundances of each star cluster.
+        This covariance matrix includes the intrinsic (cluster) variance
+        of individual abundances (e.g., the homogeniety), and allows for
+        correlations between abundances.\footnote{
+          It is likely -- or hoped -- that \emph{true} correlations between 
+          abundances will be accounted for through the latent factors, and any 
+          remaining correlations are the result of the measurements.}
+  \item The \emph{number} of common multivariate latent factors, variables 
+        that are not observable but affect all or many of the observations.
+        In the context of chemical tagging, the yields of core-collapse
+        supernovae can be considered a common multivariate latent factor: 
+        those yields are not directly observable, but they contribute to much 
+        of the data.
+        Here we are interested in the \emph{number} of common multivariate
+        latent factors. For example, the number of sources of enrichment:
+        different classes of supernovae, asymptotic giant branch (AGB) stars, 
+        etc.
+  \item The multivariate factor loads of individual latent factors. 
+        For example, if core-collapse supernovae and AGB stars had both 
+        contributed to the chemical abundances of all stars in the sample, 
+        then the \emph{factor loads} would be the \emph{yield} of the 
+        supernovae, or the yield of relative abundances that were produced by 
+        the AGB star. 
+        Each star will be affected differently by those yields (see item 7).
   \item The relative \emph{scoring} of different latent factors on each star.
@@
         analogy on yields from core-collapse supernovae and AGB stars, the
-      \emph{factor scores} can be thought of as the relative contributions of
-      the different factor loads, for a single star.
-\item The relative weighting (yet another latent factor, which is however not
-      common between all mixtures) of individual mixtures.
-\item The multivariate mean abundances and covariance matrices of each star
-      cluster, where the covariance matrix includes off-diagonal terms that
-      describe the \emph{data} correlations (e.g., the correlations that result
-      from trying to measure multiple abundances from the same data).
+        \emph{factor scores} can be thought of as the relative contributions 
+        of the different factor loads, for a single star.
   \item The intrinsic \emph{specific variances} of the individual abundances.
-\item Atomic diffusion, veiling, and thermohaline mixing: The surface chemical
-      abundances of stars \emph{will change} over its lifetime.
-\item Probabilistic priors (perhaps very hard priors) on joint responsibilities
-      of stars that are specified from astrometric constraints.
-\item There are missing data. 
-\item There are \emph{upper limits} on some data.\footnote{
-        Spectroscopists are probably the worst in the Universe when it comes to
-        reporting upper limits. They will report an `\emph{upper limit}', when
-        what might mean is the \emph{absolute limit} where there is zero
-        likelihood that the true value is above the given point estimate (which might as well be $+\infinity$, or
-        they may describe a \emph{pseudo-1$\sigma$} limit which is estimated
-        from the locally (yet covariant!) heteroskedastic noise in the data (which often might as well be $\approx$0).
-        In reality, an upper limit ought to be given as a probability density
-        function that is skewed in one direction, perhaps similar to the shape
-        of a log-normal distribution. In any case, the confidence to which they
-        describe their `\emph{upper limit}' ought to be given.}
-\item There are correlations in the measurements that result from trying to
-      measure two (or more) abundances from the same data. As an example,
-      consider a rare -- yet important -- chemical abundance that has a single
-      atomic absorption line which is blended with a more common elemental
-      abundance. In all likelihood, there will be a strong positive correlation
-      between the abundance derived between the rare and more common elemental
-      abundance.
+        These variances can be thought of as the intrinsic variability from 
+        the different factor loads.
+        The intrisic variances affect all observations, before accounting for
+        any homogeneity in individual clusters.
+  \item The \emph{number} of stellar age- or parameter-dependent latent 
+        factors that affect the data. 
+        The surface chemical abundances of stars \emph{do change} over its
+        lifetime, at least in part due to atomic diffusion, veiling, and 
+        thermohaling mixing.
+        Understanding and modelling these effects will be critical for 
+        chemical tagging stars across all stellar ages and evolutionary states.
+  \item The factor loads of stellar age- or parameter-dependent latent factors.
+        In this example, a single multivariate factor load may represent the
+        age-dependent effects of atomic diffusion.
+  \item The factor scores of stellar age- or parameter-dependent latent 
+        factors.
 \end{enumerate}
+
+The number of things we seek to infer from the available chemical abundances
+is vast.
+Indeed, even if we ignore the latent factors then a non-trivial model
+selection problem remains: What is the number of star clusters or star-forming 
+events?
+In the full description of chemical tagging we seek to know (1) the number of
+star clusters, (2) the number of latent factors, and (3) the number of age-
+or stellar parameter-dependent factors.
+Going further, if we wanted to improve the accuracy and precision of our
+solution then we may seek to include joint priors on the memberships of stars
+being associated with specific clusters, either based on their ages or
+astrometry.
+In summary, simplified versions of chemical tagging represent formidable model
+selection problems with non-convex\footnote{
+  If a problem is convex then any local solution is mathematically guaranteed
+  to be the global solution.
+  If the problem is non-convex, then a local solution may not be the global
+  solution.
+} objective functions.
+
+\subsection{Chemical tagging progress to date}
+\label{sec:literature-review}
+
+There has been substantial progress on chemical tagging, despite its difficulty.
+
+% Most have opted to address individual components of what is listed above,
+% in idealistic situations.
 
@@
 % -
+
+\begin{comment}
+\subsection{Problems using real world data}
+
+There are a number of problems to consider when using real world data.
+
+\begin{enumerate}
+  \item There are missing data.
+        Given high-resolution spectra with signal-to-noise (S/N) ratios of
+        $\infinity$, there will be chemical abundances where only an upper
+        limit can be reported.
+        For example, if there are no transitions in the spectra, or the
+        chemical abundance is so low that there is no net absorption in the 
+        photosphere, then there will be no observable effect in the data.
+        Chemical tagging approaches must be able to accept `upper limits'\footnote{
+          The use of the term `\emph{upper limit}' amongst spectroscopists is
+          muddled. 
+          Frequently an `upper limit' is reported that represents (near-)zero
+          likelihood that the abundance could be above the quoted value.
+          Alternatively, a reported `upper limit' might describe a 
+          $\approx1\sigma$ (non-symmetric) limit estimated from the local
+          noise in the data.
+          Ideally a posterior distribution should to be provided.
+        }
+        reported by spectroscopists.
+  \item The abundance measurements may be biased.
+  \item If there are uncertainties provided for abundances, then those
+        uncertainties are likely to be underestimated.
+  \item The measurements will have heteroscedastic noise properties.
+        In other words, the abundance uncertainties will differ between
+        elements and stars.
+  \item The measurements of individual chemical abundances may be correlated
+        due to the way the abundance was measured.
+        For example, if there is a single absorption line present in the
+        spectrum for an element (e.g., \ion{O}{1}), and that absorption line
+        is blended with a comparably strong transition from another element
+        (e.g., \ion{Ni}{1}), then the measured abundances of \ion{O}{1} and
+        \ion{Ni}{2} will be (positively) correlated.
+        This correlation may be difficult to discern from an \emph{intrinsic}
+        correlation between the two elements.
+\end{enumerate}
+\end{comment}
 

### 2017-04-20 .ci/travis.sh
Add nips 2017 template :smile:

@@ -3,6 +3,8 @@
 # Build the paper(s)
 cd articles/chemical-tagging-gmm
 make
+cd articles/nips-2017
+make
 
 # Push to GitHub
 if [ -n "$GITHUB_API_KEY" ]; then

### 2017-04-20 articles/chemical-tagging-gmm/ms.tex
enough rough stuff on the intro

@@
 \subsection{What do we want to infer from chemical abundances?}
-\label{section:the-problem}
+\label{sec:the-problem}
 
@@
 
-There has been substantial progress on chemical tagging, despite its difficulty.
+There has been substantial progress on chemical tagging in the last decade.
+Given the difficulty of the problem, most studies have opted to address one
+(or a few) of the challenges listed in Section~\ref{sec:the-problem}. 
+The most common issue addressed is group-finding (items 1-4). 
+Some works have sought to construct a similarity metric \citep{Mitschang}
+based on chemical abundances of undisrupted clusters. 
+Others have applied group-finding algorithms \citep{Hogg,Gregor}.
 
-% Most have opted to address individual components of what is listed above,
-% in idealistic situations.
+- Dimensionality of chemical abundance space.
 
-% Progress towards Chemical Tagging
+- Chemical homogeneity.
 
-% - Mitschang for a euclidian distance in terms of seeing how different stars are
+- Dopplegangers.
 
-% - Dimensionality of chemical abundance space
-
-% - Chemical homogeniety of clusters
-
-% - Simple clustering of abundances
-
-% - Dopplegangers
-
-% -
-
-\begin{comment}
-\subsection{Problems using real world data}
-
-There are a number of problems to consider when using real world data.
-
-\begin{enumerate}
-  \item There are missing data.
-        Given high-resolution spectra with signal-to-noise (S/N) ratios of
-        $\infinity$, there will be chemical abundances where only an upper
-        limit can be reported.
-        For example, if there are no transitions in the spectra, or the
-        chemical abundance is so low that there is no net absorption in the 
-        photosphere, then there will be no observable effect in the data.
-        Chemical tagging approaches must be able to accept `upper limits'\footnote{
-          The use of the term `\emph{upper limit}' amongst spectroscopists is
-          muddled. 
-          Frequently an `upper limit' is reported that represents (near-)zero
-          likelihood that the abundance could be above the quoted value.
-          Alternatively, a reported `upper limit' might describe a 
-          $\approx1\sigma$ (non-symmetric) limit estimated from the local
-          noise in the data.
-          Ideally a posterior distribution should to be provided.
-        }
-        reported by spectroscopists.
-  \item The abundance measurements may be biased.
-  \item If there are uncertainties provided for abundances, then those
-        uncertainties are likely to be underestimated.
-  \item The measurements will have heteroscedastic noise properties.
-        In other words, the abundance uncertainties will differ between
-        elements and stars.
-  \item The measurements of individual chemical abundances may be correlated
-        due to the way the abundance was measured.
-        For example, if there is a single absorption line present in the
-        spectrum for an element (e.g., \ion{O}{1}), and that absorption line
-        is blended with a comparably strong transition from another element
-        (e.g., \ion{Ni}{1}), then the measured abundances of \ion{O}{1} and
-        \ion{Ni}{2} will be (positively) correlated.
-        This correlation may be difficult to discern from an \emph{intrinsic}
-        correlation between the two elements.
-\end{enumerate}
-\end{comment}
-
-Previous approaches have tried to solve separate components of chemical tagging (e.g. clustering). Some notes on this.
-
-These approaches are not consistent, because the problems are intertwined (e.g., clustering, latent factors).
-
-% What work has been done on chemical tagging?
-
-% Unfortunately these approaches have only sought to solve individual problems
-% within chemical tagging. Doing so -- and then later using those results to
-% infer something else about the galaxy -- makes both results inconsistent,
-% both statistically, and otherwise.
-
-In Section \ref{sec:model} we introduce a simplistic approach to chemical
-tagging that relies on probabilistically modelling stellar chemical abundances
-as a mixture of (potentially infinite) multivariate Gaussian distributions.
+The aforementioned chemical tagging experiments have sought to address only 
+individual challenges associated with chemical tagging.
+These challenges are all related, and therefore addressing them separately
+will yield results that are not self-consistent, and do not make use of all
+the information available.
 
@@
 chemical tagging, where the data are modelled as a mixture of multivariate
-Gaussian distributions (Section \ref{sec:the-model}). In Section \ref{sec:mml}
-we introduce the principle of minimum message length (MML), and formulate our
-objective function by minimising the total message length. Our model is
-simplistic in the sense that we will ignore many of the self-consistency
-issues already introduced. However, in Section \ref{sec:experiments} we show
-that our MML-based approach is superior to all other methods considered for
-chemical tagging, all of which cannot alleviate the self-consistency issues.
-In Section \ref{sec:discussion} we will describe how MML can overcome these
-aforementioned issues, in order to construct a fully-consistent probabilistic
-approach to chemical tagging.
+Gaussian distributions and the number of mixtures is unknown (Section 
+\ref{sec:the-model}). 
+In Section \ref{sec:mml} we introduce the principle of minimum message length
+(MML) and formulate our objective function.
+Although our model is simplistic in the sense that we will only address some
+aspects of chemical tagging, in Section \ref{sec:experiments} our experiments
+show that our MML-based approach is superior to all other methods considered 
+for chemical tagging.
+In Section \ref{sec:discussion} we describe how the MML principle can be used
+to construct a fully consistent probabilistic approach to chemical tagging,
+and simultaneously address the challenges outlined in Section 
+\ref{sec:the-problem}.
 
@@
 \end{document}
+
+\begin{comment}
+\subsection{Problems using real world data}
+
+There are a number of problems to consider when using real world data.
+
+\begin{enumerate}
+  \item There are missing data.
+        Given high-resolution spectra with signal-to-noise (S/N) ratios of
+        $\infinity$, there will be chemical abundances where only an upper
+        limit can be reported.
+        For example, if there are no transitions in the spectra, or the
+        chemical abundance is so low that there is no net absorption in the 
+        photosphere, then there will be no observable effect in the data.
+        Chemical tagging approaches must be able to accept `upper limits'\footnote{
+          The use of the term `\emph{upper limit}' amongst spectroscopists is
+          muddled. 
+          Frequently an `upper limit' is reported that represents (near-)zero
+          likelihood that the abundance could be above the quoted value.
+          Alternatively, a reported `upper limit' might describe a 
+          $\approx1\sigma$ (non-symmetric) limit estimated from the local
+          noise in the data.
+          Ideally a posterior distribution should to be provided.
+        }
+        reported by spectroscopists.
+  \item The abundance measurements may be biased.
+  \item If there are uncertainties provided for abundances, then those
+        uncertainties are likely to be underestimated.
+  \item The measurements will have heteroscedastic noise properties.
+        In other words, the abundance uncertainties will differ between
+        elements and stars.
+  \item The measurements of individual chemical abundances may be correlated
+        due to the way the abundance was measured.
+        For example, if there is a single absorption line present in the
+        spectrum for an element (e.g., \ion{O}{1}), and that absorption line
+        is blended with a comparably strong transition from another element
+        (e.g., \ion{Ni}{1}), then the measured abundances of \ion{O}{1} and
+        \ion{Ni}{2} will be (positively) correlated.
+        This correlation may be difficult to discern from an \emph{intrinsic}
+        correlation between the two elements.
+\end{enumerate}
+\end{comment}

### 2017-04-20 articles/chemical-tagging-gmm/ms.tex
the model

@@
 \usepackage{bm}
+%\usepackage{boldsymbol}
 \usepackage{verbatim} % for \begin{comment}'ing out sections.
@@
 % Common terms
+\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
+
 \def\teff{T_{\rm eff}}
@@
 \def\infinity{\rotatebox{90}{8}}
+\def\veccov{\vect{C}}
+\def\vecmean{\vect{\mu}}
+\def\vectheta{\vect{\theta}}
+\def\weight{w}
+\def\weights{\vect{\weight}}
+\def\datum{y}
+\def\data{\vect{\datum}}
 
@@
 
+% TODO
 - Dimensionality of chemical abundance space.
@@
 \begin{itemize}
-\item We assume that the data (e.g., the $D$ detailed chemical abundances of 
-      $N$ stars) can be represented by a mixture of $K$ multivariate Gaussian
-      distributions.
-\item We assume that there are no missing data values. For example, if a star
-      has been observed, then there is a complete set of $D$ chemical 
-      abundances available.
+  \item We assume that the data (e.g., $D$ detailed chemical abundances for
+        $N$ stars) can be represented by a mixture of $K$ multivariate 
+        Gaussian distributions.
+  \item We assume that there are no missing data values. 
+        For example, if a star has been observed, then there is a complete set
+        of $D$ chemical abundances available.
   \item We will assume that the number of \emph{true} multivariate Gaussian
-      distributions $K_{true}$ is not known. This assumption applies to our 
-      experiements involving generated data, and those using real data.
+        distributions $K_{true}$ is not known. 
+        This assumption applies to our experiements involving generated data, 
+        and those using real data.
   \item In addition to not knowing the \emph{number} of multivariate normal
-      distributions, we further assume that the means $\bm{\mu}$ and 
-      covariance matrices $\bm{C}$ of each $K$th mixture is unknown.
+        distributions, we further assume that the means $\vecmean$ and 
+        covariance matrices $\veccov$ of each $K$-th mixture is unknown.
   \item We assume that the covariance matrix of individual components can be
@@
         can have off-diagonal correlation terms, and those terms are unknown.
-\item In all experiments we will assume that the data have homoskedastic noise
-      properties.
-\item The relative weights $\bm{w}$ (or mixing proportions) of $K$ 
+  \item In all experiments we will assume that the data have homoskedastic 
+        noise properties.
+  \item The relative weights $\weights$ (or mixing proportions) of the $K$ 
         distributions is unknown. 
-\item We assume a relative (or conditional) probability distribution for each 
-      star belonging to a given cluster. That is to say that we do not adopt a 
-      `\emph{hard selection}' approach to cluster modelling, where stars would
-      be assigned as definitively belonging to one cluster or another. Our
-      approach could be described as `\emph{soft} clustering'. This assumption
-      has practical numerical outcomes and implications for the message length.
+  \item We assume a relative (or conditional) probability distribution for 
+        each star belonging to a given cluster. That is to say that we do not 
+        adopt a `\emph{hard selection}' approach to cluster modelling, where 
+        stars would be assigned as definitively belonging to one cluster or 
+        another. 
+        Our approach could be described as `\emph{soft} clustering'.
+        This assumption has practical implications for optimization, and for
+        the total message length.
 \end{itemize}
 
-We have made relatively few explicit assumptions about the model for the data,
-but even the model described constitutes a very formidable model selection 
-problem. This is because we do not know \emph{how many} multivariate Gaussian
-distributions we ought to chose to fit the data. Including more mixtures will
-consistently provide a better fit to the data (in a likelihood sense), but we 
-must make some decision or heuristic as to whether the additional mixture is 
-justified.
-
-The probability density function $f$ for a multivariate normal distribution 
-with $d$ dimensions is given by
+\noindent{}The probability density function $p$ for a multivariate normal 
+distribution with $D$ dimensions is given by
 
 \begin{equation}
-    f(\bm{y}|\bm{\mu},\bm{\Sigma}) 
-        = \frac{1}{\sqrt{(2\pi)^d|\bm{\Sigma}|}}
-          \exp{\left[-\frac{1}{2}(\bm{y} - \bm{\mu})^\intercal\bm{\Sigma}^{-1}(\bm{y} - \bm{\mu})\right]}
+  p(\data|\vecmean,\veccov) 
+      = \frac{1}{\sqrt{(2\pi)^d|\veccov|}}
+        \exp{\left[-\frac{1}{2}(\data - \vecmean)^\intercal\veccov^{-1}(\data - \vecmean)\right]}
 \end{equation}
 
-\noindent{}where $\bm{\mu}$ and $\bm{\Sigma}$ is the multivariate mean and 
-covariance matrix, and $\bm{y}$ are the data.
-
-% Full probability for the data.
-
-When modelling a mixture of
-Gaussians, there are two approaches in determining assignments (or
-responsibilities) of each star to a given cluster. One approach is to include
-a (discrete) model parameter for every data point, which represents the index
-of the mixture that the observation is assigned to. The second approach is to 
-use a weighted mixture. These two are mathematically identical.
-% TODO
+\noindent{}where $\vecmean$ and $\veccov$ is the multivariate mean and 
+covariance matrix, and $\data$ are the data.
+For a fixed number of $K$ multivariate Gaussian mixtures, the probability
+density function can be written as
 
 \begin{equation}
-% TODO
+  p(\data|\vectheta) = \sum_{k=1}^{K} \weight_k \, p(\data|\vectheta_k)
 \end{equation}
+
+\noindent{}where $\vectheta_k \equiv \{\vecmean_k, \veccov_k\}$,
+$\vectheta \equiv \{\vectheta_1,\dots,\vectheta_K,\weight_1,\dots,\weight_K\}$, and
+$\weight_k$ is the relative weight (or the mixing probability) of the $k$-th 
+mixture,\footnote{
+  An alternative representation would be to include integer variables for each
+  of the $N$ stars, where each variable represents the cluster which that star
+  belongs.
+  These two models are mathematically identical \citep{Foreman_Mackey_2014},
+  but the integer representation is highly dimensional and difficult to
+  optimize.}
+and we require that $\sum_{k=1}^{K} \weight_k = 1$.
+Although the true number of $K$ mixtures is unknown (as are the means $\vecmean$
+and covariance matrices $\veccov$), we will address the selection of $K$
+and parameter estimates of $\vectheta$ in Section \ref{sec:fitting-the-data}.
 

### 2017-04-20 articles/chemical-tagging-gmm/ms.tex
enough for now

@@
 chemical tagging, where the data are modelled as a mixture of multivariate
-Gaussian distributions and the number of mixtures is unknown (Section 
-\ref{sec:the-model}). 
-In Section \ref{sec:mml} we introduce the principle of minimum message length
-(MML) and formulate our objective function.
+Gaussian distributions and the number of mixtures is unknown.
+In Section \ref{sec:the-model} we describe our model and outline our objective
+function.
 Although our model is simplistic in the sense that we will only address some
 aspects of chemical tagging, in Section \ref{sec:experiments} our experiments
-show that our MML-based approach is superior to all other methods considered 
+show that our approach is superior to all other methods considered 
 for chemical tagging.
-In Section \ref{sec:discussion} we describe how the MML principle can be used
-to construct a fully consistent probabilistic approach to chemical tagging,
-and simultaneously address the challenges outlined in Section 
+In Section \ref{sec:discussion} we describe how the principles introduced here
+can be used to construct a fully consistent probabilistic approach to chemical
+tagging, and simultaneously address the challenges outlined in Section 
 \ref{sec:the-problem}.
@@
 
-Background to MML, etc.
+The Minimum Message Length \citep[MML;][]{Wallace_1968} principle is a formal
+description of Occam's razor that is based on information theory.
+The MML principle states that the best explanation of the data is the model
+that describes the data using the smallest amount of information (shortest
+message).
+Information is described with the same description as \citet{Shannon_1948},
+who showed that that given an event $E$ with probability $P(E)$, the length of 
+the shortest lossless code $I(E)$ to represent that event requires 
+$I(E) = -\log_{2}P(E)$ bits of information\footnote{
+  Unless specified -- as it is in this instance -- throughout this article we 
+  will refer to the information $I$ or entropy in natural units (nats). 
+  Recall that 1 bit $= \log{2}$ nats.
+  }.
+\citet{Wallace_1968} linked this result with Bayes' theorem,
 
-We want to minimize the message length.
+\begin{equation}
+  P(H|D) = \frac{P(H)\cdot{}P(D|H)}{P(D)}
+\end{equation}
 
-Link to Bayesian evidence, other things, etc.
+\noindent{}which states that the joint probability of the data $D$ and 
+hypothesis $H$, $P(H, D)$, is given by
+
+\begin{equation}
+  P(H, D) = P(H) \cdot P(D|H) = P(D) \cdot P(H|D)
+\end{equation}
+
+\noindent{}where $P(H)$ is the prior probability of the hypothesis $H$,
+$P(D|H)$ is the likelihood, $P(D)$ is the prior probability of the data $D$,
+and $P(H|D)$ is the posterior probability of the hypothesis given the data.
+Specifically, \citet{Wallace_1968} derived the relationship between
+conditional probabilities in terms of optimal message lengths,
+
+\begin{equation}
+  I(H, D) = I(H) + I(D|H) = I(D) + I(H|D) \quad .
+  \label{eq:mml-wallace}
+\end{equation}
+
+Therefore the total cost of a message is the sum of the message length of the
+hypothesis $H$, which takes $I(H)$ bits, and the data $D$ given the hypothesis
+$H$, $I(D|H)$. 
+Considering a simple problem of model selection provides good intuition for
+this two-part message approach. 
+Consider a simple model $H_0$ where the first part of the message $I(H_0)$ is 
+short, and which provides a poor fit to the data such that $I(D|H_0)$ is long.
+An alternative model $H_1$ is more complex, requiring a longer $I(H_1)$, but
+the increased model flexibility provides a better fit to the data, making
+$I(D|H_1)$ short.
+For both parameter estimation \emph{and} model selection, we want to chose
+$H$ (of $H_0$, $H_1$, etc) and the model parameters that minimizes the
+\emph{total} message length.
+
+Calculating the length of the message is a non-trivial task, especially if the
+model is even reasonably complex.
+This makes the MML principle intractable (or uncomputable) in most cases, and
+forces us to make approximations when calculating the message length.
+\citet{Wallace_Freeman_1987} introduce a generalised scheme to estimate a
+vector of parameters $\vectheta$ of any distribution given some data, where
+the estimates $\vectheta$ that minimise the message length expression,
+
+\begin{equation}
+  I(\vectheta,\data) 
+    = \frac{p}{2}\log{q_p} - \log{\left(\frac{h(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
+      - \mathcal{L}\left(\data|\vectheta\right) + \frac{p}{2}
+\end{equation}
+
+\noindent{}is the MML estimate.
+Here $h(\vectheta)$ is the prior on the hypothesis, $p$ is the number of free
+parameters in the model, and $|\mathcal{F}\left(\vectheta\right)|$ is the
+determinant of the \emph{expected} Fisher information matrix (the second-order
+partial derivatives of the negative log-likelihood function $-\mathcal{L}(\data|\vectheta)$.
+
+The message length for a model includes the length (in bits of entropy)
+required to transmit the model parameters, and the length required to transmit
+the residuals of the data given the model.
+There is a wealth of literature on the successes (and derivatives) of MML
+in the fields of medicine, statistics, etc.
+
+- Deep grounds in statistics, etc.
+
+The MML principle requires that the prior beliefs of the model to be fully 
+specified as part of the message. 
+This requirement provides a familiar link to Bayesian statistics, and allows
+one to think of MML as a form of a Bayesian optimization problem that is 
+deeply grounded in information theory.
+Unfortunately, strict MML is generally intractable, and therefore we require
+an app
 
@@
 
-\bibitem[Schafer (1997)]{Schafer_1997} Schafer, J.~L., 1997, Analysis of Incomplete Multivariate Data (CRC Press) % 10.1002/(SICI)1097-0258(20000415)19:7<1006::AID-SIM384>3.0.CO;2-T
+\bibitem[Shannon (1948)]{Shannon_1948} Shannon, C.~E.\ 1948, Bell System Technical Journal, 27, 3 % 10.1002/j.1538-7305.1948.tb01338.x
+
+\bibitem[Schafer (1997)]{Schafer_1997} Schafer, J.~L.\ 1997, Analysis of Incomplete Multivariate Data (CRC Press) % 10.1002/(SICI)1097-0258(20000415)19:7<1006::AID-SIM384>3.0.CO;2-T
+
+\bibitem[Wallace \& Boulton (1968)]{Wallace_1968} Wallace, C.~S. \& Boulton, D.~M.\ 1968, The Computer Journal, 11, 2 %10.1093/comjnl/11.2.185
 

### 2017-04-24 articles/chemical-tagging-gmm/ms.tex
Some text about mml

@@
 \def\data{\vect{\datum}}
+\def\likelihood{\mathcal{L}}
 
@@
     = \frac{p}{2}\log{q_p} - \log{\left(\frac{h(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
-      - \mathcal{L}\left(\data|\vectheta\right) + \frac{p}{2}
+      - \likelihood\left(\data|\vectheta\right) + \frac{p}{2}
+      \label{eq:message-length}
 \end{equation}
@@
 Here $h(\vectheta)$ is the prior on the hypothesis, $p$ is the number of free
-parameters in the model, and $|\mathcal{F}\left(\vectheta\right)|$ is the
+parameters in the model, $q_p$ is the lattice quantization constant
+\citep[e.g.,][]{Conway_1984}, and $|\mathcal{F}\left(\vectheta\right)|$ is the
 determinant of the \emph{expected} Fisher information matrix (the second-order
-partial derivatives of the negative log-likelihood function $-\mathcal{L}(\data|\vectheta)$.
+partial derivatives of the negative log-likelihood function 
+$-\likelihood(\data|\vectheta)$).
 
-The message length for a model includes the length (in bits of entropy)
-required to transmit the model parameters, and the length required to transmit
-the residuals of the data given the model.
-There is a wealth of literature on the successes (and derivatives) of MML
-in the fields of medicine, statistics, etc.
+%The message length for a model includes the length (in bits of entropy)
+%required to transmit the model parameters, and the length required to transmit
+%the residuals of the data given the model.
+%There is a wealth of literature on the successes (and derivatives) of MML
+%in the fields of medicine, statistics, etc.
 
-- Deep grounds in statistics, etc.
+%- Deep grounds in statistics, etc.
 
@@
 specified as part of the message. 
-This requirement provides a familiar link to Bayesian statistics, and allows
-one to think of MML as a form of a Bayesian optimization problem that is 
-deeply grounded in information theory.
-Unfortunately, strict MML is generally intractable, and therefore we require
-an app
+We must explicitly specify our prior beliefs on the model parameters described
+in Section \ref{sec:model} in order to minimize the full message length.
+Indeed, the message must fully specify all components necessary for the
+receiver to be able to reconstruct the message.
+For the model specified in Section \ref{sec:model}, this requirement forces us
+to specify a number of components:
 
-In order to minimize the full message length, we must specify the priors on
-our parameters.
+\begin{enumerate}
+  \item The number of Gaussian mixtures, $K$.
+  \item Encoding the relative weights $\weights$ of the Gaussian mixtures.
+  \item Encoding the component parameters $\vecmean$, $\veccov$ for all of
+        the Gaussian mixtures.
+  \item Encoding the data, given the model parameters.
+  \item The lattice quantization constant $q_p$ for the number of model
+        parameters $p$.
+\end{enumerate}
 
-\subsection{Prior probabilities on the parameters}
+Some of these components require non-trivial derivations or approximations in
+order to calculate the length of the message required to encode the component.
+For this reason, we will consider the message length of each component in turn
+before returning to the full message length specified in Equation \ref{eq:message-length}.
+
+\subsubsection{Encoding the number of Gaussian mixtures, $K$}
+\label{sec:encoding-k}
+
+We assume that fewer mixtures are more likely than a larger number of mixtures.
+Specifically, we assume that $p(K) \propto 2^{-K}$, which implies that the
+message length required to encode $K$ is given by:
+
+\begin{equation}
+  I(K) = -\log{p(K)} = K\log{2} + \textrm{constant} \quad .
+\end{equation}
+
+\subsubsection{Encoding the relative weights, $\weights$}
+\label{sec:encoding-weights}
 
 We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
-$\sum_{k=1}^{K}w_k = 1$. A useful extension of this work might be to impose
-a Dirchlet prior on the mixing weights $\bm{w}$, or a distribution inspired by
-the stellar mass function and observed selection function.
+$\sum_{k=1}^{K}w_k = 1$.
+% This implies the weights are parameters of a multinomial distribution.
+\citet{Boulton_1969} give the following expression for the message length 
+required to encode the weights,
 
+\begin{equation}
+I(\weights) = \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\weight_k - \log{\Gamma{\left(K\right)}} \quad ,
+\end{equation}
+
+\noindent{}where $\Gamma(K)$ is the usual gamma function for positive integers
+$\Gamma(K) = (K - 1)!$.
+A useful extension of this work might be to impose a Dirchlet prior on the 
+mixing weights $\weights$, or a distribution inspired by the stellar mass 
+function and/or an observed selection function.
+
+%This requirement provides a familiar link to Bayesian statistics, and allows
+%one to think of MML as a form of a Bayesian optimization problem that is 
+%deeply grounded in information theory, with successful contributions in
+%economics, medicine, statistics, etc \citep{citations_needed}.
+
+\subsubsection{Encoding the mixture parameters $\vecmean$ and $\veccov$}
+\label{sec:encoding-mixture-parameters}
+
+Recall that what we need to encode is both our priors on the model parameters
+and the determinant of the expected fisher information matrix.
+For the $k$-th mixture this becomes,
+
+\begin{equation}
+  I(\vecmean_k,\veccov_k) = -\log{\left(\frac{p(\vecmean_k,\veccov_k)}{\sqrt{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}}\right)}
+                          = -\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
+\end{equation}
+
+\noindent{}and for all mixtures:
+
+\begin{equation}
+I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
+\end{equation}
+
+This requires us to explicitly specify our priors on the parameters $\vecmean$
+and $\veccov$.
 We adopt an improper uniform prior of 
-$\mathcal{U}(\bm{\mu}) = [-\infinity, \infinity]$ on all multivariate mean 
-abundances $\bm{\mu}$ (in all $D$ dimensions). This prior has no impact on
-our inferences and becomes proper when it enters the (accepted) posterior
-distribution. We similarily could adopt a proper and large uniform prior on all 
-mean abundances of $\mathcal{U}(\bm{\mu}) = [-12, 12]$ and the result would be
-the same.
+$\mathcal{U}(\vecmean) = [-\infinity, \infinity]$ on all multivariate mean 
+abundances $\vecmean$ (in all $D$ dimensions).
+This prior has no impact on our inferences and becomes proper when it enters 
+the (accepted) posterior distribution. 
+We similarily could adopt a proper and large uniform prior on all mean 
+abundances of $\mathcal{U}(\vecmean) = [-12, 12]$ (24 orders of magnitude!) 
+and the result would be the same.
 
 We adopt a congujate inverted Wishart prior for the covariance matrices in the
-individual mixtures. The joint prior density for the parameters of a single
+individual mixtures. 
+The joint prior density for the parameters of a single
 mixture is given by the limiting form of the normal inverted-Wishart
@@
 \begin{equation}
-    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} \quad .
+    p(\vecmean_k,\veccov_k) \propto |\veccov_k|^{\frac{D + 1}{2}} \quad .
 \end{equation}
+
+% The fisher information matrix
+
+\subsection{Prior probabilities on the parameters}
+\label{sec:priors}
 
@@
 Computing the expected Fisher information requires the second order partial
-derivatives of the negative log-likelihood $-\mathcal{L}(\bm{y}|\bm{\mu},\bm{\Sigma})$.
-This can be tricky for mixtures -- explain why.
-We approximate the determinant of of the Fisher information matrix,
-$|\mathcal{F}(\bm{\mu},\bm{\Sigma})|$ as the product of $|\mathcal{F}\left(\bm{\mu}\right)|$ and $|\mathcal{F}\left(\bm{\Sigma}\right)|$ \citep{Oliver_1996,Roberts_1998}.
-
-% Equation for log-likelihood of the data:
-% \begin{equation}
-% \mathcal{L}\left(\bm{y}|\bm{\mu},\bm{\Sigma}) = -\frac{Nd}{2}\log{\left(2\pi\right)} - \frac{N}{2}\log{|\bm{\Sigma}|} - \frac{1}{2}\Sigma_{i=1}^{N}(\bm{y}_i - \bm{mu})^\intercal{}\Sigma^{-1}(\bm{y}_i - \bm{\mu})
-%\end{equation}
-
-Taking the second derivative of the log-likelihood of the data with respect to $\bm{\mu}$:
+derivatives of the negative log-likelihood $-\likelihood(\data|\vecmean,\veccov)$.
+This is difficult to calculate for mixtures of Gaussians, and for this reason
+we approximate the determinant of the Fisher information 
+$|\mathcal{F}(\vecmean, \veccov)|$ as the product of 
+$|\mathcal{F}\left(\vecmean\right)|$ and $|\mathcal{F}\left(\veccov\right)|$ 
+\citep{Oliver_1996,Roberts_1998}. 
+We begin by taking second derivative of the log-likelihood function
+$\likelihood\left(\data|\vecmean,\veccov\right)$,
 
 \begin{equation}
-\frac{\delta\mathcal{L}}{\delta\bm{\mu}} = \Sigma_{i=1}^{N}\bm{C}^{-1}(\bm{y}_i - \bm{\mu})
+  \likelihood\left(\data|\vecmean, \veccov\right) = 
+  -\frac{Nd}{2}\log{\left(2\pi\right)} - \frac{N}{2}\log{|\veccov|} - \frac{1}{2}\Sigma_{i=1}^{N}\left(\data_i - \vecmean\right)^\intercal{}\veccov^{-1}\left(\bm{y}_i - \vecmean\right)
 \end{equation}
 
-\begin{equation}
-\frac{\delta^2\mathcal{L}}{\delta\bm{\mu}^2} = N\bm{C}^{-1}
-\end{equation}
+\noindent{}with respect to $\vecmean$,
 
-Computing the Fisher information of the covariance matrix $|\mathcal{F}\left(\bm{C}\right)|$ is harder. For a full covariance matrix with off-diagonal terms, the total
-number of free parameters required is $d(d+1)/2$.
+\begin{eqnarray}
+  \frac{\delta\likelihood}{\delta\vecmean} = \Sigma_{i=1}^{N}\veccov^{-1}(\data_i - \vecmean)
+  & \hspace{1em}\textrm{and}\hspace{1em}
+  & \frac{\delta^2\likelihood}{\delta\vecmean^2} = -N\veccov^{-1} \quad ,
+\end{eqnarray}
+
+\noindent{}and for a $d$-dimensional vector $\vecmean$ the Fisher information
+matrix is:
+
+\begin{eqnarray}
+  |\mathcal{F}(\vecmean)| = \left|-\frac{\delta^2\likelihood}{\delta\vecmean^2}\right| = N^d|\veccov|^{-1} \quad .
+\end{eqnarray}
+
+Computing the Fisher information of the covariance matrix 
+$|\mathcal{F}\left(\veccov\right)|$ is harder. 
+For a full covariance matrix with off-diagonal terms, the total number of free 
+parameters required is $d(d+1)/2$.
 \citep{Magnus_1988} derived an analytical expression:
+
 % Dwyer_1967, Bozdogan_1990? Drton_2009?
@@
 \begin{equation}
-|\mathcal{F}\left(\bm{C}\right) = N^\frac{d(d+1)}{2}2^{-d}|\bm{C}|^{-(d+1)}
+  |\mathcal{F}\left(\veccov\right)| = N^\frac{d(d+1)}{2}2^{-d}|\veccov|^{-(d+1)}
 \end{equation}
 
-Giving the approximate determinant of the Fisher information $|\mathcal{F}\left(\bm{\mu},\bm{C}\right)|$ as:
+\noindent{}Giving the approximate determinant of the expected Fisher information 
+$|\mathcal{F}\left(\vecmean,\veccov\right)|$ as:
 
 \begin{eqnarray}
-  |\mathcal{F}\left(\bm{\mu},\bm{C}\right)| & \approx & |\mathcal{F}\left(\bm{\mu}\right)|\cdot|\mathcal{F}\left(\bm{C}\right)| \\
-  & \approx & N^{d}|\bm{C}|^{-1}N^\frac{d(d+1)}{2}2^{-d}|\bm{C}|^{-(d+1)} \\
-  |\mathcal{F}\left(\bm{\mu},\bm{C}\right)| & \approx & N^\frac{d(d+3)}{2}2^{-d}|\bm{C}|^{-(d+2)}
+  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & |\mathcal{F}\left(\vecmean\right)|\cdot|\mathcal{F}\left(\veccov\right)| \nonumber \\
+  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & N^{d}|\veccov|^{-1}N^\frac{d(d+1)}{2}2^{-d}|\veccov|^{-(d+1)} \nonumber \\
+  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & N^\frac{d(d+3)}{2}2^{-d}|\veccov|^{-(d+2)} \quad .
 \end{eqnarray}
 
+\subsection{Encoding the message}
+
+Given our priors, likelihood function, and approximate expressions for the
+Fisher information, we are now in a position to calculate the total message
+length.
+There are a number of components that go into the message:
+
 A common method to estimate the parameters of a Gaussian distribution is by
-maximum likelihood, where the log-likelihood $\mathcal{L}$ is maximized. In
+maximum likelihood, where the log-likelihood $\likelihood$ is maximized. In
 order to do this, one computes the gradient of the log-likelihood function

### 2017-04-24 articles/chemical-tagging-gmm/ms.tex
some bad text

@@
 \begin{equation}
-I(\weights) = \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\weight_k - \log{\Gamma{\left(K\right)}} \quad ,
+I(\weights) = \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k - \log{\Gamma{\left(K\right)}} \quad ,
 \end{equation}
@@
 
-% The fisher information matrix
-
-\subsection{Prior probabilities on the parameters}
-\label{sec:priors}
-
-\subsection{Expected Fisher information}
-
-Computing the expected Fisher information requires the second order partial
-derivatives of the negative log-likelihood $-\likelihood(\data|\vecmean,\veccov)$.
-This is difficult to calculate for mixtures of Gaussians, and for this reason
-we approximate the determinant of the Fisher information 
-$|\mathcal{F}(\vecmean, \veccov)|$ as the product of 
-$|\mathcal{F}\left(\vecmean\right)|$ and $|\mathcal{F}\left(\veccov\right)|$ 
+Computing the expected Fisher information $\mathcal{F}$ requires the second 
+order partial derivatives of the negative log-likelihood 
+$-\likelihood(\data|\vecmean,\veccov)$.
+We approximate the determinant of the Fisher information 
+$|\mathcal{F}(\vecmean_k, \veccov_k)|$ as the product of 
+$|\mathcal{F}\left(\vecmean_k\right)|$ and $|\mathcal{F}\left(\veccov_k\right)|$ 
 \citep{Oliver_1996,Roberts_1998}. 
 We begin by taking second derivative of the log-likelihood function
-$\likelihood\left(\data|\vecmean,\veccov\right)$,
+$\likelihood\left(\data|\vecmean,\veccov\right)$ for a multivariate Gaussian
+distribution,
 
@@
 \begin{eqnarray}
-  |\mathcal{F}(\vecmean)| = \left|-\frac{\delta^2\likelihood}{\delta\vecmean^2}\right| = N^d|\veccov|^{-1} \quad .
+  |\mathcal{F}(\vecmean_k)| = \left|-\frac{\delta^2\likelihood}{\delta\vecmean^2}\right| = N^d|\veccov|^{-1} \quad .
 \end{eqnarray}
@@
 
-\subsection{Encoding the message}
+Thus for all $K$ mixtures, Equation \ref{eq:10} expands to:
 
-Given our priors, likelihood function, and approximate expressions for the
-Fisher information, we are now in a position to calculate the total message
-length.
-There are a number of components that go into the message:
+\begin{eqnarray}
+I(\vecmean, \veccov) &=& -\sum_{k=1}^K\frac{D + 1}{2}\log{|\veccov_k|} 
+                         + \frac{1}{2}\sum_{k=1}^K\left[\frac{D(D+3)}{2}\log{N} - D\log{2} - (D + 2)\log{|\veccov_k|}\right] \nonumber \\
+I(\vecmean, \veccov) &=& -\sum_{k=1}^K\frac{D + 1}{2}\log{|\veccov_k|}
+                         + \frac{KD(D + 3)}{4}
+                         + \frac{D\log{2}}{2}
+                         - \frac{1}{2}\sum_{k=1}^K (D+2)\log{|\veccov_k|} \nonumber\\
+I(\vecmean, \veccov) &=& -\sum_{k=1}^K(D + 1)\log{|\veccov_k|} + \frac{1}{2}\left(\frac{KD(D + 3)}{2} + D\log{2}\right)
+\end{eqnarray}
 
-A common method to estimate the parameters of a Gaussian distribution is by
-maximum likelihood, where the log-likelihood $\likelihood$ is maximized. In
-order to do this, one computes the gradient of the log-likelihood function
-with respect to the parameters and solves the resulting equations.
+\subsubsection{Encoding the data, given the model parameters}
+\label{sec:encoding-data}
 
-- Encoding the message.
+Each datum can only be stated with a precision which is given by the accuracy
+of the measurement.
+If we assume homoskedastic noise properties for the observed data, and the
+precision that each elemental abundance can be measured is $\epsilon$, then
+the \emph{probability} of a datum is given by 
+$\textrm{Pr}(\datum_n) = \epsilon^D\textrm{Pr}(\datum_i|\mathcal{M})$
+where $\mathcal{M}$ represents the model and $\textrm{Pr}(\datum_i|\mathcal{M})$
+is the \emph{probability density}.
+Note that $\epsilon$ is a constant, and is only used to maintain validity 
+between \emph{probability} and \emph{probability density}.
+For these experiments we adopt $\epsilon = 0.01$, but note that the value of
+$\epsilon$ has no effect on our inferences.
 
-- Process of expectation and maximization for each step.
+The length of the total encoding of a datum given the model is,
+\begin{equation}
+  I(\data_n) = -\log{\textrm{Pr}(\data_n)} = -D\log\epsilon - \log\sum_{k=1}^{K}w_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
+\end{equation}
+
+\noindent{}and for the entire data:
+
+\begin{equation}
+  I(\data|\vectheta) = -ND\log\epsilon - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k)
+\end{equation}
+
+\subsubsection{Encoding the lattice constant}
+
+\subsection{The Objective Function}
+\label{sec:objective-function}
+
+The message lengths of individual components allows us to fully specify an
+objective function to minimise the total message length, given any number of
+Gaussian mixtures $K$ and parameter estimates $\vectheta$:
+
+\begin{equation}
+% TODO:
+  I(\vectheta,\data) = I(K) + I(\weights) + \sum_{k=1}^{K}I(\vectheta_k) + I(\data|\vectheta) + \textrm{lattice} \quad .
+\end{equation}
+
+Substituting the expressions outlined in previous subsections, we arrive at:
+
+\begin{equation}
+  % FULL EQUATION.
+  I(\vectheta,\data) = long
+\end{equation}
+
+Note that our objective function includes the number of mixtures $K$, a
+discrete valued parameter. 
+Because $K$ is not continuous, this complicates our optimization procedure.
+
+Here we will outline our optimization procedure to minimise the message length
+when $K$ is known and fixed, before briefly describing our search strategy to
+move between trial values of $K$.
 
@@
 
+This research has made use of NASA's Astrophysics Data System.
 This research is supported by an Australian Research Council Discovery Project

### 2017-04-24 articles/chemical-tagging-gmm/experiments/experiment_v0.py
eek

@@ -1,0 +1,102 @@
+
+
+import numpy as np
+
+from astropy.table import Table
+from snob import mixture_ka as snob
+from sklearn import mixture
+
+
+catalog = Table.read("catalog.fits")
+
+
+realisations = {}
+
+# Number of Monte-Carlo realisations to do for each number of true clusters
+M = 10
+
+# Number of potential clusters.
+N = len(set(catalog["group"])) - 1 # subtract 1 to ignore field stars
+min_clusters = 1
+
+predictors = ("RA", "DEC", "VRAD")
+
+results = []
+
+np.random.seed(1234567890)
+
+for n in range(min_clusters, N):
+
+    for m in range(M):
+        if m > 0 and n == N: break # because at n = N, we get all clusters.
+
+        # Pick which clusters.
+        cluster_indices = np.random.choice(range(N), size=n, replace=False)
+
+        # Get all the stars from those clusters.
+        match = np.in1d(catalog["group"], cluster_indices)
+
+        # Construct the matrix of data.
+        y = np.array([catalog[p][match] for p in predictors]).T
+
+
+        # Determine number of Gaussians from MML
+        model = snob.GaussianMixture(y)
+        op_mu, op_cov, op_weight = model.fit(covariance_regularization=0)#1e-6)
+        mml_num = op_weight.size
+    
+        # Determine number of components by AIC/BIC.
+        aic = []
+        for k in range(1, 1 + 2*N):
+            model = mixture.GaussianMixture(n_components=k)
+            fitted_model = model.fit(y)
+
+            aic.append(fitted_model.aic(y))
+
+            if k > 2 \
+            and np.all(np.diff(aic[-3:]) > 0):
+                break
+
+        best_by_aic = 1 + np.argmin(aic)
+
+
+        # Determine number of components by BIC.
+        bic = []
+        for k in range(1, 1 + 2*N):
+            model = mixture.GaussianMixture(n_components=k)
+            fitted_model = model.fit(y)
+
+            bic.append(fitted_model.bic(y))
+
+            if k > 2 \
+            and np.all(np.diff(bic[-3:]) > 0):
+                break
+
+        best_by_bic = 1 + np.argmin(bic)
+
+        results.append((n, m, best_by_aic, best_by_bic, mml_num))
+
+        print(results[-1])
+
[…]

### 2017-04-24 articles/chemical-tagging-gmm/experiments/experiment_v1.py

@@ -1,0 +1,120 @@
+
+
+import numpy as np
+
+from astropy.table import Table
+from snob import mixture_ka as snob
+from sklearn import mixture
+
+
+catalog = Table.read("catalog.fits")
+
+
+realisations = {}
+
+# Number of Monte-Carlo realisations to do for each number of true clusters
+M = 10
+
+# Number of potential clusters.
+N = len(set(catalog["group"])) - 1 # subtract 1 to ignore field stars
+min_clusters = 1
+
+
+predictors \
+    = ("FE1", "MG1", "TI1", "SI1", "CA1",)# "AL1", "CR1", "MN1", "NI1", "CO1")#, "BA2")
+
+predictors = ("FE1", "MG1", )#"TI1", "SI1", "CA1", "AL1", "CR1", "BA2")
+
+
+finite = np.all(np.array([
+    np.isfinite(catalog[predictor]) for predictor in predictors]), axis=0)
+
+results = []
+
+np.random.seed(1234567890)
+
+for n in range(min_clusters, N):
+
+    for m in range(M):
+        if m > 0 and n == N: break # because at n = N, we get all clusters.
+
+        # Pick which clusters.
+        cluster_indices = np.random.choice(range(N), size=n, replace=False)
+
+        # Get all the stars from those clusters.
+        match = finite * np.in1d(catalog["group"], cluster_indices)
+
+        # Construct the matrix of data.
+        y = np.array([catalog[p][match] for p in predictors]).T
+        
+        raise a
+
+        # Determine number of Gaussians from MML
+        model = snob.GaussianMixture(y)
+        op_mu, op_cov, op_weight, meta = model.fit(covariance_regularization=0)
+        mml_num = op_weight.size
+
+        # Consider alternative MML, where we initialize it at the true solution
+
+        model2 = snob.GaussianMixture(y)
+        op_mu2, op_cov2, op_weight2, meta2 = model.fit(
+            covariance_regularization=0,
+            __initialize=)
+    
+        # Determine number of components by AIC/BIC.
+        aic = []
+        for k in range(1, 1 + 2*N):
+            model = mixture.GaussianMixture(n_components=k)
+            fitted_model = model.fit(y)
+
+            aic.append(fitted_model.aic(y))
+
+            if k > 2 \
+            and np.all(np.diff(aic[-3:]) > 0):
[…]

### 2017-04-24 snob/mixture_ka.py

@@ -117,8 +117,13 @@ class GaussianMixture(object):
             covariance_regularization=covariance_regularization,
             callback=callback)
 
-        # Initialize as a one-component mixture.
-        mu, cov, weight = _initialize(y, callback=callback)
+        # Do we have an initialization keyword?
+        if kwargs.get("__initialize", None) is not None:
+            mu, cov, weight = kwargs.pop("__initialize")
+
+        else:
+            # Initialize as a one-component mixture.
+            mu, cov, weight = _initialize(y, callback=callback)
 
         iterations = 1
         N_cp = _parameters_per_component(D, self.covariance_type)
@@ -183,7 +188,8 @@ class GaussianMixture(object):
                 break
 
         # TODO: a full_output response.
-        return (mu, cov, weight)
+        meta = dict(message_length=mindl)
+        return (mu, cov, weight, meta)
         
     fit = _fit_kasarapu_allison
 
@@ -597,6 +603,7 @@ def _compute_cholesky_decomposition(covariances, covariance_type):
                 cholesky_cov = scipy.linalg.cholesky(covariance, lower=True) 
             except scipy.linalg.LinAlgError:
                 raise ValueError(singular_matrix_error)
+
 
             cholesky_decomposition[m] = scipy.linalg.solve_triangular(
                 cholesky_cov, np.eye(D), lower=True).T

### 2017-04-24 articles/chemical-tagging-gmm/experiments/experiment_v1.py
bad stuff

@@ -31,6 +31,9 @@ finite = np.all(np.array([
 
 results = []
 
+running_delta_bic = 0
+running_delta_mml = 0
+
 np.random.seed(1234567890)
 
 for n in range(min_clusters, N):
@@ -47,20 +50,38 @@ for n in range(min_clusters, N):
         # Construct the matrix of data.
         y = np.array([catalog[p][match] for p in predictors]).T
         
-        raise a
+        K = cluster_indices.size
+        D = y.shape[1]
+        true_mu = np.empty((K, D))
+        true_cov = np.empty((K, D, D))
+        true_weights = np.empty(K)
+
+        for k, cluster_index in enumerate(cluster_indices):
+            _match = finite * (catalog["group"] == cluster_index) 
+            _data = np.array([catalog[p][_match] for p in predictors])
+            true_mu[k] = np.mean(_data, axis=1)
+            true_weights[k] = _match.sum()
+            true_cov[k] = np.cov(_data)
+
+        true_weights /= true_weights.sum()
 
         # Determine number of Gaussians from MML
         model = snob.GaussianMixture(y)
-        op_mu, op_cov, op_weight, meta = model.fit(covariance_regularization=0)
+        op_mu, op_cov, op_weight, meta = model.fit(covariance_regularization=1e-6)
         mml_num = op_weight.size
 
         # Consider alternative MML, where we initialize it at the true solution
 
         model2 = snob.GaussianMixture(y)
         op_mu2, op_cov2, op_weight2, meta2 = model.fit(
-            covariance_regularization=0,
-            __initialize=)
-    
+            covariance_regularization=1e-6,
+            __initialize=(true_mu, true_cov, true_weights))
+
+        if op_weight2.size != op_weight.size:
+            #assert meta2["message_length"] < meta["message_length"]
+            print("DID BETTER FROM TRUE")
+            mml_num = op_weight2.size
+
         # Determine number of components by AIC/BIC.
         aic = []
         for k in range(1, 1 + 2*N):
@@ -93,6 +114,9 @@ for n in range(min_clusters, N):
         results.append((n, m, y.shape[0], best_by_aic, best_by_bic, mml_num))
 
         print(results[-1])
+        running_delta_mml += abs(mml_num - n)
+        running_delta_bic += abs(best_by_bic - n)
+        print("MML/BIC", running_delta_mml, running_delta_bic)
 
 results = np.array(results)
 

### 2017-04-24 snob/mixture_ka.py

@@ -20,6 +20,8 @@ from collections import defaultdict
 logger = logging.getLogger(__name__)
 
 
+_DEFAULT_COVARIANCE_TYPE = "diag"
+
 class GaussianMixture(object):
 
     r"""
@@ -45,20 +47,20 @@ class GaussianMixture(object):
 
     parameter_names = ("mean", "cov", "weight")
 
-    def __init__(self, y, covariance_type="free", **kwargs):
+    def __init__(self, y, covariance_type=_DEFAULT_COVARIANCE_TYPE, **kwargs):
 
         y = np.atleast_2d(y)
         if 1 > y.size:
             raise ValueError("no data!")
         
-        available = ("free", "diag", "tied", "tied_diag")
+        available = ("full", "diag", "tied", "tied_diag")
         covariance_type = covariance_type.strip().lower()
         if covariance_type not in available:
             raise ValueError("covariance type '{}' is invalid. "\
                              "Must be one of: {}".format(
                                 covariance_type, ", ".join(available)))
 
-        if covariance_type not in ("free", "tied"):
+        if covariance_type not in ("full", "diag"):
             raise NotImplementedError("don't get your hopes up")
 
         self._y = y
@@ -266,8 +268,8 @@ def responsibility_matrix(y, mu, cov, weight, full_output=False):
     assert np.all(np.isfinite(log_likelihood))
     """
 
-    precision_cholesky = _compute_cholesky_decomposition(cov, "full")
-    weighted_log_prob = _estimate_log_gaussian_prob(y, mu, precision_cholesky, "full") \
+    precision_cholesky = _compute_cholesky_decomposition(cov, _DEFAULT_COVARIANCE_TYPE)
+    weighted_log_prob = _estimate_log_gaussian_prob(y, mu, precision_cholesky, _DEFAULT_COVARIANCE_TYPE) \
                       + np.log(weight)
     log_prob_norm = scipy.misc.logsumexp(weighted_log_prob, axis=1)
     with np.errstate(under="ignore"):
@@ -360,7 +362,7 @@ def _parameters_per_component(D, covariance_type):
         mean and covariance matrix of :math:`D` Gaussian distributions.
     """
 
-    if covariance_type == "free":
+    if covariance_type == "full":
         return int(D + D*(D + 1)/2.0)
     elif covariance_type == "diag":
         return 2 * D
[…]

### 2017-04-28 sandbox_mixture_slf.py
err

@@ -5,7 +5,7 @@ from snob import mixture_slf as slf
 #y = np.loadtxt("cluster_abundances.txt")
 
 n_samples, n_features, n_clusters, rank = 1000, 50, 4, 1
-sigma = 0.5
+sigma = 0.1
 true_specific_variances = sigma**2 * np.ones((1, n_features))
 
 

### 2017-04-28 snob/mixture_slf.py

@@ -17,8 +17,8 @@ logger = logging.getLogger(__name__)
 class SLFGaussianMixture(object):
 
     r"""
-    Model data from (potentially) many multivariate Gaussian distributions
-    with a single (multivariate) latent factor.
+    Model data from a fixed number of multivariate Gaussian distributions with
+    a single (multivariate) latent factor.
 
     The latent factor and mixture parameters are iteratively updated using
     expectation-maximization, with minimum message length describing the cost
@@ -221,7 +221,7 @@ class SLFGaussianMixture(object):
             # Check for convergence.
             change = np.abs((results[-1] - results[-2])/results[-2])
             
-            logger.debug("E-M step {}: {} {}".format(
+            print("E-M step {}: {} {}".format(
                 iteration, results[-1], change))
 
             if change <= self.threshold \
@@ -299,7 +299,6 @@ class SLFGaussianMixture(object):
         else:
             b = (self.factor_loads/np.sqrt(self.specific_variances)).flatten()
 
-
         for iteration in range(self.max_inner_iterations):
 
             # The iteration scheme is from Wallace & Freeman (1992)
@@ -313,8 +312,7 @@ class SLFGaussianMixture(object):
             b_sq = np.sum(b**2)
             if b_sq == 0:
                 raise NotImplementedError("a latent factor is not justified")
-                break
-
+                
             # Note: Step (c) of Wallace & Freeman (1992) suggest to compute
             #       Y as Y_{kj} = \frac{V_{kj}}{\sigma_{k}\sigma_{j}}, where
             #       V_{kj} is the kj entry of the correlation matrix, but that
@@ -352,6 +350,8 @@ class SLFGaussianMixture(object):
 
         return self.set_parameters(factor_loads, factor_scores, 
             specific_variances, means, weights)
+
+
 
 
 

### 2017-04-28 articles/gmm-search/ms.tex
some thoughts about a mixture search

%% This document is part of the snob project
%% Copyright 2017 the authors. All rights reserved.

\documentclass{aastex61}

\usepackage{bm}
%\usepackage{boldsymbol}
\usepackage{verbatim} % for \begin{comment}'ing out sections.

% For revision history
\IfFileExists{vc.tex}{\input{vc.tex}}{
    \newcommand{\githash}{UNKNOWN}
    \newcommand{\giturl}{UNKNOWN}}

% Define commands
\newcommand{\article}{\emph{Article}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}

% Surveys
\newcommand{\apogee}{\acronym{APOGEE}}
\newcommand{\ges}{\acronym{GES}}
\newcommand{\hermes}{\acronym{HERMES}}
\newcommand{\galah}{\acronym{GALAH}}
\newcommand{\fourmost}{\acronym{4MOST}}
\newcommand{\weave}{\acronym{WEAVE}}
\newcommand{\gaia}{\project{Gaia}}

% Common terms
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}

\def\teff{T_{\rm eff}}
\def\logg{\log{g}}

% Have some maaaath.
\def\infinity{\rotatebox{90}{8}}
\def\veccov{\vect{C}}
\def\vecmean{\vect{\mu}}
\def\vectheta{\vect{\theta}}
\def\weight{w}
\def\weights{\vect{\weight}}
\def\datum{y}
\def\data{\vect{\datum}}
\def\likelihood{\mathcal{L}}

% Affiliation(s)
\newcommand{\moca}{
	\affil{School of Physics and Astronomy, Monash University, 
		   Melbourne, Clayton VIC 3800, Australia}}
\newcommand{\claytonfit}{
	\affil{Faculty of Information Technology, Monash University,
	       Melbourne, Clayton VIC 3800, Australia}}
\newcommand{\caulfieldfit}{
	\affil{Faculty of Information Technology, Monash University,
		   Melbourne, Caulfield East VIC 3145, Australia}}

\received{}
\revised{}
\accepted{}

\submitjournal{TBD} % ApJ, MNRAS, PASA?

\shorttitle{A search strategy for Gaussian mixtures using MML}
\shortauthors{Casey et al.}

\begin{document}

\title{A search method for Gaussian mixtures using MML}

\author{Aleti, Casey, Dowe, Lattanzio}

\begin{abstract}

\end{abstract}

\keywords{}

\section{Introduction} 
\label{sec:introduction}

There are $N$ data points each with $D$ dimensions, which are to be modelled
by $K$ mixture of $D$-dimensional gaussian distributions, each with a relative
weighting $\weight_k$ such that $\sum_{k=1}^{K}\weight_k = 1$. 
The data have the same error value, $y_{err}$, in all $D$ dimensions, for all $N$ observations.

The full expression for the message length of a
given by,

\begin{eqnarray}
I_K & = & K\log{2} % I_k
    + \frac{(K - 1)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{w_k} - \log{|\Gamma(K)|} % I_w
    + \mathcal{L}(\data|\vectheta) - DN\log{y_{err}} \\ % Likelihood \\
  & + & \frac{1}{2}\sum_{k=1}^{K}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] % I_t
    - \frac{Q}{2}\log(2\pi) + \frac{\log{Q\pi}}{2} % lattice, minus the lattice of part 2
\end{eqnarray}

\noindent{}where $Q = \frac{1}{2}DK(D + 3) + K - 1$, the number of free parameters, and
$\mathcal{L}$ is the log-likelihood of a multivariate gaussian distribution.

Say we wanted to calculate whether another mixture was warranted. If another
mixture were preferred then we would want:

\begin{eqnarray}
  \Delta{}I_{K+1} - I_{K} < 0
\end{eqnarray}

The expression is given as:
\begin{eqnarray}
\Delta{I_{K+1} - I_K} & = & (K + 1)\log{2} - K\log{2} \\ % I_k^(new) - %I_k^(old)
  &&+ \frac{(K)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K+1}\log{w_k}^{(new)} - \log{|\Gamma(K+1)|} \\ % I_w^(new)
  &&- \frac{(K - 1)}{2}\log{N} + \frac{1}{2}\sum_{k=1}^{K}\log{w_k} + \log{|\Gamma(K)|}\\ % I_w^(old)
  &&+ \mathcal{L}^{(new)} - DN\log{y_{err}} \\ % Likelihood (new)\\
  &&- \mathcal{L}^{(old)} + DN\log{y_{err}} \\ % Likelihood (old) \\
  &&+ \frac{1}{2}\sum_{k=1}^{K+1(new)}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] \\ % I_t^(new)
  &&- \frac{1}{2}\sum_{k=1}^{K(old)}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] \\ % I_t^(old)
  &&- \frac{Q^{(new)}}{2}\log(2\pi) + \frac{\log{Q^{(new)}\pi}}{2} \\ % lattice, minus the lattice of part 2
  &&+ \frac{Q^{(old)}}{2}\log(2\pi) - \frac{\log{Q^{(old)}\pi}}{2} % lattice, minus the lattice of part 2
\end{eqnarray}

\noindent{}By making use of $\log{\Gamma(K)} - \log{\Gamma(K + 1)} = -\log{K}$ and re-arranging the expression:

\begin{eqnarray}
\Delta{}I_{K+1} - I_K &=& \log{2} % \Delta{}I_k
    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
&& + \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
&& + \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) - \left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
&& + \frac{\log(2\pi)}{2}(Q^{(old)} - Q^{(new)}) + \frac{\pi}{2}\left(\log{Q^{(new)}} - \log{Q^{(old)}}\right)
\label{eq:13}
\end{eqnarray}

Expanding the $Q$ terms:

\begin{eqnarray}
Q^{(old)} - Q^{(new)} &=& \frac{1}{2}DK(D + 3) + K - 1 - \frac{1}{2}D(K + 1)(D + 3) + (K + 1) - 1 \nonumber \\
Q^{(old)} - Q^{(new)} &=& -\frac{1}{2}D(D+3) + 2K  - 1
\label{eq:14}
\end{eqnarray}

\noindent{}And making use of the following logarithmic identities,
\begin{eqnarray}
  \log{Q^{(new)}} &=& \log{\left(\frac{1}{2}D(K+1)(D + 3) + K\right)} \nonumber \\
                  &=& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} + \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} \\
  \log{Q^{(old)}} &=& \log{\left(\frac{1}{2}DK(D + 3) + K - 1\right)} \nonumber \\
                  &=& \log{\left(\frac{1}{2}DK(D + 3)\right)} + \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)}
\end{eqnarray}

\noindent{}gives us,

\begin{eqnarray}
  \log{Q^{(new)}} - \log{Q^{(old)}} &=& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
                                    &&+ \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} - \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)} \quad .\\
\end{eqnarray}
The second row of terms can be ignored because they are very small (typically less than 1 bit). This is because as $K \rightarrow \infinity$, $2K/D(K+1)(D+3) \rightarrow 1$, thus $\log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} \rightarrow \log{2}$. Similarly as $D \rightarrow \infinity$, $2K/D(K+1)(D+3) \rightarrow 0$.

As $K \rightarrow \infinity$ then $2(K-1)/DK(D+3) \rightarrow 1$ and as $D \rightarrow \infinity$ then $2(K-1)/DK(D+3) \rightarrow 0$ and thus $\log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)} \approx 0$.

\noindent{}Ignoring these minor terms:

\begin{eqnarray}
  \log{Q^{(new)}} - \log{Q^{(old)}} &\approx& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
  \log{Q^{(new)}} - \log{Q^{(old)}} &\approx& \log{(K + 1)} - \log{K}
  \label{eq:19}
\end{eqnarray}

\noindent{}Substituting Eqs. \ref{eq:19} and \ref{eq:14} into \ref{eq:13} yields:

\begin{eqnarray}
\Delta{}I_{K+1} - I_K &\approx& \log{2} % \Delta{}I_k
    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
&& + \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
&& + \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) - \left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
&& + \frac{\log(2\pi)}{2}(-\frac{1}{2}D(D+3) + 2K  - 1) + \frac{\pi}{2}\left(\log{(K + 1)} - \log{K}\right)
\end{eqnarray}

\begin{thebibliography}{}

% Note: I include the DOI as a comment on the same line after every \bibitem
%       entry. I do this in case I ever need to extract full bibstems for the
%       bibliography, which can be done using this curl command and the DOI:

%       curl -LH "Accept: application/x-bibtex" http://dx.doi.org/10.5555/12345678

%       If no DOI is available then I list the ADS identifier.
\end{thebibliography}

\end{document}

### 2017-04-28 .ci/travis.sh

@@ -5,6 +5,8 @@ cd articles/chemical-tagging-gmm
 make
 cd articles/nips-2017
 make
+cd articles/gmm-search
+make
 
 # Push to GitHub
 if [ -n "$GITHUB_API_KEY" ]; then

### 2017-04-28 sandbox_mixture_search.py

@@ -1,0 +1,97 @@
+
+import numpy as np
+from snob import mixture_search as mixture
+
+
+
+# Generate data from the example in Section 13.4 of P & A (2015)
+np.random.seed(888)
+N = 1000
+weight = np.array([0.3, 0.3, 0.3, 0.1])
+mu = np.array([
+    [-4, -4],
+    [-4, -4],
+    [2, 2],
+    [-1, -6]
+])
+
+cov = np.array([
+    [
+        [1, 0.5],
+        [0.5, 1]
+    ],
+    [
+        [6, -2],
+        [-2, 6]
+    ],
+    [
+        [2, -1],
+        [-1, 2]
+    ],
+    [
+        [0.125, 0],
+        [0, 0.125]
+    ]
+])
+
+
+y = np.vstack([np.random.multivariate_normal(
+        mu[i], cov[i], size=int(N * weight[i])) \
+    for i in range(len(weight))])
+
+"""
+Approximating \sum\log{w_k}...
+
+bar = []
+for i in range(1, 101):
+
+    foo = np.random.uniform(size=(1000, i))
+    foo = foo.T/np.sum(foo, axis=1)
+    print(i)
+
+    mean = np.mean(np.log(foo).sum(axis=0))
+    std = np.std(np.log(foo).sum(axis=0))
+
+    bar.append([i, mean, std])
+
+    
+
+bar = np.array(bar)
+
+fig, ax = plt.subplots()
+ax.scatter(bar.T[0], bar.T[2])
+#ax.scatter(bar.T[0], bar.T[1])
+#ax.errorbar(bar.T[0], bar.T[1], bar.T[2], fmt=None)
+
+raise a
+"""
+
+"""
+# Approximating \log(\sum{r}) (the log of the effective memberships...)
+bar = []
+for i in range(1, 101):
+
+    foo = np.random.uniform(size=(1000, i))
+    foo = foo.T/np.sum(foo, axis=1)
+    foo *= 900 # The sample size, say.
+
+    mean = np.mean(np.log(foo).sum(axis=0))
+    std = np.std(np.log(foo).sum(axis=0))
+
+    bar.append([i, mean, std])
+
+
+bar = np.array(bar)
+
+fig, axes = plt.subplots(2)
+axes[0].scatter(bar.T[0], bar.T[1])
+axes[1].scatter(bar.T[0], bar.T[2])
+
+raise a
+"""
+
+model = mixture.GaussianMixture()
+mu, cov, weight, meta = model.fit(y)
+
+
+raise a

### 2017-04-28 snob/mixture_search.py

@@ -1,0 +1,1162 @@
+
+"""
+An estimator for modelling data from a mixture of Gaussians, 
+using an objective function based on minimum message length.
+"""
+
+__all__ = [
+    "GaussianMixture", 
+    "kullback_leibler_for_multivariate_normals",
+    "responsibility_matrix",
+    "split_component", "merge_component", "delete_component", 
+] 
+import logging
+import numpy as np
+import scipy
+from collections import defaultdict
+
+logger = logging.getLogger(__name__)
+
+
+class GaussianMixture(object):
+
+    r"""
+    Model data from (potentially) many multivariate Gaussian distributions, 
+    using minimum message length (MML) as the objective function.
+
+    :param covariance_type: [optional]
+        The structure of the covariance matrix for individual components.
+        The available options are: `full` for a free covariance matrix, or
+        `diag` for a diagonal covariance matrix (default: ``diag``).
+
+    :param covariance_regularization: [optional]
+        Regularization strength to add to the diagonal of covariance matrices
+        (default: ``0``).
+
+    :param threshold: [optional]
+        The relative improvement in message length required before stopping an
+        expectation-maximization step (default: ``1e-5``).
+
+    :param max_em_iterations: [optional]
+        The maximum number of iterations to run per expectation-maximization
+        loop (default: ``10000``).
+    """
+
+    parameter_names = ("mean", "covariance", "weight")
+
+    def __init__(self, covariance_type="full", covariance_regularization=0, 
+        threshold=1e-5, max_em_iterations=10000, **kwargs):
+
+        available = ("full", "diag", )
+        covariance_type = covariance_type.strip().lower()
+        if covariance_type not in available:
+            raise ValueError("covariance type '{}' is invalid. "\
+                             "Must be one of: {}".format(
+                                covariance_type, ", ".join(available)))
+
+        if 0 > covariance_regularization:
+            raise ValueError(
+                "covariance_regularization must be a non-negative float")
+
+        if 0 >= threshold:
+            raise ValueError("threshold must be a positive value")
+
+        if 1 > max_em_iterations:
+            raise ValueError("max_em_iterations must be a positive integer")
+
+        self._threshold = threshold
[…]

### 2017-05-01 articles/chemical-tagging-gmm/ms.tex
Expand math for model, some minor text changes

--- 2017-04-24 articles/chemical-tagging-gmm/ms.tex
@@
 stars could be used to reconstruct the evolution of the Milky Way, providing
-insight on supernovae yields, galactic dynamics (e.g., constraints on radial 
-mixing), the initial mass function, as well as the formation and destruction
-of star clusters. Some progress has been made towards chemical tagging, but 
+insight on supernovae yields, galactic dynamics, the initial mass function, 
+as well as the formation and destruction of star clusters.
+
+Some progress has been made towards chemical tagging, but 
 here we argue that even the \emph{extent} of the problem has not been realised.
@@
 only the most massive or isolated star clusters are able to survive the star
-formation process \citep{someone2}, leaving most clusters to become 
+formation process \citep[e.g.,][]{Lada_2010}, leaving most clusters to become 
 dynamically unbound after about 1~Gyr.
@@
 
-%The Milky Way is our best laboratory for understanding galaxy formation and
-%evolution. 
-Making strong inferences about the Milky Way's formation and evolution from
-chemical tagging will require a transformative number of stars with precise
-abundances.
-Ground- and space-based surveys are poised to deliver those data this decade. 
+Despite the Milky Way being our best laboratory for understanding galaxy
+formation, a transformative number of stars with precise abundances will be
+required to chemically tag the Galaxy.
+Ground- and space-based surveys are poised to deliver those data in the coming
+decade. 
 The Apache Point Observatory Galactic Evolution Experiment 
@@
 a formidable parameter estimation and model selection problem.
-Before reviewing the literature on chemical tagging, here we explicitly
-formulate the scope of the analysis problem.
+Given the difficulty of the problem, most chemical tagging studies have 
+focussed on addressing one \emph{aspect} of chemical tagging.
+For example, even the simplest applications of chemical tagging requires the
+identification of groups in the data, with stars being assigned to certain
+groups (a clustering problem), and to somehow chose the number of groups that
+are best represented by the data (a model selection problem).
+Previous works include definitions of L1-like metrics \citep{Mitschang_2014},
+where open clusters are used to define confidence intervals of similarity
+\citep{Mitschang_2013}, and whether similar stars constitute a group or not.
+Other studies have directly applied standard clustering algorithms,
+\citep[e.g., \acronym{DBSCAN}, $k$-means;][]{Blanco_Cuaresma_2015,Hogg_2016} 
+and fixed the number of clusters.
+Inferring the number of true clusters -- one of the principal goals of 
+chemical tagging -- was not attempted in these simplified experiments.
 
-\subsection{What do we want to infer from chemical abundances?}
+Other works have focussed [TODO EXPAND]:
+\begin{itemize}
+  \item Dimensionality of chemical abundance space (Ting et al).
+  \item Chemical homogenity of open clusters (Bovy, Da Silva)
+  \item Dopplegangers (Ness)
+  \item A single common latent factor, integrated over cosmic time (Bland-Hawthorn)
+\end{itemize}
+
+The aforementioned chemical tagging experiments have sought to address only 
+individual challenges associated with chemical tagging.
+These challenges are all related, and therefore addressing them separately
+will give results that ado not make use of all
+the information available. 
+Here, as we introduce the problem of chemical tagging, we list the inferences
+we wish to make from a large number of chemical abundances.
+
+\subsection{What's the goal of chemical tagging? What do we want to infer?}
 \label{sec:the-problem}
@@
         lifetime, at least in part due to atomic diffusion, veiling, and 
-        thermohaling mixing.
+        thermohaling mixing. % TODO: Add Korn, others?
         Understanding and modelling these effects will be critical for 
@@
 } objective functions.
-
-\subsection{Chemical tagging progress to date}
-\label{sec:literature-review}
-
-There has been substantial progress on chemical tagging in the last decade.
-Given the difficulty of the problem, most studies have opted to address one
-(or a few) of the challenges listed in Section~\ref{sec:the-problem}. 
-The most common issue addressed is group-finding (items 1-4). 
-Some works have sought to construct a similarity metric \citep{Mitschang}
-based on chemical abundances of undisrupted clusters. 
-Others have applied group-finding algorithms \citep{Hogg,Gregor}.
-
-% TODO
-- Dimensionality of chemical abundance space.
-
-- Chemical homogeneity.
-
-- Dopplegangers.
-
-The aforementioned chemical tagging experiments have sought to address only 
-individual challenges associated with chemical tagging.
-These challenges are all related, and therefore addressing them separately
-will yield results that are not self-consistent, and do not make use of all
-the information available.
 
@@
 
-\noindent{}The probability density function $p$ for a multivariate normal 
+\noindent{}The probability density function $f$ for a multivariate normal 
 distribution with $D$ dimensions is given by
@@
 \begin{equation}
-  p(\data|\vecmean,\veccov) 
+  f(\data|\vecmean,\veccov) 
       = \frac{1}{\sqrt{(2\pi)^d|\veccov|}}
@@
 \begin{equation}
-  p(\data|\vectheta) = \sum_{k=1}^{K} \weight_k \, p(\data|\vectheta_k)
+  f(\data|\vectheta) = \sum_{k=1}^{K} \weight_k \, f(\data|\vectheta_k)
 \end{equation}
@@
   belongs.
-  These two models are mathematically identical \citep{Foreman_Mackey_2014},
+  These two models are mathematically identical \citep[e.g., see discussion in][]{Foreman_Mackey_2014},
   but the integer representation is highly dimensional and difficult to
@@
 \noindent{}which states that the joint probability of the data $D$ and 
-hypothesis $H$, $P(H, D)$, is given by
+hypothesis $H$, $P(H,D)$, is given by
 
 \begin{equation}
-  P(H, D) = P(H) \cdot P(D|H) = P(D) \cdot P(H|D)
+  P(H,D) = P(H) \cdot P(D|H) = P(D) \cdot P(H|D)
 \end{equation}
@@
 \begin{equation}
-  I(H, D) = I(H) + I(D|H) = I(D) + I(H|D) \quad .
+  I(H,D) = I(H) + I(D|H) = I(D) + I(H|D) \quad .
   \label{eq:mml-wallace}
@@
 forces us to make approximations when calculating the message length.
-\citet{Wallace_Freeman_1987} introduce a generalised scheme to estimate a
-vector of parameters $\vectheta$ of any distribution given some data, where
-the estimates $\vectheta$ that minimise the message length expression,
-
+Using a Taylor expansion, \citet{Wallace_1987} introduce a generalised 
+scheme to estimate a parameter vector $\vectheta$ (of any distribution), that
+minimises the message length expression:
 \begin{equation}
+\label{eq:message-length}
   I(\vectheta,\data) 
-    = \frac{p}{2}\log{q_p} - \log{\left(\frac{h(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
-      - \likelihood\left(\data|\vectheta\right) + \frac{p}{2}
-      \label{eq:message-length}
+    = \frac{Q}{2}\log\kappa{\left(Q\right)} 
+    - \log{\left(\frac{p(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
+    - \likelihood\left(\data|\vectheta\right) 
+    + \frac{Q}{2} \quad .
 \end{equation}
 
-\noindent{}is the MML estimate.
-Here $h(\vectheta)$ is the prior on the hypothesis, $p$ is the number of free
-parameters in the model, $q_p$ is the lattice quantization constant
-\citep[e.g.,][]{Conway_1984}, and $|\mathcal{F}\left(\vectheta\right)|$ is the
-determinant of the \emph{expected} Fisher information matrix (the second-order
-partial derivatives of the negative log-likelihood function 
+Here $p(\vectheta)$ is the prior on the hypothesis, $q$ is the number of free
+parameters in the model, $\kappa\left(Q\right)$ is a function to approximate the lattice
+quantisation constant for $Q$ free parameters \citep[e.g.,][]{Conway_1984}, 
+and $|\mathcal{F}\left(\vectheta\right)|$ is the determinant of the 
+\emph{expected} Fisher information matrix (the second-order partial 
+derivatives of the negative log-likelihood function 
 $-\likelihood(\data|\vectheta)$).
@@
 
-The MML principle requires that the prior beliefs of the model to be fully 
-specified as part of the message. 
-We must explicitly specify our prior beliefs on the model parameters described
-in Section \ref{sec:model} in order to minimize the full message length.
-Indeed, the message must fully specify all components necessary for the
-receiver to be able to reconstruct the message.
-For the model specified in Section \ref{sec:model}, this requirement forces us
-to specify a number of components:
+The MML principle requires that the message fully specifies all components
+necessary for the receiver to be able to reconstruct the message, as well as
+our prior beliefs of the model.
+The message must include:
+
+%This requirement provides a familiar link to Bayesian statistics, and allows
+%one to think of MML as a form of a Bayesian optimization problem that is 
+%deeply grounded in information theory, with successful contributions in
+%economics, medicine, statistics, etc \citep{citations_needed}.
 
@@
   \item The number of Gaussian mixtures, $K$.
-  \item Encoding the relative weights $\weights$ of the Gaussian mixtures.
-  \item Encoding the component parameters $\vecmean$, $\veccov$ for all of
-        the Gaussian mixtures.
+  \item Encoding the relative weights $\weights$ of the $K$ Gaussian mixtures.
+  \item Encoding the component parameters $\vecmean$, $\veccov$ for all $K$
+        Gaussian mixtures.
   \item Encoding the data, given the model parameters.
-  \item The lattice quantization constant $q_p$ for the number of model
-        parameters $p$.
+  \item The lattice quantization constant $\kappa(q)$ for the number of model
+        parameters $q$.
 \end{enumerate}
@@
 Some of these components require non-trivial derivations or approximations in
-order to calculate the length of the message required to encode the component.
+order to calculate the length of the optimal loseless encoding. 
 For this reason, we will consider the message length of each component in turn
@@
 Specifically, we assume that $p(K) \propto 2^{-K}$, which implies that the
-message length required to encode $K$ is given by:
+optimal message length required to encode $K$ is given by:
 
@@
 $\sum_{k=1}^{K}w_k = 1$. 
-% This implies the weights are parameters of a multinomial distribution.
-\citet{Boulton_1969} give the following expression for the message length 
-required to encode the weights,
+This implies that the weights can be treated as parameters of a multinomial
+distribution, and the length of their optimally encoded message is given by
+\citet{Boulton_1969},
 
 \begin{equation}
-I(\weights) = \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\weight_k - \log{\Gamma{\left(K\right)}} \quad ,
+  I(\weights) 
+    = \frac{K - 1}{2}\log{N} 
+    - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k 
+    - \log{\Gamma{\left(K\right)}} \quad ,
 \end{equation}
@@
 
-%This requirement provides a familiar link to Bayesian statistics, and allows
-%one to think of MML as a form of a Bayesian optimization problem that is 
-%deeply grounded in information theory, with successful contributions in
-%economics, medicine, statistics, etc \citep{citations_needed}.
-
 \subsubsection{Encoding the mixture parameters $\vecmean$ and $\veccov$}
@@
 
-Recall that what we need to encode is both our priors on the model parameters
-and the determinant of the expected fisher information matrix.
+In order to properly encode the mixture parameters $\vecmean$ and $\veccov$
+for all $K$ mixtures, we must encode both our prior belief on those parameters,
+and the determinant of the expected Fisher information matrix.
 For the $k$-th mixture this becomes,
@@
   I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
+  \label{eq:10}
 \end{equation}
 
-This requires us to explicitly specify our priors on the parameters $\vecmean$
-and $\veccov$.
-We adopt an improper uniform prior of 
-$\mathcal{U}(\vecmean) = [-\infinity, \infinity]$ on all multivariate mean 
-abundances $\vecmean$ (in all $D$ dimensions).
-This prior has no impact on our inferences and becomes proper when it enters 
-the (accepted) posterior distribution. 
-We similarily could adopt a proper and large uniform prior on all mean 
-abundances of $\mathcal{U}(\vecmean) = [-12, 12]$ (24 orders of magnitude!) 
-and the result would be the same.
+We introduce our priors on $\vecmean$ and $\veccov$ before approximating the
+expected Fisher information matrix.
+We adopt a uniform prior of $\mathcal{U}(\vecmean) = [-\infinity, \infinity]$ 
+on all multivariate mean abundances $\vecmean$ (in all $D$ dimensions).
+We note that while this prior is improper, it becomes proper when it enters
+the (accepted) posterior distribution, and it has no impact on our inferences.
+Indeed, we similarly could adopt a (proper and) large uniform prior on all 
+mean abundances of $\mathcal{U}(\vecmean) = [-12, 12]$ (24 orders of magnitude!)
+and our inference would be the same.
 
@@
 mixture is given by the limiting form of the normal inverted-Wishart
-density \citep[e.g., Section 5.2.3 of ][]{Schafer_1990}:
+density \citep[e.g., Section 5.2.3 of ][]{Schafer_1997}:
 
@@
     p(\vecmean_k,\veccov_k) \propto |\veccov_k|^{\frac{D + 1}{2}} \quad .
+    \label{eq:prior-mean-cov}
 \end{equation}
 
-% The fisher information matrix
-
-\subsection{Prior probabilities on the parameters}
-\label{sec:priors}
-
-\subsection{Expected Fisher information}
-
-Computing the expected Fisher information requires the second order partial
-derivatives of the negative log-likelihood $-\likelihood(\data|\vecmean,\veccov)$.
-This is difficult to calculate for mixtures of Gaussians, and for this reason
-we approximate the determinant of the Fisher information 
-$|\mathcal{F}(\vecmean, \veccov)|$ as the product of 
-$|\mathcal{F}\left(\vecmean\right)|$ and $|\mathcal{F}\left(\veccov\right)|$ 
+Computing the expected Fisher information $\mathcal{F}(\vecmean_k,\veccov_k)$ 
+requires the second order partial derivatives of the negative log-likelihood 
+$-\likelihood(\data|\vecmean,\veccov)$.
+We approximate the determinant of the Fisher information 
+$|\mathcal{F}(\vecmean_k, \veccov_k)|$ as the product of 
+$|\mathcal{F}\left(\vecmean_k\right)|$ and $|\mathcal{F}\left(\veccov_k\right)|$ 
 \citep{Oliver_1996,Roberts_1998}. 
@@
 
-\noindent{}and for a $d$-dimensional vector $\vecmean$ the Fisher information
+\noindent{}and for a $D$-dimensional vector $\vecmean$ the Fisher information
 matrix is:
@@
 \begin{eqnarray}
-  |\mathcal{F}(\vecmean)| = \left|-\frac{\delta^2\likelihood}{\delta\vecmean^2}\right| = N^d|\veccov|^{-1} \quad .
+  |\mathcal{F}(\vecmean_k)| = \left|-\frac{\delta^2\likelihood}{\delta\vecmean^2}\right| = N^D|\veccov|^{-1} \quad .
+  \label{eq:approx-fisher-mean}
 \end{eqnarray}
@@
 $|\mathcal{F}\left(\veccov\right)|$ is harder.
-For a full covariance matrix with off-diagonal terms, the total number of free 
-parameters required is $d(d+1)/2$.
-\citep{Magnus_1988} derived an analytical expression:
+Using theory of matrix derivatives based on matrix vectorization \citep{Dwyer_1967},
+\citet{Magnus_1988} derived the following analytical expression for the 
+determinant of the Fisher information for a covariance matrix
+$|\mathcal{F}\left(\veccov_k\right)|$,
 
-% Dwyer_1967, Bozdogan_1990? Drton_2009?
 % TODO: This derivation is a nightmare to find
@@
 \begin{equation}
-  |\mathcal{F}\left(\veccov\right)| = N^\frac{d(d+1)}{2}2^{-d}|\veccov|^{-(d+1)}
+  |\mathcal{F}\left(\veccov_k\right)| = N^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)}
+  \label{eq:approx-fisher-cov}
 \end{equation}
 
-\noindent{}Giving the approximate determinant of the expected Fisher information 
-$|\mathcal{F}\left(\vecmean,\veccov\right)|$ as:
+\noindent{}where $D(D+1)/2$ arises from the number of free parameters in a
+covariance matrix with off-diagonal terms.
+Equations \ref{eq:approx-fisher-mean} and \ref{eq:approx-fisher-cov} allow us
+to approximate the determinant of the expected Fisher information matrix
+$|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|$ as:
 
 \begin{eqnarray}
-  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & |\mathcal{F}\left(\vecmean\right)|\cdot|\mathcal{F}\left(\veccov\right)| \nonumber \\
-  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & N^{d}|\veccov|^{-1}N^\frac{d(d+1)}{2}2^{-d}|\veccov|^{-(d+1)} \nonumber \\
-  |\mathcal{F}\left(\vecmean,\veccov\right)| & \approx & N^\frac{d(d+3)}{2}2^{-d}|\veccov|^{-(d+2)} \quad .
+  |\mathcal{F}\left(\vecmean_k,\veccov_k\right)| & \approx & |\mathcal{F}\left(\vecmean_k\right)|\cdot|\mathcal{F}\left(\veccov_k\right)| \nonumber \\
+  |\mathcal{F}\left(\vecmean_k,\veccov_k\right)| & \approx & N^{D}|\veccov_k|^{-1}N^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
+  |\mathcal{F}\left(\vecmean_k,\veccov_k\right)| & \approx & N^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
+  \label{eq:fisher-mean-cov}
 \end{eqnarray}
 
-\subsection{Encoding the message}
+Thus for all $K$ mixtures, substituting Equations \ref{eq:prior-mean-cov} and 
+\ref{eq:fisher-mean-cov} into Equation \ref{eq:10} gives:
 
-Given our priors, likelihood function, and approximate expressions for the
-Fisher information, we are now in a position to calculate the total message
-length.
-There are a number of components that go into the message:
+\begin{eqnarray}
+I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|} \nonumber  \\
+I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{\left(|\veccov_k|^{\frac{D + 1}{2}}\right)}
+ + \frac{1}{2}\sum_{k=1}^{K}\log{\left(N^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right)} \nonumber \\
+I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\frac{D + 1}{2}\log|\veccov_k|
+ + \frac{1}{2}\sum_{k=1}^{K}\left[\frac{D(D+3)}{2}\log{N} -D\log{2} -(D+2)\log|\veccov_k|\right] \nonumber \\
+I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\frac{D + 1}{2}\log|\veccov_k|
+ + \frac{1}{2}K\left[\frac{D\left(D+3\right)}{2}\log{N} - D\log{2}\right]
+ -\sum_{k=1}^{K}\frac{D+2}{2}\log|\veccov_k| \nonumber \\
+I(\vecmean,\veccov) &=& \frac{1}{2}\sum_{k=1}^{K}\log|\veccov_k|
+ + \frac{1}{2}K\left[\frac{D\left(D+3\right)}{2}\log{N} - D\log{2}\right] \quad .
+\end{eqnarray}
 
-A common method to estimate the parameters of a Gaussian distribution is by
-maximum likelihood, where the log-likelihood $\likelihood$ is maximized. In
-order to do this, one computes the gradient of the log-likelihood function
-with respect to the parameters and solves the resulting equations.
+\subsubsection{Encoding the data, given the model parameters}
+\label{sec:encoding-data}
 
-- Encoding the message.
+Each datum can only be stated with a precision which is given by the accuracy
+of the measurement.
+If we assume homoskedastic noise properties for the observed data, and the
+precision that each elemental abundance can be measured is $\mathcal{E}$, then
+the \emph{probability} of a datum is given by 
+$\textrm{Pr}(\datum_n) = \mathcal{E}^D\textrm{Pr}(\datum_i|\mathcal{M})$
+where $\mathcal{M}$ represents the model and $\textrm{Pr}(\datum_i|\mathcal{M})$
+is the \emph{probability density}.
+Note that $\mathcal{E}$ is a constant, and is only used to maintain validity 
+between \emph{probability} and \emph{probability density}.
+For these experiments we adopt $\mathcal{E} = 0.01$, but note that the value of
+$\mathcal{E}$ has no effect on our inferences.
 
-- Process of expectation and maximization for each step.
+The length of the total encoding of a datum given the model is,
+\begin{equation}
+  I(\data_n) = -\log{\textrm{Pr}(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
+\end{equation}
+
+\noindent{}and for the entire data:
+
+\begin{eqnarray}
+  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .\\
+%  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{k=1}^{K}\log{\weight_k} - \sum_{n=1}^{N}\sum_{k=1}^{K}f_k\left(\data_n|\vecmean_k,\veccov_k\right) \quad .
+\end{eqnarray}
+
+\subsubsection{Encoding the lattice constant}
+\label{sec:encoding-lattice-constant}
+
+The lattice quantisation constant arises from the approximation to the (so-called)
+strict MML, where parameters are quantised into intervals in high dimensional
+space.
+First we will calculate the total number of free parameters in our model.
+Recall that we assume our covariance matrices have non-zero off-diagonal terms,
+requiring $\frac{KD(D+1)}{2}$ parameters for all $K$ covariance matrices
+(where we implicitly assume $D > 1$).
+The $K$ $D$-dimensional vector means require a total of $KD$ parameters.
+Under the assumption $K > 1$, the $K$ relative weights $\weights$ only require
+$K - 1$ parameters because $\sum_{k=1}^{K}\weight_k = 1$.
+Therefore the total number of free parameters $Q$ is given by,
+
+\begin{eqnarray}
+  Q &=& \frac{KD(D+1)}{2} + KD + K - 1 \nonumber \\
+  Q &=& K\left[\frac{D\left(D+3\right)}{2} + 1\right] - 1 \quad .
+  \label{eq:q}
+\end{eqnarray}
+
+We use an approximation for the logarithm of the lattice constant
+\citep[see Sections 5.2.12 and 3.3.4 of ][]{Wallace_2005} where,
+
+\begin{eqnarray}
+    I(y) \approx - \log\frac{p(\theta')}{\sqrt{F(\theta')}} 
+                   - \log{f(y|\theta')}
+                   - (Q/2)\log{2\pi}
+                   + \frac{1}{2}\log{(\pi{}Q)}
+                   - 1
+\end{eqnarray}
+
+\noindent{}such that,
+\begin{eqnarray}
+  \log\kappa(Q) &=& \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1
+\end{eqnarray}
+
+\noindent{}and the expected error on $\log\kappa(Q)$ is less than 0.1~nit.
+The $Q$ (lattice) terms in Equation \ref{eq:message-length} becomes:
+
+\begin{eqnarray}
+\frac{Q}{2}\left(\log\kappa\left(Q\right) + 1\right) &=& \frac{Q}{2}\left(\frac{\log{Q\pi}}{Q} - \log{2\pi} - 1 + 1\right) \nonumber \\
+\frac{Q}{2}\left(\log\kappa\left(Q\right) + 1\right) &=& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) \quad .
+\end{eqnarray}
+
+\subsection{The Objective Function}
+\label{sec:objective-function}
+
+The message lengths of individual components allows us to fully specify an
+objective function to minimise the total message length, given any number of
+Gaussian mixtures $K$ and parameter estimates $\vectheta$:
+
+\begin{equation}
+  I(\vectheta,\data) = I(K) + I(\weights) + \sum_{k=1}^{K}I(\vectheta_k) + I(\data|\vectheta) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right)\quad .
+\end{equation}
+
+\noindent{}Substituting the expressions outlined in previous subsections 
+(except for $Q$, as per Equation \ref{eq:q}, which we leave unexpanded
+for brevity), we arrive at our complete objective function:
+
+\begin{eqnarray}
+  I(\vectheta,\data) &=&
+      K\log{2} % I(K)
+    + \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k - \log{\Gamma(K)} % I(\weights) 
+    + \frac{1}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\ % 
+   &+&\frac{K}{2}\left[\frac{D(D+3)}{2}\log{N} - D\log{2}\right] % I(\vectheta_k) = I(\vecmean_k,\veccov_k)
+    - ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) % I(\data|\vectheta)
+    + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) \quad . % lattice
+    \label{eq:objective-function}
+\end{eqnarray}
+
+Our objective function includes the number of mixtures, $K$, a discrete valued
+parameter.
+We seek the number of mixtures $K$ and the parameter estimates $\vectheta$
+that minimise our objective function (Eq. \ref{eq:objective-function}).
+
+Here we will outline our optimization procedure to minimise the message length
+when $K$ is fixed, before describing our search strategy to move between trial 
+values of $K$.
+
+[TODO]
+- Expectation maximization, using MML (unbiased) updates
+
+\subsection{The Search Strategy}
+\label{sec:search-strategy}
 
@@
 
+\bibitem[Blanco-Cuaresma et al.(2015)]{Blanco_Cuaresma_2015} Blanco-Cuaresma, S., Soubiran, C., Heiter, U., et al.\ 2015, \aap, 577, A47 % 10.1051/0004-6361/201425232
+
+\bibitem[Boulton \& Wallace(1969)]{Boulton_1969} Boulton, D.~M. \& Wallace, C.~S.\ 1969, Journal of Theoretical Biology, 23, 269-278
+
+\bibitem[Conway \& Sloane(1984)]{Conway_1984} Conway, J.~H. \& Sloane, N.~J.~A.\ 1984, Journal on Algebraic and Discrete Methods, 5, 3, 294-305 % 10.1137/0605031
+
 \bibitem[Dalton et al.(2012)]{weave} Dalton, G., Trager, S.~C., Abrams, D.~C., et al.\ 2012, \procspie, 8446, 84460P % 10.1117/12.925950
@@
 
+%Some applications of matrix derivatives in multivariate analysis.
+
+\bibitem[Dwyer (1967)]{Dwyer_1967} Dwyer, P.~S.\ 1967, Journal of the American Statistical Association, 62, 318, 607-625
+
+\bibitem[Foreman-Mackey(2014)]{Foreman_Mackey_2014} Foreman-Mackey, D.\ 2014, Blog Post: Mixture Models., Zenodo, https://doi.org/10.5281/zenodo.15856 % 10.5281/zenodo.15856
+
 \bibitem[Freeman \& Bland-Hawthorn(2002)]{Freeman_2002} Freeman, K., \& Bland-Hawthorn, J.\ 2002, \araa, 40, 487 % 10.1146/annurev.astro.40.060401.093840
@@
 
+\bibitem[Hogg et al.(2016)]{Hogg_2016} Hogg, D.~W., Casey, A.~R., Ness, M., et al.\ 2016, \apj, 833, 262 % 10.3847/1538-4357/833/2/262
+
+\bibitem[Lada(2010)]{Lada_2010} Lada, C.~J.\ 2010, Philosophical Transactions of the Royal Society of London Series A, 368, 713 % 10.1098/rsta.2009.0264
+
 \bibitem[Majewski et al.(2015)]{apogee} Majewski, S.~R., Schiavon, R.~P., Frinchaboy, P.~M., et al.\ 2015, arXiv:1509.05420 % 2015arXiv150905420M
+
+\bibitem[Magnus \& Neudecker(1988)]{Magnus_1988} Magnus, J.~R. \& Neudecker, H.\ 1998, Matrix differential calculus with applications in statistics and econometrics. Wiley, New York.
+
+\bibitem[Mitschang et al.(2013)]{Mitschang_2013} Mitschang, A.~W., De Silva, G., Sharma, S., \& Zucker, D.~B.\ 2013, \mnras, 428, 2321 % 10.1093/mnras/sts194
+
+\bibitem[Mitschang et al.(2014)]{Mitschang_2014} Mitschang, A.~W., De Silva, G., Zucker, D.~B., et al.\ 2014, \mnras, 438, 2753  % 10.1093/mnras/stt2320
+
+\bibitem[Oliver et al.(1996)]{Oliver_1996} Oliver, J.~J., Baxter, R.~A., Wallace, C.~S.\ 1996, Unsupervised Learning using MML, Proceedings of the 13th International Conference on Machine Learning, Morgan Kaufmann, 364-372
 
@@
 
-\bibitem[Shannon (1948)]{Shannon_1948} Shannon, C.~E.\ 1948, Bell System Technical Journal, 27, 3 % 10.1002/j.1538-7305.1948.tb01338.x
+\bibitem[Roberts et al.(1998)]{Roberts_1998} Roberts, S., Husmeier, D., Rezek, I., Penny, W.\ 1998, Bayesian approaches to Gaussian mixture modeling.
+IEEE Transactions on Pattern Analysis and Machine Intelligence, 20, 11, 1133-1142
 
-\bibitem[Schafer (1997)]{Schafer_1997} Schafer, J.~L.\ 1997, Analysis of Incomplete Multivariate Data (CRC Press) % 10.1002/(SICI)1097-0258(20000415)19:7<1006::AID-SIM384>3.0.CO;2-T
+\bibitem[Shannon(1948)]{Shannon_1948} Shannon, C.~E.\ 1948, Bell System Technical Journal, 27, 3 % 10.1002/j.1538-7305.1948.tb01338.x
 
-\bibitem[Wallace \& Boulton (1968)]{Wallace_1968} Wallace, C.~S. \& Boulton, D.~M.\ 1968, The Computer Journal, 11, 2 %10.1093/comjnl/11.2.185
+\bibitem[Schafer(1997)]{Schafer_1997} Schafer, J.~L.\ 1997, Analysis of Incomplete Multivariate Data (CRC Press) % 10.1002/(SICI)1097-0258(20000415)19:7<1006::AID-SIM384>3.0.CO;2-T
+
+\bibitem[Wallace \& Boulton(1968)]{Wallace_1968} Wallace, C.~S. \& Boulton, D.~M.\ 1968, The Computer Journal, 11, 2 %10.1093/comjnl/11.2.185
+
+\bibitem[Wallace \& Freeman(1987)]{Wallace_1987} Wallace, C.~S. \& Freeman, P.~R.\ 1987, Journal of the Royal Statistical Society. Series B (Methodological), 49, 3, 240-265
+
+\bibitem[Wallace(2005)]{Wallace_2005} Wallace, C.~S.\ 2005, Statistical and inductive inference using minimum message length. Springer-Verlag, NJ, USA
 

### 2017-05-01 .ci/travis.sh
minor

@@ -1,11 +1,11 @@
 #!/bin/bash -x
 
 # Build the paper(s)
-cd articles/chemical-tagging-gmm
+cd $TRAVIS_BUILD_DIR/articles/chemical-tagging-gmm
 make
-cd articles/nips-2017
+cd $TRAVIS_BUILD_DIR/articles/nips-2017
 make
-cd articles/gmm-search
+cd $TRAVIS_BUILD_DIR/articles/gmm-search
 make
 
 # Push to GitHub

### 2017-05-01 .ci/travis.sh
skip nips

@@ -3,8 +3,8 @@
 # Build the paper(s)
 cd $TRAVIS_BUILD_DIR/articles/chemical-tagging-gmm
 make
-cd $TRAVIS_BUILD_DIR/articles/nips-2017
-make
+#cd $TRAVIS_BUILD_DIR/articles/nips-2017
+#make
 cd $TRAVIS_BUILD_DIR/articles/gmm-search
 make
 

### 2017-05-02 sandbox_mixture_slf.py
SLF with clusters working, even on heteroskedastic data

@@ -4,12 +4,12 @@ from snob import mixture_slf as slf
 
 #y = np.loadtxt("cluster_abundances.txt")
 
-n_samples, n_features, n_clusters, rank = 1000, 50, 4, 1
-sigma = 0.1
-true_specific_variances = sigma**2 * np.ones((1, n_features))
+n_samples, n_features, n_clusters, rank = 1000, 50, 3, 1
+sigma = 1
+true_homo_specific_variances = sigma**2 * np.ones((1, n_features))
 
 
-rng = np.random.RandomState(100)
+rng = np.random.RandomState(321)
 
 U, _, _ = np.linalg.svd(rng.randn(n_features, n_features))
 true_factor_loads = U[:, :rank].T
@@ -31,7 +31,7 @@ X_homo = X + sigma * bar
 # Adding heteroscedastic noise
 sigmas = sigma * rng.rand(n_features) + sigma / 2.
 X_hetero = X + rng.randn(n_samples, n_features) * sigmas
-
+true_hetero_specific_variances = sigmas**2
 
 #y = np.atleast_2d(np.loadtxt("coffee_example.txt")).T
 
@@ -42,22 +42,25 @@ model = slf.SLFGaussianMixture(n_clusters)
 #model.true_factor_loads = true_factor_loads
 #model.true_means = true_means
 #model.true_specific_variances = true_specific_variances
-model.fit(X_homo)
+model.fit(X_hetero)
 
 #model.initialize_parameters(X_homo)
 
-fig, ax = plt.subplots()
-ax.scatter(true_factor_loads, model.factor_loads)
+def scatter_common(x, y, title=None):
+    fig, ax = plt.subplots()
+    ax.scatter(x,y)
+    ax.set_title(title or "")
+    limits = np.array([ax.get_xlim(), ax.get_ylim()])
+    limits = (limits.min(), limits.max())
+    ax.plot(limits, limits, c="#666666", linestyle=":", linewidth=0.5, zorder=-1)
+    ax.set_xlim(limits)
+    ax.set_ylim(limits)
+    
+    return fig
 
-fig, ax = plt.subplots()
-ax.scatter(true_factor_scores, model.factor_scores)
-
-
-fig, ax = plt.subplots()
-ax.scatter(true_specific_variances, model.specific_variances)
-
-
-fig, ax = plt.subplots()
+scatter_common(true_factor_loads, model.factor_loads, "factor loads")
+scatter_common(true_factor_scores, model.factor_scores, "factor scores")
[…]

### 2017-05-02 snob/mixture_slf.py

@@ -74,7 +74,7 @@ class SLFGaussianMixture(object):
         self.initialization_method = initialization_method
         self.meta = {}
 
-        self.set_parameters(*([None] * len(self.parameter_names)))
+        self.set_parameters(**dict([(p, None) for p in self.parameter_names]))
 
         return None
 
@@ -85,6 +85,14 @@ class SLFGaussianMixture(object):
         Return the means, :math:`\mu_k`.
         """
         return self._means
+
+
+    @property
+    def weights(self):
+        r"""
+        Return the mixture weights, :math:`\w_k`.
+        """
+        return self._weights
 
 
     @property
@@ -134,6 +142,8 @@ class SLFGaussianMixture(object):
         # the square-root of the principal eigenvalue
         b = V[0] * S[0]**0.5
 
+        #b = np.random.uniform(size=b.shape) - 0.5
+
         specific_sigmas = np.sqrt(np.diag(np.dot(w.T, w)) \
                     / ((b*b + 1.0) * (N - 1)))
         factor_loads = specific_sigmas * b
@@ -142,12 +152,14 @@ class SLFGaussianMixture(object):
         b_sq = np.sum(b**2)
         factor_scores = np.dot(y, b) * (1 - K/(N - 1) * b_sq)/(1 + b_sq)
         specific_variances = specific_sigmas**2
+        specific_variances = specific_variances.reshape((1, -1))
 
         # Subtract off the factor loads and factor scores, then estimate the
         # cluster properties.
-        factor_scores = np.atleast_2d(factor_scores)
-        factor_loads = np.atleast_2d(factor_loads)
-        y_nlf = data - np.dot(factor_scores.T, factor_loads)
+        factor_scores = np.atleast_2d(factor_scores).reshape((-1, 1))
+        factor_loads = np.atleast_2d(factor_loads).reshape((1, -1))
+
+        y_nlf = data - np.dot(factor_scores, factor_loads)
 
         if self.initialization_method == "kmeans":
 
@@ -168,24 +180,29 @@ class SLFGaussianMixture(object):
             raise ValueError("Unknown initialization method: {}".format(
                 self.initialization_method))
 
-        # Return the estimate of the model parameters given the responsibility
-        # matrix.
-        return self._maximization(y, responsibility.T)
+        # Calculate the weights from the responsibility initialisation.
+        effective_membership = np.sum(responsibility, axis=0)
+        weights = (effective_membership + 0.5)/(N + K/2.0)
+
[…]

### 2017-05-14 snob/nips_search.py
diag and other stuff

@@ -22,7 +22,7 @@ logger = logging.getLogger(__name__)
 
 
 
-def _total_parameters(K, D):
+def _total_parameters(K, D, covariance_type):
     r"""
     Return the total number of model parameters :math:`Q`, if a full 
     covariance matrix structure is assumed.
@@ -428,7 +428,7 @@ class GaussianMixture(BaseGaussianMixture):
                 y, responsibility=responsibility)
 
         # Store the mixture.
-        slogdet = np.sum(np.log(np.linalg.det(mixture.covariance)))
+        slogdet = np.sum(_slogdet(mixture.covariance, mixture.covariance_type))
         self._proposed_mixtures.append(mixture)
         self._mixture_predictors.append([
             mixture.weight.size,
@@ -501,7 +501,8 @@ class GaussianMixture(BaseGaussianMixture):
         weight = np.ones(1)
         mean = np.mean(y, axis=0).reshape((1, -1))
 
-        covariance = _estimate_covariance_matrix(y, np.ones((1, 1)), mean,
+        N, D = y.shape
+        covariance = _estimate_covariance_matrix(y, np.ones((1, N)), mean,
             self.covariance_type, self.covariance_regularization)
 
         # Set parameters.
@@ -547,13 +548,13 @@ class GaussianMixture(BaseGaussianMixture):
 
         # Calculate the change in message length.
         current_ll = np.max(predictors.T[2][predictors.T[0] == current_K])
-        sign, slogdet = np.linalg.slogdet(self.covariance)
+        slogdet = _slogdet(self.covariance, self.covariance_type)
 
         dI_expectation = dK * (
             (1 - D/2.0)*np.log(2) + 0.25 * (D*(D+3) + 2)*np.log(N/(2*np.pi))) \
             + 0.5 * (D*(D+3)/2.0 - 1) * (slw_expectation - np.sum(np.log(self.weight))) \
             - np.array([np.sum(np.log(current_K + np.arange(_))) for _ in dK])\
-            + 0.5 * np.log(_total_parameters(K, D)/float(_total_parameters(current_K, D))) \
+            + 0.5 * np.log(_total_parameters(K, D, self.covariance_type)/float(_total_parameters(current_K, D, self.covariance_type))) \
             - (D + 2)/2.0 * (np.sum(slogdet)) \
             + current_ll + nll_mslogdetcov_expectation
         
[…]

### 2017-05-14 sandbox_nips_search.py
searcher'

@@ -40,79 +40,29 @@ y = np.vstack([np.random.multivariate_normal(
     for i in range(len(weight))])
 
 
-
+#
 #y = np.loadtxt("cluster_example.txt")
 
-#y = np.loadtxt("birch3.txt")
-
-
-#n_samples = 500
-
-# Generate random sample, two components
-#np.random.seed(0)
-#C = np.array([[0., -0.1], [1.7, .4]])
-#X = np.r_[np.dot(np.random.randn(n_samples, 2), C),
-#          .7 * np.random.randn(n_samples, 2) + np.array([-6, 3])]
-#y = X
-
+y = np.loadtxt("s4.txt")
 
 #from sklearn import datasets
-#iris = datasets.load_iris()
-#y = iris.data
+#y, _ = datasets.make_blobs(n_samples=1000, n_features=2, centers=16)
 
-"""
-Approximating \sum\log{w_k}...
-
-bar = []
-for i in range(1, 101):
-
-    foo = np.random.uniform(size=(1000, i))
-    foo = foo.T/np.sum(foo, axis=1)
-    print(i)
-
-    mean = np.mean(np.log(foo).sum(axis=0))
-    std = np.std(np.log(foo).sum(axis=0))
-
-    bar.append([i, mean, std])
-
-    
-
-bar = np.array(bar)
-
-fig, ax = plt.subplots()
-ax.scatter(bar.T[0], bar.T[2])
-#ax.scatter(bar.T[0], bar.T[1])
-#ax.errorbar(bar.T[0], bar.T[1], bar.T[2], fmt=None)
-
-raise a
-"""
-
-"""
-# Approximating \log(\sum{r}) (the log of the effective memberships...)
-bar = []
-for i in range(1, 101):
-
-    foo = np.random.uniform(size=(1000, i))
-    foo = foo.T/np.sum(foo, axis=1)
-    foo *= 900 # The sample size, say.
-
-    mean = np.mean(np.log(foo).sum(axis=0))
-    std = np.std(np.log(foo).sum(axis=0))
-
-    bar.append([i, mean, std])
+#y = np.loadtxt("birch2.txt")
 
 
-bar = np.array(bar)
-
-fig, axes = plt.subplots(2)
-axes[0].scatter(bar.T[0], bar.T[1])
-axes[1].scatter(bar.T[0], bar.T[2])
-
-raise a
-"""
-
-search_model = mixture.GaussianMixture()
[…]

### 2017-05-15 sandbox_nips_search.py
shit

@@ -1,6 +1,6 @@
 
 import numpy as np
-from snob import nips_search as mixture
+from snob import nips_search2 as mixture
 
 
 
@@ -43,16 +43,17 @@ y = np.vstack([np.random.multivariate_normal(
 #
 #y = np.loadtxt("cluster_example.txt")
 
-y = np.loadtxt("s4.txt")
+#y = np.loadtxt("s4.txt")
 
-#from sklearn import datasets
-#y, _ = datasets.make_blobs(n_samples=1000, n_features=2, centers=16)
+from sklearn import datasets
+y, _ = datasets.make_blobs(n_samples=1000, n_features=30, centers=10,
+  cluster_std=0.5, center_box=(-100, 100))
 
 #y = np.loadtxt("birch2.txt")
 
 
-search_model = mixture.GaussianMixture(covariance_type="full", predict_mixtures=100)
-search_model.search(y)
+search_model = mixture.GaussianMixture(covariance_type="full", predict_mixtures=1)
+search_model.growing_search(y)
 
 
 raise a

### 2017-05-15 snob/nips_search.py

@@ -775,33 +775,45 @@ class GaussianMixture(BaseGaussianMixture):
         logger.debug("Re-initializing with K-means++ at K = {}".format(K))
 
         # Initialize new centroids by k-means++
-        mean = kmeans._k_init(y, K, kmeans.row_norms(y, squared=True),
-            kmeans.check_random_state(None))
+        mixtures = []
+        mls = []
+        for z in range(30):
 
-        # Calculate weights by L2 distances to closest centers.
-        distance = np.sum((y[:, :, None] - mean.T)**2, axis=1).T
+            mean = kmeans._k_init(y, K, kmeans.row_norms(y, squared=True),
+                kmeans.check_random_state(None))
 
-        N, D = y.shape
-        responsibility = np.zeros((K, N))
-        responsibility[np.argmin(distance, axis=0), np.arange(N)] = 1.0
+            # Calculate weights by L2 distances to closest centers.
+            distance = np.sum((y[:, :, None] - mean.T)**2, axis=1).T
 
-        weight = responsibility.sum(axis=1)/N
+            N, D = y.shape
+            responsibility = np.zeros((K, N))
+            responsibility[np.argmin(distance, axis=0), np.arange(N)] = 1.0
 
-        covariance = _estimate_covariance_matrix(y, responsibility, mean,
-            self.covariance_type, self.covariance_regularization)
+            weight = responsibility.sum(axis=1)/N
 
-        mixture = self.__class__(
-            threshold=self.threshold,
-            covariance_type=self.covariance_type,
-            max_em_iterations=self.max_em_iterations,
-            covariance_regularization=self.covariance_regularization)
+            covariance = _estimate_covariance_matrix(y, responsibility, mean,
+                self.covariance_type, self.covariance_regularization)
 
-        # Initialize it.
-        mixture.set_parameters(mean=mean, weight=weight, covariance=covariance)
+            mixture = self.__class__(
+                threshold=self.threshold,
+                covariance_type=self.covariance_type,
+                max_em_iterations=self.max_em_iterations,
+                covariance_regularization=self.covariance_regularization)
 
-        # Run E-M on the partial mixture.
-        R, meta = mixture._expectation_maximization(
-            y, parent_responsibility=responsibility)
+            # Initialize it.
+            mixture.set_parameters(mean=mean, weight=weight, covariance=covariance)
+
+            # Run E-M on the partial mixture.
[…]

### 2017-05-16 sandbox_nips_search.py
fucking :bugs:

@@ -1,7 +1,7 @@
 
 import numpy as np
-from snob import nips_search2 as mixture
-
+from snob import nips_search3 as mixture
+#from snob import mixture_ka as mixture
 
 
 # Generate data from the example in Section 13.4 of P & A (2015)
@@ -41,19 +41,19 @@ y = np.vstack([np.random.multivariate_normal(
 
 
 #
-#y = np.loadtxt("cluster_example.txt")
+y = np.loadtxt("cluster_example.txt")
 
-#y = np.loadtxt("s4.txt")
+y = np.loadtxt("s4.txt")
 
-from sklearn import datasets
-y, _ = datasets.make_blobs(n_samples=1000, n_features=30, centers=10,
-  cluster_std=0.5, center_box=(-100, 100))
+#from sklearn import datasets
+#y, _ = datasets.make_blobs(n_samples=1000, n_features=2, centers=10,
+#  cluster_std=0.5, center_box=(-100, 100))
 
 #y = np.loadtxt("birch2.txt")
 
 
-search_model = mixture.GaussianMixture(covariance_type="full", predict_mixtures=1)
-search_model.growing_search(y)
+search_model = mixture.GaussianMixture(covariance_type="full", predict_mixtures=50)
+foo = search_model.search(y)
 
 
 raise a
@@ -68,6 +68,8 @@ covariance_type = "full"
 
 
 raise  a
+
+from snob import mixture_ka as mixture
 
 model1 = mixture.GaussianMixture()
 mu_1, cov_1, weight_1, meta_1 = model1.fit(y, 1)
@@ -102,7 +104,8 @@ expected_delta_I = np.log(2) \
 expected_delta_I = expected_delta_I/np.log(2)
 
 dk = 1
-expected_delta_I2 = dk * ((1 - D/2.) * np.log(2) + 0.25 * (D*(D+3) + 2) * np.log(N/(2*np.pi))) \
+expected_delta_I2 = dk * (
+  (1 - D/2.) * np.log(2) + 0.25 * (D*(D+3) + 2) * np.log(N/(2*np.pi))) \
                   + 0.5 * (D*(D+3)/2. - 1) * (np.sum(np.log(weight_2)) - np.sum(np.log(weight_1))) \
                   - np.sum([np.log(K + _) for _ in range(dk)]) \
                   - meta_2["log_likelihood"].sum() + meta_1["log_likelihood"].sum() \

### 2017-05-16 snob/nips_search3.py

@@ -1,0 +1,1549 @@
+
+"""
+An estimator for modelling data from a mixture of Gaussians, 
+using an objective function based on minimum message length.
+"""
+
+__all__ = [
+    "GaussianMixture", 
+    "kullback_leibler_for_multivariate_normals",
+    "responsibility_matrix",
+    "split_component", "merge_component", "delete_component", 
+] 
+import logging
+import numpy as np
+import scipy
+from scipy import stats
+import scipy.optimize as op
+from collections import defaultdict
+from sklearn.cluster import k_means_ as kmeans
+
+logger = logging.getLogger(__name__)
+
+
+class GaussianMixture(object):
+
+    r"""
+    Model data from (potentially) many multivariate Gaussian distributions, 
+    using minimum message length (MML) as the objective function.
+
+    :param covariance_type: [optional]
+        The structure of the covariance matrix for individual components.
+        The available options are: `full` for a free covariance matrix, or
+        `diag` for a diagonal covariance matrix (default: ``diag``).
+
+    :param covariance_regularization: [optional]
+        Regularization strength to add to the diagonal of covariance matrices
+        (default: ``0``).
+
+    :param threshold: [optional]
+        The relative improvement in message length required before stopping an
+        expectation-maximization step (default: ``1e-5``).
+
+    :param max_em_iterations: [optional]
+        The maximum number of iterations to run per expectation-maximization
+        loop (default: ``10000``).
+    """
+
+    parameter_names = ("mean", "covariance", "weight")
+
+    def __init__(self, covariance_type="diag", covariance_regularization=0, 
+        threshold=1e-5, max_em_iterations=10000, predict_mixtures=10,
+        **kwargs):
+
+        available = ("full", "diag", )
+        covariance_type = covariance_type.strip().lower()
+        if covariance_type not in available:
+            raise ValueError("covariance type '{}' is invalid. "\
+                             "Must be one of: {}".format(
+                                covariance_type, ", ".join(available)))
+
+        if 0 > covariance_regularization:
+            raise ValueError(
+                "covariance_regularization must be a non-negative float")
+
+        if 0 >= threshold:
+            raise ValueError("threshold must be a positive value")
+
+        if 1 > max_em_iterations:
+            raise ValueError("max_em_iterations must be a positive integer")
+
+        self._threshold = threshold
+        self._max_em_iterations = max_em_iterations
[…]

### 2017-05-16 sandbox_nips_search.py
finally ffs

@@ -43,17 +43,28 @@ y = np.vstack([np.random.multivariate_normal(
 #
 y = np.loadtxt("cluster_example.txt")
 
-y = np.loadtxt("s4.txt")
+#y = np.loadtxt("s4.txt")
 
-#from sklearn import datasets
-#y, _ = datasets.make_blobs(n_samples=1000, n_features=2, centers=10,
-#  cluster_std=0.5, center_box=(-100, 100))
-
-#y = np.loadtxt("birch2.txt")
+from sklearn import datasets
+y, _ = datasets.make_blobs(n_samples=1000, n_features=2, centers=50,
+  cluster_std=1, center_box=(-20, 20))
 
 
-search_model = mixture.GaussianMixture(covariance_type="full", predict_mixtures=50)
+
+#y = np.loadtxt("a3.txt")
+
+
+search_model = mixture.GaussianMixture(covariance_type="full", predict_mixtures=10)
 foo = search_model.search(y)
+
+
+fig, ax = plt.subplots()
+ax.scatter(y.T[0], y.T[1])
+
+means = foo[0]
+scat = ax.scatter(means.T[0], means.T[1], c=foo[2], edgecolor="k", s=50)
+plt.colorbar(scat)
+
 
 
 raise a

### 2017-05-16 snob/nips_search3.py

@@ -47,7 +47,7 @@ class GaussianMixture(object):
 
     parameter_names = ("mean", "covariance", "weight")
 
-    def __init__(self, covariance_type="diag", covariance_regularization=0, 
+    def __init__(self, covariance_type="diag", covariance_regularization=1e-6, 
         threshold=1e-5, max_em_iterations=10000, predict_mixtures=10,
         **kwargs):
 
@@ -152,7 +152,7 @@ class GaussianMixture(object):
                 # Exhaustively delete all components.
                 for m in range(M):
                     p = delete_component(y, mu, cov, weight, R, m, **kwds)
-
+                    
                     # Keep best deleted component.
                     if p[-1] < best_perturbations["delete"][-1]:
                         best_perturbations["delete"] = [m] + list(p)
@@ -208,12 +208,30 @@ class GaussianMixture(object):
 
         while True:
 
+
             M = weight.size
+            
             best_perturbations = defaultdict(lambda: [np.inf])
-                
+            logger.debug("K = {}, I = {}, ll = {}".format(M, message_length, ll))
+            logger.debug("Smallest Neff = {} ({})".format(
+                np.min(weight * y.shape[0]), np.argmin(weight * y.shape[0])))
+
             # Exhaustively split all components.
             for m in range(M):
-                p = split_component(y, mu, cov, weight, R, m, **kwds)
+                try:
+                    p = split_component(y, mu, cov, weight, R, m, **kwds)
+                except ValueError:
+                    logger.debug("Failed to split component {}".format(m))
+                    continue
+
+                logger.debug("Split component {} of {}: {}".format(m, M, p[-1]))
+
+                # If the split resulted in a maximum Neff of < 1, don't keep
+                # this.
+                if np.min(p[2] * y.shape[0]) < 1:
+                    logger.debug("Ignoring mixture because Neff < 1")
+                    continue
+
                 self._mixture_predictors.append(p[-2]["predictors"])
 
                 # Keep best split component.
@@ -223,7 +241,19 @@ class GaussianMixture(object):
             if M > 1:
                 # Exhaustively delete all components.
                 for m in range(M):
-                    p = delete_component(y, mu, cov, weight, R, m, **kwds)
+                    try:
[…]

### 2017-05-16 articles/nips-2017/ms.tex
stuff

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
% \PassOptionsToPackage{numbers, compress}{natbib}
% before loading nips_2017
%
% to avoid loading the natbib package, add option nonatbib:
% \usepackage[nonatbib]{nips_2017}

%\usepackage{nips_2017}

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2017}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\title{
  Unsupervised learning of Gaussian mixture models in linear time
  using variational Bayesian methods
}

% The \author macro works with any number of authors. There are two
% commands used to separate the names and addresses of multiple
% authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to
% break the lines. Using \AND forces a line break at that point. So,
% if LaTeX puts 3 of 4 authors names on the first line, and the last
% on the second line, try using \AND instead of \And before the third
% author name.

\author{
  Andrew R.~Casey\\
  Monash University\\
  \texttt{andrew.casey@monash.edu}\\
  %% examples of more authors
  \And 
  Aldeida Aleti\\
  Monash University\\
  \texttt{aldeida.aleti@monash.edu}\\
  \AND
  David Dowe\\
  Monash University\\
  \texttt{david.dowe@monash.edu}\\
  \And
  John C.~Lattanzio\\
  Monash University\\
  \texttt{john.lattanzio@monash.edu}\\
}

\begin{document}
% \nipsfinalcopy is no longer used

\maketitle

\begin{abstract}
  The number of components in a Gaussian mixture model is usually determined
  by applying a chosen penalty heuristic.
  This approach requires optimizing a set of mixtures of increasing complexity,
  and unnecessarily separates model selection from parameter estimation.

  We derive an objective function for a finite Gaussian mixture model using
  Minimum Message Length (MML), and derive an expression for future mixtures
  that have yet to be observed.

  Although future mixtures have yet to be trialled, we show that we can
  calculate probability distributions for those future mixtures, allowing 
  for a \emph{jumping} search algorithm that simultaneously performs model
  selection and parameter estimation for Gaussian mixtures in linear time.

  The abstract paragraph should be indented \nicefrac{1}{2}~inch
  (3~picas) on both the left- and right-hand margins. Use 10~point
  type, with a vertical spacing (leading) of 11~points.  The word
  \textbf{Abstract} must be centered, bold, and in point size 12. Two
  line spaces precede the abstract. The abstract must be limited to
  one paragraph.
\end{abstract}

\section{Submission of papers to NIPS 2017}

NIPS requires electronic submissions.  The electronic submission site
is
\begin{center}
  \url{https://cmt.research.microsoft.com/NIPS2017/}
\end{center}

Please read carefully the instructions below and follow them
faithfully.

\subsection{Style}

Papers to be submitted to NIPS 2017 must be prepared according to the
instructions presented here. Papers may only be up to eight pages
long, including figures. Since 2009 an additional ninth page
\emph{containing only acknowledgments and/or cited references} is
allowed. Papers that exceed nine pages will not be reviewed, or in any
other way considered for presentation at the conference.

The margins in 2017 are the same as since 2007, which allow for
$\sim$$15\%$ more words in the paper compared to earlier years.

Authors are required to use the NIPS \LaTeX{} style files obtainable
at the NIPS website as indicated below. Please make sure you use the
current files and not previous versions. Tweaking the style files may
be grounds for rejection.

\subsection{Retrieval of style files}

The style files for NIPS and other conference information are
available on the World Wide Web at
\begin{center}
  \url{http://www.nips.cc/}
\end{center}
The file \verb+nips_2017.pdf+ contains these instructions and
illustrates the various formatting requirements your NIPS paper must
satisfy.

The only supported style file for NIPS 2017 is \verb+nips_2017.sty+,
rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{}
  2.09, Microsoft Word, and RTF are no longer supported!}

The new \LaTeX{} style file contains two optional arguments:
\verb+final+, which creates a camera-ready copy, and \verb+nonatbib+,
which will not load the \verb+natbib+ package for you in case of
package clash.

At submission time, please omit the \verb+final+ option. This will
anonymize your submission and add line numbers to aid review.  Please
do \emph{not} refer to these line numbers in your paper as they will
be removed during generation of camera-ready copies.

The file \verb+nips_2017.tex+ may be used as a ``shell'' for writing
your paper. All you have to do is replace the author, title, abstract,
and text of the paper with your own.

The formatting instructions contained in these style files are
summarized in Sections \ref{gen_inst}, \ref{headings}, and
\ref{others} below.

\section{General formatting instructions}
\label{gen_inst}

The text must be confined within a rectangle 5.5~inches (33~picas)
wide and 9~inches (54~picas) long. The left margin is 1.5~inch
(9~picas).  Use 10~point type with a vertical spacing (leading) of
11~points.  Times New Roman is the preferred typeface throughout, and
will be selected for you by default.  Paragraphs are separated by
\nicefrac{1}{2}~line space (5.5 points), with no indentation.

The paper title should be 17~point, initial caps/lower case, bold,
centered between two horizontal rules. The top rule should be 4~points
thick and the bottom rule should be 1~point thick. Allow
\nicefrac{1}{4}~inch space above and below the title to rules. All
pages should start at 1~inch (6~picas) from the top of the page.

For the final version, authors' names are set in boldface, and each
name is centered above the corresponding address. The lead author's
name is to be listed first (left-most), and the co-authors' names (if
different address) are set to follow. If there is only one co-author,
list both author and co-author side by side.

Please pay special attention to the instructions in Section \ref{others}
regarding figures, tables, acknowledgments, and references.

\section{Headings: first level}
\label{headings}

All headings should be lower case (except for first word and proper
nouns), flush left, and bold.

First-level headings should be in 12-point type.

\subsection{Headings: second level}

Second-level headings should be in 10-point type.

\subsubsection{Headings: third level}

Third-level headings should be in 10-point type.

\paragraph{Paragraphs}

There is also a \verb+\paragraph+ command available, which sets the
heading in bold, flush left, and inline with the text, with the
heading followed by 1\,em of space.

\section{Citations, figures, tables, references}
\label{others}

These instructions apply to everyone.

\subsection{Citations within the text}

The \verb+natbib+ package will be loaded for you by default.
Citations may be author/year or numeric, as long as you maintain
internal consistency.  As to the format of the references themselves,
any style is acceptable as long as it is used consistently.

The documentation for \verb+natbib+ may be found at
\begin{center}
  \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
\end{center}
Of note is the command \verb+\citet+, which produces citations
appropriate for use in inline text.  For example,
\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}
produces
\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

If you wish to load the \verb+natbib+ package with options, you may
add the following before loading the \verb+nips_2017+ package:
\begin{verbatim}
   \PassOptionsToPackage{options}{natbib}
\end{verbatim}

If \verb+natbib+ clashes with another package you load, you can add
the optional argument \verb+nonatbib+ when loading the style file:
\begin{verbatim}
   \usepackage[nonatbib]{nips_2017}
\end{verbatim}

As submission is double blind, refer to your own published work in the
third person. That is, use ``In the previous work of Jones et
al.\ [4],'' not ``In our previous work [4].'' If you cite your other
papers that are not widely available (e.g., a journal paper under
review), use anonymous author names in the citation, e.g., an author
of the form ``A.\ Anonymous.''

\subsection{Footnotes}

Footnotes should be used sparingly.  If you do require a footnote,
indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Place the footnotes at the bottom of the
page on which they appear.  Precede the footnote with a horizontal
rule of 2~inches (12~picas).

Note that footnotes are properly typeset \emph{after} punctuation
marks.\footnote{As in this example.}

\subsection{Figures}

All artwork must be neat, clean, and legible. Lines should be dark
enough for purposes of reproduction. The figure number and caption
always appear after the figure. Place one line space before the figure
caption and one line space after the figure. The figure caption should
be lower case (except for first word and proper nouns); figures are
numbered consecutively.

You may use color figures.  However, it is best for the figure
captions and the paper body to be legible if the paper is printed in
either black/white or in color.
\begin{figure}[h]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

\subsection{Tables}

All tables must be centered, neat, clean and legible.  The table
number and title always appear before the table.  See
Table~\ref{sample-table}.

Place one line space before the table title, one line space after the
table title, and one line space after the table. The table title must
be lower case (except for first word and proper nouns); tables are
numbered consecutively.

Note that publication-quality tables \emph{do not contain vertical
  rules.} We strongly suggest the use of the \verb+booktabs+ package,
which allows for typesetting high-quality, professional tables:
\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}
This package was used to typeset Table~\ref{sample-table}.

\begin{table}[t]
  \caption{Sample table title}
  \label{sample-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule{1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Final instructions}

Do not change any aspects of the formatting parameters in the style
files.  In particular, do not modify the width or length of the
rectangle the text should fit into, and do not change font sizes
(except perhaps in the \textbf{References} section; see below). Please
note that pages should be numbered.

\section{Preparing PDF files}

Please prepare submission files with paper size ``US Letter,'' and
not, for example, ``A4.''

Fonts were the main cause of problems in the past years. Your PDF file
must only contain Type 1 or Embedded TrueType fonts. Here are a few
instructions to achieve this.

\begin{itemize}

\item You should directly generate PDF files using \verb+pdflatex+.

\item You can check which fonts a PDF files uses.  In Acrobat Reader,
  select the menu Files$>$Document Properties$>$Fonts and select Show
  All Fonts. You can also use the program \verb+pdffonts+ which comes
  with \verb+xpdf+ and is available out-of-the-box on most Linux
  machines.

\item The IEEE has recommendations for generating PDF files whose
  fonts are also acceptable for NIPS. Please see
  \url{http://www.emfield.org/icuwb2010/downloads/IEEE-PDF-SpecV32.pdf}

\item \verb+xfig+ "patterned" shapes are implemented with bitmap
  fonts.  Use "solid" shapes instead.

\item The \verb+\bbold+ package almost always uses bitmap fonts.  You
  should use the equivalent AMS Fonts:
\begin{verbatim}
   \usepackage{amsfonts}
\end{verbatim}
followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or
\verb+\mathbb{C}+ for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You
can also use the following workaround for reals, natural and complex:
\begin{verbatim}
   \newcommand{\RR}{I\!\!R} %real numbers
   \newcommand{\Nat}{I\!\!N} %natural numbers
   \newcommand{\CC}{I\!\!\!\!C} %complex numbers
\end{verbatim}
Note that \verb+amsfonts+ is automatically loaded by the
\verb+amssymb+ package.

\end{itemize}

If your file contains type 3 fonts or non embedded TrueType fonts, we
will ask you to fix it.

\subsection{Margins in \LaTeX{}}

Most of the margin problems come from figures positioned by hand using
\verb+\special+ or other commands. We suggest using the command
\verb+\includegraphics+ from the \verb+graphicx+ package. Always
specify the figure width as a multiple of the line width as in the
example below:
\begin{verbatim}
   \usepackage[pdftex]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.pdf}
\end{verbatim}
See Section 4.4 in the graphics bundle documentation
(\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})

A number of width problems arise when \LaTeX{} cannot properly
hyphenate a line. Please give LaTeX hyphenation hints using the
\verb+\-+ command when necessary.

\subsubsection*{Acknowledgments}

Use unnumbered third level headings for the acknowledgments. All
acknowledgments go at the end of the paper. Do not include
acknowledgments in the anonymized submission, only in the final paper.

\section*{References}

References follow the acknowledgments. Use unnumbered first-level
heading for the references. Any choice of citation style is acceptable
as long as you are consistent. It is permissible to reduce the font
size to \verb+small+ (9 point) when listing the references. {\bf
  Remember that you can use a ninth page as long as it contains
  \emph{only} cited references.}
\medskip

\small

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms
for connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and
T.K.\ Leen (eds.), {\it Advances in Neural Information Processing
  Systems 7}, pp.\ 609--616. Cambridge, MA: MIT Press.

[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS:
  Exploring Realistic Neural Models with the GEneral NEural SImulation
  System.}  New York: TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of
learning and recall at excitatory recurrent synapses and cholinergic
modulation in rat hippocampal region CA3. {\it Journal of
  Neuroscience} {\bf 15}(7):5249-5262.

\end{document}

### 2017-05-16 articles/nips-2017/ms.tex
abstract

@@
   Unsupervised learning of Gaussian mixture models in linear time
-  using variational Bayesian methods
+  using a variational Bayesian method
 }
@@
   \texttt{andrew.casey@monash.edu}\\
-  %% examples of more authors
   \And 
@@
 \begin{abstract}
-  The number of components in a Gaussian mixture model is usually determined
-  by applying a chosen penalty heuristic.
-  This approach requires optimizing a set of mixtures of increasing complexity,
-  and unnecessarily separates model selection from parameter estimation.
-
-  We derive an objective function for a finite Gaussian mixture model using
-  Minimum Message Length (MML), and derive an expression for future mixtures
-  that have yet to be observed.
-
-  Although future mixtures have yet to be trialled, we show that we can
-  calculate probability distributions for those future mixtures, allowing 
-  for a \emph{jumping} search algorithm that simultaneously performs model
-  selection and parameter estimation for Gaussian mixtures in linear time.
-
-  The abstract paragraph should be indented \nicefrac{1}{2}~inch
-  (3~picas) on both the left- and right-hand margins. Use 10~point
-  type, with a vertical spacing (leading) of 11~points.  The word
-  \textbf{Abstract} must be centered, bold, and in point size 12. Two
-  line spaces precede the abstract. The abstract must be limited to
-  one paragraph.
+  The appropriate number of components in a Gaussian mixture model is usually 
+  determined by adopting a heuristic to penalize the number of free parameters.
+  This approach unnecessarily separates model selection from parameter estimation,
+  and requires a large set of mixtures with an increasing number of components
+  to be optimized.
+  We formulate an objective function for a finite Gaussian mixture model using
+  Minimum Message Length (MML), and derive an expression to approximate that
+  objective function for mixtures with any number of components. 
+  These approximations provide a probability density for the evaluated objective
+  function of future mixtures that have yet to be computed, allowing us to 
+  \emph{jump} towards the optimal number of components.
+  We show that our approach can simultaneously perform parameter estimation and 
+  model selection for multivariate Gaussian mixtures in linear time.
 \end{abstract}
 
-\section{Submission of papers to NIPS 2017}
+\section{Introduction}
+% Gaussian mixture modelling is a common problem.
+% When the number of mixtures is unknown, the problem is normally 
+% attacked by optimizing multiple different models, and performing
+% model selection after-the-fact.
+% Model selection is typically governed by a penalized heuristic,
+% e.g., BIC, AIC, etc.
 
-NIPS requires electronic submissions.  The electronic submission site
-is
-\begin{center}
-  \url{https://cmt.research.microsoft.com/NIPS2017/}
-\end{center}
+% Introducing MML
+% Different approaches in terms of where to start from:
+% Some things about how MML has been used to grow
+% Some example where a mixture has been used where K=N,
+% because merging/deleting components is computationally
+% more efficient than splitting.
 
-Please read carefully the instructions below and follow them
-faithfully.
+% In these approaches, the long-term memory of previously
+% considered mixtures is usually discarded.
+% Here we will outline an objective function using MML
+% We will show that the difference in the message length
+% between the current state and future unobserved states
+% can be approximated.
 
-\subsection{Style}
+\section{The objective function}
 
-Papers to be submitted to NIPS 2017 must be prepared according to the
-instructions presented here. Papers may only be up to eight pages
-long, including figures. Since 2009 an additional ninth page
-\emph{containing only acknowledgments and/or cited references} is
-allowed. Papers that exceed nine pages will not be reviewed, or in any
-other way considered for presentation at the conference.
+\section{Search strategy}
 
-The margins in 2017 are the same as since 2007, which allow for
-$\sim$$15\%$ more words in the paper compared to earlier years.
+% Initialization
 
-Authors are required to use the NIPS \LaTeX{} style files obtainable
-at the NIPS website as indicated below. Please make sure you use the
-current files and not previous versions. Tweaking the style files may
-be grounds for rejection.
+% Perturbation strategy with a twist
 
-\subsection{Retrieval of style files}
+\section{Experimental results}
 
-The style files for NIPS and other conference information are
-available on the World Wide Web at
-\begin{center}
-  \url{http://www.nips.cc/}
-\end{center}
-The file \verb+nips_2017.pdf+ contains these instructions and
-illustrates the various formatting requirements your NIPS paper must
-satisfy.
+%
 
-The only supported style file for NIPS 2017 is \verb+nips_2017.sty+,
-rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{}
-  2.09, Microsoft Word, and RTF are no longer supported!}
+% BAYES FTW
 
-The new \LaTeX{} style file contains two optional arguments:
-\verb+final+, which creates a camera-ready copy, and \verb+nonatbib+,
-which will not load the \verb+natbib+ package for you in case of
-package clash.
+\section{Conclusion}
 
-At submission time, please omit the \verb+final+ option. This will
-anonymize your submission and add line numbers to aid review.  Please
-do \emph{not} refer to these line numbers in your paper as they will
-be removed during generation of camera-ready copies.
-
-The file \verb+nips_2017.tex+ may be used as a ``shell'' for writing
-your paper. All you have to do is replace the author, title, abstract,
-and text of the paper with your own.
-
-The formatting instructions contained in these style files are
-summarized in Sections \ref{gen_inst}, \ref{headings}, and
-\ref{others} below.
-
-\section{General formatting instructions}
-\label{gen_inst}
-
-The text must be confined within a rectangle 5.5~inches (33~picas)
-wide and 9~inches (54~picas) long. The left margin is 1.5~inch
-(9~picas).  Use 10~point type with a vertical spacing (leading) of
-11~points.  Times New Roman is the preferred typeface throughout, and
-will be selected for you by default.  Paragraphs are separated by
-\nicefrac{1}{2}~line space (5.5 points), with no indentation.
-
-The paper title should be 17~point, initial caps/lower case, bold,
-centered between two horizontal rules. The top rule should be 4~points
-thick and the bottom rule should be 1~point thick. Allow
-\nicefrac{1}{4}~inch space above and below the title to rules. All
-pages should start at 1~inch (6~picas) from the top of the page.
-
-For the final version, authors' names are set in boldface, and each
-name is centered above the corresponding address. The lead author's
-name is to be listed first (left-most), and the co-authors' names (if
-different address) are set to follow. If there is only one co-author,
-list both author and co-author side by side.
-
-Please pay special attention to the instructions in Section \ref{others}
-regarding figures, tables, acknowledgments, and references.
-
-\section{Headings: first level}
-\label{headings}
-
-All headings should be lower case (except for first word and proper
-nouns), flush left, and bold.
-
-First-level headings should be in 12-point type.
-
-\subsection{Headings: second level}
-
-Second-level headings should be in 10-point type.
-
-\subsubsection{Headings: third level}
-
-Third-level headings should be in 10-point type.
-
-\paragraph{Paragraphs}
-
-There is also a \verb+\paragraph+ command available, which sets the
-heading in bold, flush left, and inline with the text, with the
-heading followed by 1\,em of space.
-
-\section{Citations, figures, tables, references}
-\label{others}
-
-These instructions apply to everyone.
-
-\subsection{Citations within the text}
-
-The \verb+natbib+ package will be loaded for you by default.
-Citations may be author/year or numeric, as long as you maintain
-internal consistency.  As to the format of the references themselves,
-any style is acceptable as long as it is used consistently.
-
-The documentation for \verb+natbib+ may be found at
-\begin{center}
-  \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
-\end{center}
-Of note is the command \verb+\citet+, which produces citations
-appropriate for use in inline text.  For example,
-\begin{verbatim}
-   \citet{hasselmo} investigated\dots
-\end{verbatim}
-produces
-\begin{quote}
-  Hasselmo, et al.\ (1995) investigated\dots
-\end{quote}
-
-If you wish to load the \verb+natbib+ package with options, you may
-add the following before loading the \verb+nips_2017+ package:
-\begin{verbatim}
-   \PassOptionsToPackage{options}{natbib}
-\end{verbatim}
-
-If \verb+natbib+ clashes with another package you load, you can add
-the optional argument \verb+nonatbib+ when loading the style file:
-\begin{verbatim}
-   \usepackage[nonatbib]{nips_2017}
-\end{verbatim}
-
-As submission is double blind, refer to your own published work in the
-third person. That is, use ``In the previous work of Jones et
-al.\ [4],'' not ``In our previous work [4].'' If you cite your other
-papers that are not widely available (e.g., a journal paper under
-review), use anonymous author names in the citation, e.g., an author
-of the form ``A.\ Anonymous.''
-
-\subsection{Footnotes}
-
-Footnotes should be used sparingly.  If you do require a footnote,
-indicate footnotes with a number\footnote{Sample of the first
-  footnote.} in the text. Place the footnotes at the bottom of the
-page on which they appear.  Precede the footnote with a horizontal
-rule of 2~inches (12~picas).
-
-Note that footnotes are properly typeset \emph{after} punctuation
-marks.\footnote{As in this example.}
-
-\subsection{Figures}
-
-All artwork must be neat, clean, and legible. Lines should be dark
-enough for purposes of reproduction. The figure number and caption
-always appear after the figure. Place one line space before the figure
-caption and one line space after the figure. The figure caption should
-be lower case (except for first word and proper nouns); figures are
-numbered consecutively.
-
-You may use color figures.  However, it is best for the figure
-captions and the paper body to be legible if the paper is printed in
-either black/white or in color.
-\begin{figure}[h]
-  \centering
-  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-  \caption{Sample figure caption.}
-\end{figure}
-
-\subsection{Tables}
-
-All tables must be centered, neat, clean and legible.  The table
-number and title always appear before the table.  See
-Table~\ref{sample-table}.
-
-Place one line space before the table title, one line space after the
-table title, and one line space after the table. The table title must
-be lower case (except for first word and proper nouns); tables are
-numbered consecutively.
-
-Note that publication-quality tables \emph{do not contain vertical
-  rules.} We strongly suggest the use of the \verb+booktabs+ package,
-which allows for typesetting high-quality, professional tables:
-\begin{center}
-  \url{https://www.ctan.org/pkg/booktabs}
-\end{center}
-This package was used to typeset Table~\ref{sample-table}.
-
-\begin{table}[t]
-  \caption{Sample table title}
-  \label{sample-table}
-  \centering
-  \begin{tabular}{lll}
-    \toprule
-    \multicolumn{2}{c}{Part}                   \\
-    \cmidrule{1-2}
-    Name     & Description     & Size ($\mu$m) \\
-    \midrule
-    Dendrite & Input terminal  & $\sim$100     \\
-    Axon     & Output terminal & $\sim$10      \\
-    Soma     & Cell body       & up to $10^6$  \\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-\section{Final instructions}
-
-Do not change any aspects of the formatting parameters in the style
-files.  In particular, do not modify the width or length of the
-rectangle the text should fit into, and do not change font sizes
-(except perhaps in the \textbf{References} section; see below). Please
-note that pages should be numbered.
-
-\section{Preparing PDF files}
-
-Please prepare submission files with paper size ``US Letter,'' and
-not, for example, ``A4.''
-
-Fonts were the main cause of problems in the past years. Your PDF file
-must only contain Type 1 or Embedded TrueType fonts. Here are a few
-instructions to achieve this.
-
-\begin{itemize}
-
-\item You should directly generate PDF files using \verb+pdflatex+.
-
-\item You can check which fonts a PDF files uses.  In Acrobat Reader,
-  select the menu Files$>$Document Properties$>$Fonts and select Show
-  All Fonts. You can also use the program \verb+pdffonts+ which comes
-  with \verb+xpdf+ and is available out-of-the-box on most Linux
-  machines.
-
-\item The IEEE has recommendations for generating PDF files whose
-  fonts are also acceptable for NIPS. Please see
-  \url{http://www.emfield.org/icuwb2010/downloads/IEEE-PDF-SpecV32.pdf}
-
-\item \verb+xfig+ "patterned" shapes are implemented with bitmap
-  fonts.  Use "solid" shapes instead.
-
-\item The \verb+\bbold+ package almost always uses bitmap fonts.  You
-  should use the equivalent AMS Fonts:
-\begin{verbatim}
-   \usepackage{amsfonts}
-\end{verbatim}
-followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or
-\verb+\mathbb{C}+ for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You
-can also use the following workaround for reals, natural and complex:
-\begin{verbatim}
-   \newcommand{\RR}{I\!\!R} %real numbers
-   \newcommand{\Nat}{I\!\!N} %natural numbers
-   \newcommand{\CC}{I\!\!\!\!C} %complex numbers
-\end{verbatim}
-Note that \verb+amsfonts+ is automatically loaded by the
-\verb+amssymb+ package.
-
-\end{itemize}
-
-If your file contains type 3 fonts or non embedded TrueType fonts, we
-will ask you to fix it.
-
-\subsection{Margins in \LaTeX{}}
-
-Most of the margin problems come from figures positioned by hand using
-\verb+\special+ or other commands. We suggest using the command
-\verb+\includegraphics+ from the \verb+graphicx+ package. Always
-specify the figure width as a multiple of the line width as in the
-example below:
-\begin{verbatim}
-   \usepackage[pdftex]{graphicx} ...
-   \includegraphics[width=0.8\linewidth]{myfile.pdf}
-\end{verbatim}
-See Section 4.4 in the graphics bundle documentation
-(\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})
-
-A number of width problems arise when \LaTeX{} cannot properly
-hyphenate a line. Please give LaTeX hyphenation hints using the
-\verb+\-+ command when necessary.
-
-\subsubsection*{Acknowledgments}
-
-Use unnumbered third level headings for the acknowledgments. All
-acknowledgments go at the end of the paper. Do not include
-acknowledgments in the anonymized submission, only in the final paper.
+%\subsubsection*{Acknowledgments}
 

### 2017-05-17 articles/nips-2017/ms.tex
some math

@@
 \usepackage{microtype}      % microtypography
+
+\usepackage{amssymb} % for \intercal
+\usepackage{bm}      % for bold vectors
+\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
+\def\veccov{\vect{C}}
+\def\vecmean{\vect{\mu}}
+\def\vectheta{\vect{\theta}}
+\def\weight{w}
+\def\weights{\vect{\weight}}
+\def\datum{y}
+\def\data{\vect{\datum}}
+\def\likelihood{\mathcal{L}}
+\def\current{^{\mathrm{(K)}}}
+\def\future{^{\mathrm{(K} + \Delta\mathrm{K)}}}
 
@@
 \section{The objective function}
+\label{sec:objective-function}
+
+\noindent{}The probability density function $f$ for a multivariate normal 
+distribution with $D$ dimensions is,
+
+\begin{equation}
+  f(\data|\vecmean,\veccov) 
+      = \frac{1}{\sqrt{(2\pi)^d|\veccov|}}
+        \exp{\left[-\frac{1}{2}(\data - \vecmean)^\intercal\veccov^{-1}(\data - \vecmean)\right]}
+\end{equation}
+
+\noindent{}where $\vecmean$ and $\veccov$ is the multivariate mean and 
+covariance matrix, and $\data$ are the data.
+For a fixed number of $K$ multivariate Gaussian mixtures, the probability
+density function is given by:
+
+\begin{equation}
+  f(\data|\vectheta) = \sum_{k=1}^{K} \weight_k \, f(\data|\vectheta_k)
+\end{equation}
+
+\noindent{}Where $\vectheta_k \equiv \{\vecmean_k, \veccov_k\}$,
+$\vectheta \equiv \{\vectheta_1,\dots,\vectheta_K,\weight_1,\dots,\weight_K\}$, 
+and $\weight_k$ is the relative weight of the $k$-th mixture such that
+$\sum_{k=1}^{K}\weight_k = 1$.
+
+\subsection{Minimum Message Length}
+\label{sec:mml}
+
+The Minimum Message Length \citep[MML;][]{Wallace_1968} principle is a formal
+description of Occam's razor that is based on information theory.
+% Taking Bayes and Shannon.
+% TODO
+The MML principle states that the best explanation of the data is the model
+that describes the data using the smallest amount of information (shortest
+message), where information is described as per \citet{Shannon_1948}.
+
+Calculating the optimal lossless message length of a complex model can be
+non-trivial or intractable. Here we make use of the \citet{Wallace_1987}
+approximation of the message length:
+
+\begin{equation}
+\label{eq:message-length}
+  I(\vectheta,\data) 
+    = \frac{Q}{2}\log\kappa{\left(Q\right)} 
+    - \log{\left(\frac{p(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
+    - \likelihood\left(\data|\vectheta\right) 
+    + \frac{Q}{2} \quad .
+\end{equation}
+
+Here $p(\vectheta)$ is the prior, $Q$ is the number of free parameters in the
+model, $\kappa\left(Q\right)$ is a function to approximate the lattice
+quantisation constant for $Q$ free parameters \citep[e.g.,][]{Conway_1984}, 
+and $|\mathcal{F}\left(\vectheta\right)|$ is the determinant of the 
+\emph{expected} Fisher information matrix (the second-order partial 
+derivatives of the negative log-likelihood function 
+$-\likelihood(\data|\vectheta)$).
+
+The MML principle requires that we encode everything required to reconstruct
+the message, including our prior beliefs on the model parameters.The message 
+for a finite Gaussian mixture must therefore encode:
+\begin{enumerate}
+  \item The number of Gaussian mixtures, $K$.
+  \item The relative weights $\weights$ of the $K - 1$ Gaussian mixtures. Only
+        $K - 1$ must be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
+  \item The component parameters $\vecmean$ and $\veccov$ for all $K$ Gaussian
+        mixtures.
+  \item The data, given the model parameters.
+  \item The lattice quantisation constant $\kappa(Q)$ for the number of model
+        parameters $Q$.
+\end{enumerate}
+
+We provide the message length of each component in turn before outlining our
+objective function.  First, we assume a prior on $K$ of $p(K) \propto 2^{-K}$
+such that the length of the optimal lossless message to encode $K$ is 
+$I(K) = -\log{p(K)} = K\log{2} + \textrm{constant}$.  Given a uniform prior
+on $\weights$ and the requirement that $\sum_{k=1}^{K}\weight_k = 1$, the 
+weights can be treated as parameters of a multinomial distribution, which can
+be optimally encoded in a message of length:
+\begin{equation}
+  I(\weights) 
+    = \frac{K - 1}{2}\log{N} 
+    - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k 
+    - \log{\Gamma{\left(K\right)}} \quad .
+\end{equation}
+
+In order to properly encode the mixture parameters $\vecmean$ and $\veccov$
+for all $K$ mixtures, we must encode both our prior belief on those parameters,
+and the determinant of the expected Fisher information matrix. For the $k$-th
+mixture this becomes,
+
+\begin{equation}
+  I(\vecmean_k,\veccov_k) = -\log{\left(\frac{p(\vecmean_k,\veccov_k)}{\sqrt{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}}\right)}
+                          = -\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
+\end{equation}
+
+\noindent{}and for all mixtures:
+
+\begin{equation}
+  I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|} \quad .
+  \label{eq:I_component_params}
+\end{equation}
+
+We adopt an improper uniform prior of $\mathcal{U}(\vecmean) = [-\infty, +\infty]$ 
+on $\vecmean$, and a conjugate inverted Wishart prior for the covariance matrix
+of individual mixtures \citep[e.g., Section 5.2.3. of ][]{Schafer_1997}:
+\begin{equation}
+  p(\vecmean_k, \veccov_k) \propto |\veccov_k|^{\frac{D+1}{2}} \quad .
+  \label{eq:covariance-prior}
+\end{equation}
+
+We approximate the determinant of the Fisher information 
+$|\mathcal{F}(\vecmean_k, \veccov_k)|$ as the product of 
+$|\mathcal{F}\left(\vecmean_k\right)|$ and $|\mathcal{F}\left(\veccov_k\right)|$ 
+\citep{Oliver_1996,Roberts_1998}.  Taking the second derivative of the
+log-likelihood function $\likelihood\left(\data|\vecmean,\veccov\right)$,
+with respect to $\vecmean$ yields,
+\begin{equation}
+  |\mathcal{F}\left(\vecmean_k\right)| = (N\weight_k)^{D}|\veccov_k|^{-1} \quad .
+\end{equation}
+
+We make use of an analytical expression for the determinant of the Fisher
+information for a covariance matrix $|\mathcal{F}(\veccov_k)|$ \citep{Dwyer_1967,Magnus_1988,Kasarapu_2015},
+\begin{equation}
+  |\mathcal{F}\left(\veccov_k\right)| = (N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \quad ,
+\end{equation}
+
+\noindent{}giving the following approximation for the determinant of the
+expected Fisher information matrix,
+\begin{eqnarray}
+  |\mathcal{F}(\vecmean_k,\veccov_k)| & \approx & |\mathcal{F}(\vecmean_k)|\cdot|\mathcal{F}(\veccov_k)| \nonumber \\
+  |\mathcal{F}(\vecmean_k,\veccov_k)| & \approx & (N\weight_k)^{D}|\veccov_k|^{-1}(N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
+  |\mathcal{F}(\vecmean_k,\veccov_k)| & \approx & (N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
+\end{eqnarray}
+
+\noindent{}which, with Eq. \ref{eq:covariance-prior}, we can substitute into Eq. \ref{eq:I_component_params}:
+
+\begin{eqnarray}
+I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{|\veccov_k|^{\frac{D + 1}{2}}}
+ + \frac{1}{2}\sum_{k=1}^{K}\log{\left[(N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right]} \nonumber \\
+I(\vecmean,\veccov) &=& -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
+ + \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{\left(N\weight_k\right)} -\frac{KD}{2}\log{2} -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
+%I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{\left(N\weight_k\right)} -\frac{KD}{2}\log{2} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
+% +-\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
+% TODO: UPDATE....
+\end{eqnarray}
+
+If we assume that the data have homoskedastic noise properties, then the 
+precision of each measurement $\mathcal{E}$ relates the probability of a
+datum $\textrm{Pr}(\datum_n)$ can be related to the probability density of
+the datum, given the model 
+$\textrm{Pr}(\datum_n) = \mathcal{E}^{D}\textrm{Pr}(y_i|\mathcal{M})$.
+In the paper we adopt $\mathcal{E} = 0.01$, but we note that the value of
+$\mathcal{E}$ has no effect on our inferences.  The message length of encoding
+a datum given the model is,
+\begin{equation}
+  I(\data_n) = -\log{\textrm{Pr}(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
+\end{equation}
+
+\noindent{}and for the entire data:
+\begin{eqnarray}
+  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .
+\end{eqnarray}
+
+The final part of the message is to encode the lattice quantisation constant,
+which arises from the approximation to the (so-called) strict MML, where
+parameters are quantised into intervals in high dimensional space in order to
+be losslessly encoded.  We use an approximation of the lattice constant
+\citep[see Sections 5.1.12 and 3.3.4 of ][]{Wallace_2005} such that,
+\begin{equation}
+  \log\kappa(Q) = \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1 \quad ,
+\end{equation}
+
+\noindent{}where $Q$ is the total number of free parameters in the model. Each
+component requires $D$ parameters to specify $\vecmean$, and a covariance
+matrix with off-diagonal terms requires $\frac{1}{2}D(D+1)$ parameters.
+Because only $K - 1$ weights need to be encoded, the total number of free
+parameters is given by ${Q = K\left[\frac{D(D+3)}{2} + 1\right] - 1}$.
+
+\subsection{Message length for a finite mixture of multivariate Gaussian components}
+
+The message lengths of the optimally encoded components allows us to construct
+our full objective function:
+\begin{equation}
+  I(\vectheta,\data) = I(K) + I(\weights) + \sum_{k=1}^{K}I(\vectheta_k) + I(\data|\vectheta) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right)\quad .
+\end{equation}
+
+\noindent{}Substituting the expressions outlined in the previous subsection
+(except for $Q$, which we leave unexpanded for brevity), we arrive at the full
+message length for a finite mixture of $K$ multivariate Gaussian components:
+\begin{eqnarray}
+  I(\vectheta,\data) &=&
+      K\log{2} % I(K)
+    + \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k - \log{\Gamma(K)} % I(\weights) 
+    + \frac{1}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\ % 
+   &+&\frac{K}{2}\left[\frac{D(D+3)}{2}\log{N} - D\log{2}\right] % I(\vectheta_k) = I(\vecmean_k,\veccov_k)
+    - ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \nonumber \\ % I(\data|\vectheta)
+   &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) \quad . % lattice
+    \label{eq:objective-function}
+\end{eqnarray}
+
+\subsection{Predicting the message length of uncomputed mixtures}
+
+Consider that the number of true components $K_{true}$ is unknown, and that we
+have an optimal mixture of $K$ components in the sense that the model
+parameters $\vectheta$ minimise the total message length in Eq.
+\ref{eq:objective-function}.  We are interested to know whether the total
+message length will decrease if we add or remove components.  For
+example, our current state may be an optimized mixture with $K=2$ and we want
+to know if an optimal mixture with $K=3$ will result in a shorter message
+length.  By denoting $I\current$ as the message length of the current (optimal) 
+mixture and $I\future$ as mixtures that have yet to be computed, we
+want to know if $I\future - I\current < 0$, or rather we seek to calculate 
+some probability $\textrm{Pr}(I\future - I\current < 0)$ before having to 
+optimize the future mixture $I\future$.  
+Taking Eq. \ref{eq:objective-function}, we can derive an expression for the 
+difference in message lengths between future mixtures and the current state:
+\begin{eqnarray}
+  I\future - I\current &=& \Delta{K}\left[
+  \left(1 - \frac{D}{2}\right)\log{2} + \frac{1}{4}\left(D(D+3)+2\right)\log\frac{N}{2\pi}\right] \nonumber \\
+  &-& \log\Gamma(K+\Delta{K}) + \log\Gamma(K) + \frac{1}{2}\left(\log{Q\future} - \log{Q\current}\right) \nonumber \\
+  &+& \frac{1}{2}\left(\frac{D(D+3)}{2} - 1\right)\left(\sum_{k=1}^{K+\Delta{K}}\log\weight_k\future - \sum_{k=1}^{K}\weight_k\current\right) \nonumber \\
+  &-& \sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future + \sum_{n=1}^{N}\log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\current \nonumber \\
+  &+& \frac{D + 2}{2}\left(\sum_{k=1}^{K}\log|\veccov_k|\current - \sum_{k=1}^{K + \Delta{K}}\log|\veccov_k|\future\right) \quad .
+\end{eqnarray}
 

### 2017-05-17 articles/nips-2017/ms.tex
spell out the problematic terms

@@
   &+& \frac{D + 2}{2}\left(\sum_{k=1}^{K}\log|\veccov_k|\current - \sum_{k=1}^{K + \Delta{K}}\log|\veccov_k|\future\right) \quad .
+  \label{eq:delta_I}
 \end{eqnarray}
+
+Although there are many terms in Eq. \ref{eq:delta_I}, most are readily
+available and can be immediately computed for given trials of $\Delta{K}$.  
+Only three terms are exceptions, which expectedly relate to the parameter 
+estimates of future, uncomputed mixtures:
+\begin{enumerate}
+  \item The sum of the log of the weights of future mixtures,\\
+        $\sum_{k=1}^{K+\Delta{K}}\log\weight_k\future$.
+  \item The negative log-likelihood of future mixtures,\\ $-\sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future$.
+  \item The negative sum of the log of the determinant of covariance matrices
+        of future mixtures,\\
+        $-\sum_{k=1}^{K + \Delta{K}}\log|\veccov_k|\future$.
+\end{enumerate}
 

### 2017-05-17 articles/nips-2017/ms.tex
god that's shit

@@
         $\sum_{k=1}^{K+\Delta{K}}\log\weight_k\future$.
-  \item The negative log-likelihood of future mixtures,\\ $-\sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future$.
+  \item The negative log-likelihood of future mixtures,\\
+        $-\sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future$.
   \item The negative sum of the log of the determinant of covariance matrices
@@
 
+The mixing weights follow a multinomial distribution.  For this reason,
+although the weights of future mixtures are not known, it can be shown that
+the sum of the log of $K$ weights follow the function,
+\begin{equation}
+  \sum_{k=1}^{K+\Delta{K}}\log\weight_k\future \approx -K\log{K} + g_0
+  \label{eq:g}
+\end{equation}
+
+\noindent{}such that if the integration constant $g_0$ is known, Eq. \ref{eq:g}
+provides an upper bound on the possible contribution to the message length
+(maximum entropy).  In practice we store the values of $K$ and 
+$\sum_{k=1}^{K}\log\weight_k$ for optimized mixtures, and update our estimate
+of the integration constant when predicting the message length of future
+mixtures.  Depending on the distribution of the data, the sum of the log of
+the weights can have a value lower than $-K\log{K} + g_0$, but the same
+functional form will be followed.  For this reason, we fit the approximating
+function,
+\begin{equation}
+G(K) = -Kg_1\log{K} + g_0 \approx \sum_{k=1}^{K+\Delta{K}}\log\weight_k 
+\end{equation}
+
+\noindent{}using the stored values of $\{K,\sum_{k=1}^{K}\log\weight_k\}$ from
+previously optimized mixtures.  These two linear fitting routines provide an
+estimate of the expectation, the variance, \emph{and} an upper bound on the
+sum of the log of the weights of unobserved mixtures.
+
+There is no equivalent function we can derive for the log-likelihood of future
+mixtures, or the sum of the log of the determinants.  
+% Li and Barron
+The change in the log-likelihood and the determinant of the covariance
+matrices will depend on the data.
+% Li and Barron
+One could suppose that the determinants of the covariance matrices will on
+average decrease with additional components, but to our knowledge there is 
+no theoretical basis for the functional form.
+Lacking a grounded theoretical basis for these two functions, we collate them
+into a single term and approximate it with a second-order polynomial function,
+$H(K)$,
+\begin{equation}
+H(K) = h_2K^2 + h_1K + h_0 \approx -\sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future - \frac{D+2}{2}\sum_{k=1}^{K + \Delta{K}}\log|\veccov_k|\future \quad .
+\end{equation}
+
+% FIt and determine a variance.
+
+Given the functions $G(K)$ and $H(K)$, we update our estimates of the 
+coefficients $\vec{g}$ and $\vec{h}$ after any proposed mixture has been
+optimized (Section \ref{sec:search-strategy}).  While $G(K)$ is an analytic
+function, $H(K)$ is merely adopted for simplicity.  A more grounded
+approximation is welcomed.  However, in Section \ref{sec:experimental-results}
+we show that our $H(K)$ approximation is sufficient, even with a simplistic
+search strategy.
+
 \section{Search strategy}
+\label{sec:search-strategy}
 

### 2017-05-17 articles/nips-2017/ms.tex
bah

@@
 \section{Introduction}
+\label{sec:introduction}
+
 % Gaussian mixture modelling is a common problem.
@@
 \noindent{}where $\vecmean$ and $\veccov$ is the multivariate mean and 
-covariance matrix, and $\data$ are the data.
+covariance matrix, and the data $\data$ is an $N\times{}D$ array.
 For a fixed number of $K$ multivariate Gaussian mixtures, the probability
@@
 \begin{equation}
-  \sum_{k=1}^{K+\Delta{K}}\log\weight_k\future \approx -K\log{K} + g_0
+  \sum_{k=1}^{K+\Delta{K}}\log\weight_k\future \leq -K\log{K} + g_0
   \label{eq:g}
@@
 
-% Initialization
+\subsection{Initialization}
+
+\subsection{Mixture optimization}
+
+Every time a mixture is perturbed and optimized, we store the
+resulting mixture predictors $\{K, \sum_{k=1}^{K}\log\weight_k, X\}$ in order
+to update our approximating functions $G(K)$ and $H(K)$.
+
+\subsection{Perturbation search}
+
+Starting from our initial mixture, we adopt a perturbation search strategy 
+which considers the following possible perturbations: merge, delete, split,
+and jump.  We briefly describe each step in turn.
+
+% DESCRIBE EACH ONE.
+
+While perturbation search strategies have existed for Gaussian
+mixture models for some time, none have included a \emph{jump} operation.
+With our approximating functions to calculate probability densities for the
+message length of future mixtures, including a \emph{jump} operation allows
+us to quickly move to the appropriate number of mixture components.
 
@@
 
-%
+% The K=4 mixture in F & J 2002, etc.
 
@@
 
-\section{Conclusion}
+% A huge model.
+
+% Cost of the model.
+
+\section{Discussion}
 

### 2017-09-12 sandbox_mixture_slf.py
boom

@@ -2,10 +2,9 @@ import numpy as np
 from snob import mixture_slf as slf
 
 
-#y = np.loadtxt("cluster_abundances.txt")
 
-n_samples, n_features, n_clusters, rank = 1000, 50, 3, 1
-sigma = 1
+n_samples, n_features, n_clusters, rank = 1000, 50, 6, 1
+sigma = 0.5
 true_homo_specific_variances = sigma**2 * np.ones((1, n_features))
 
 
@@ -20,9 +19,13 @@ X = np.dot(true_factor_scores, true_factor_loads)
 
 # Assign objects to different clusters.
 indices = rng.randint(0, n_clusters, size=n_samples)
+true_weights = np.zeros(n_clusters)
 true_means = rng.randn(n_clusters, n_features)
 for index in range(n_clusters):
     X[indices==index] += true_means[index]
+    true_weights[index] = (indices==index).sum()
+
+true_weights = true_weights/n_samples
 
 # Adding homoscedastic noise
 bar = rng.randn(n_samples, n_features)
@@ -33,18 +36,11 @@ sigmas = sigma * rng.rand(n_features) + sigma / 2.
 X_hetero = X + rng.randn(n_samples, n_features) * sigmas
 true_hetero_specific_variances = sigmas**2
 
-#y = np.atleast_2d(np.loadtxt("coffee_example.txt")).T
+data = X_hetero
 
-#true_means = np.zeros((1, n_features))
+model = slf.SLFGMM(n_clusters)
+model.fit(data)
 
-model = slf.SLFGaussianMixture(n_clusters)
-#model.true_factor_scores = true_factor_scores
-#model.true_factor_loads = true_factor_loads
-#model.true_means = true_means
-#model.true_specific_variances = true_specific_variances
-model.fit(X_hetero)
-
-#model.initialize_parameters(X_homo)
 
 def scatter_common(x, y, title=None):
     fig, ax = plt.subplots()
@@ -60,14 +56,15 @@ def scatter_common(x, y, title=None):
 
 scatter_common(true_factor_loads, model.factor_loads, "factor loads")
 scatter_common(true_factor_scores, model.factor_scores, "factor scores")
-scatter_common(true_hetero_specific_variances, model.specific_variances, "specific variances")
+scatter_common(true_homo_specific_variances, model.specific_variances, "specific variances")
 
 # means
 # This one is tricky because the indices are not necessarily the same.
[…]

### 2017-09-12 snob/__init__.py

@@ -8,7 +8,7 @@ from . import (estimator, gaussian, slf)
 __version__ = "0.0.1"
 
 logger = logging.getLogger(__name__)
-logger.setLevel(logging.DEBUG)
+logger.setLevel(logging.WARN)
 
 handler = logging.StreamHandler()
 handler.setFormatter(logging.Formatter(

### 2017-09-12 snob/mixture_slf.py

@@ -4,17 +4,17 @@ An estimator to model data using a fixed number of multivariate Gaussian
 distributions with a single multivariate latent factor.
 """
 
-__all__ = ["SLFGaussianMixture"]
+__all__ = ["SLFGMM"]
 
 import logging
 import numpy as np
 import scipy
-from sklearn import cluster
+from sklearn import (cluster, decomposition)
 
 logger = logging.getLogger(__name__)
 
 
-class SLFGaussianMixture(object):
+class SLFGMM(object):
 
     r"""
     Model data from a fixed number of multivariate Gaussian distributions with
@@ -24,7 +24,7 @@ class SLFGaussianMixture(object):
     expectation-maximization, with minimum message length describing the cost
     function.
 
-    :param num_mixtures:
+    :param num_components:
         The number of multivariate Gaussian mixtures to model.
 
     :param threshold: [optional]
@@ -35,10 +35,6 @@ class SLFGaussianMixture(object):
         The maximum number of iterations to run per expectation-maximization
         loop (default: ``10000``).
 
-    :param max_inner_iterations: [optional]
-        The maximum number of iterations to run when updating the latent
-        factor in the maximization step (default: ``1000``).
-
     :param initialization_method: [optional]
         The method to use to initialize the mixture parameters. Available
         options are: ``kmeans``, and ``random`` (default: ``kmeans``).
@@ -47,11 +43,11 @@ class SLFGaussianMixture(object):
     parameter_names = ("factor_loads", "factor_scores", "specific_variances",
         "means", "weights")
 
-    def __init__(self, num_mixtures, threshold=1e-5, max_em_iterations=10000,
-        max_inner_iterations=1000, initialization_method="kmeans", **kwargs):
+    def __init__(self, num_components, threshold=1e-5, max_em_iterations=10000,
+        initialization_method="kmeans++", **kwargs):
 
-        num_mixtures = int(num_mixtures)
-        if 1 > num_mixtures:
+        num_components = int(num_components)
+        if 1 > num_components:
             raise ValueError("number of mixtures must be a positive integer")
 
         if 0 >= threshold:
@@ -60,17 +56,16 @@ class SLFGaussianMixture(object):
         if 0 >= max_em_iterations:
             raise ValueError("max_em_iterations must be a positive integer")
 
-        available = ("kmeans", "random")
+        available = ("kmeans++", "random")
         initialization_method = initialization_method.strip().lower()
[…]

### 2017-09-12 snob/nips_search2.py

@@ -1,0 +1,1141 @@
+
+"""
+An estimator for modelling data from a mixture of Gaussians, 
+using an objective function based on minimum message length.
+"""
+
+__all__ = [
+    "GaussianMixture", 
+    "kullback_leibler_for_multivariate_normals",
+    "responsibility_matrix",
+    "split_component", "merge_component", "delete_component", 
+] 
+import logging
+import numpy as np
+import scipy
+from collections import defaultdict
+
+logger = logging.getLogger(__name__)
+
+
+class GaussianMixture(object):
+
+    r"""
+    Model data from (potentially) many multivariate Gaussian distributions, 
+    using minimum message length (MML) as the objective function.
+
+    :param covariance_type: [optional]
+        The structure of the covariance matrix for individual components.
+        The available options are: `full` for a free covariance matrix, or
+        `diag` for a diagonal covariance matrix (default: ``diag``).
+
+    :param covariance_regularization: [optional]
+        Regularization strength to add to the diagonal of covariance matrices
+        (default: ``0``).
+
+    :param threshold: [optional]
+        The relative improvement in message length required before stopping an
+        expectation-maximization step (default: ``1e-5``).
+
+    :param max_em_iterations: [optional]
+        The maximum number of iterations to run per expectation-maximization
+        loop (default: ``10000``).
+    """
+
+    parameter_names = ("mean", "covariance", "weight")
+
+    def __init__(self, covariance_type="diag", covariance_regularization=0, 
+        threshold=1e-5, max_em_iterations=10000, **kwargs):
+
+        available = ("full", "diag", )
+        covariance_type = covariance_type.strip().lower()
+        if covariance_type not in available:
+            raise ValueError("covariance type '{}' is invalid. "\
+                             "Must be one of: {}".format(
+                                covariance_type, ", ".join(available)))
+
+        if 0 > covariance_regularization:
+            raise ValueError(
+                "covariance_regularization must be a non-negative float")
+
+        if 0 >= threshold:
+            raise ValueError("threshold must be a positive value")
+
+        if 1 > max_em_iterations:
+            raise ValueError("max_em_iterations must be a positive integer")
+
+        self._threshold = threshold
+        self._max_em_iterations = max_em_iterations
[…]

### 2017-09-12 articles/chemical-tagging-gmm/experiments/slf-clustering-apogee-dr14/prepare.py
moving furniture

@@ -1,0 +1,131 @@
+from astropy.table import Table
+
+output_filename = "apogee-dr14-cannon.fits"
+
+cannon = Table.read("allStarCannon-l31c.2.fits")
+aspcap = Table.read("allStar-l31c.2.fits")
+
+# Some basic checks.
+cannon["APOGEE_ID"] = [each.strip() for each in cannon["APOGEE_ID"]]
+aspcap["APOGEE_ID"] = [each.strip() for each in aspcap["APOGEE_ID"]]
+
+assert len(cannon) == len(aspcap)
+assert all(aspcap["APOGEE_ID"] == cannon["APOGEE_ID"])
+
+
+# Transfer the following labels from ASPCAP table to The Cannon table:
+transfer_columns = (
+    "APSTAR_ID",
+    "TARGET_ID",
+    "ASPCAP_ID",
+    "FILE",
+    "APOGEE_ID",
+    "TELESCOPE",
+    "LOCATION_ID",
+    "FIELD",
+    "J",
+    "J_ERR",
+    "H",
+    "H_ERR",
+    "K",
+    "K_ERR",
+    "RA",
+    "DEC",
+    "GLON",
+    "GLAT",
+    "APOGEE_TARGET1",
+    "APOGEE_TARGET2",
+    "APOGEE_TARGET3",
+    "TARGFLAGS",
+    "SURVEY",
+    "NVISITS",
+    "COMBTYPE",
+    "COMMISS",
+    "SNR",
+    "STARFLAG",
+    "STARFLAGS",
+    "ANDFLAG",
+    "ANDFLAGS",
+    "VHELIO_AVG",
+    "VSCATTER",
+    "VERR",
+    "VERR_MED",
+    "OBSVHELIO_AVG",
+    "OBSVSCATTER",
+    "OBSVERR",
+    "OBSVERR_MED",
+    "SYNTHVHELIO_AVG",
+    "SYNTHVSCATTER",
+    "SYNTHVERR",
+    "SYNTHVERR_MED",
+    "RV_TEFF",
+    "RV_LOGG",
+    "RV_FEH",
+    "RV_ALPHA",
+    "RV_CARB",
+    "RV_CCFWHM",
+    "RV_AUTOFWHM",
+    "SYNTHSCATTER",
+    "STABLERV_CHI2",
+    "STABLERV_RCHI2",
+    "CHI2_THRESHOLD",
[…]

### 2017-09-12 articles/chemical-tagging-gmm/experiments/slf-clustering-apogee-dr14/retrieve.sh

@@ -1,0 +1,2 @@
+wget https://data.sdss.org/sas/dr14/apogee/spectro/redux/r8/stars/l31c/l31c.2/cannon/allStarCannon-l31c.2.fits
+wget https://data.sdss.org/sas/dr14/apogee/spectro/redux/r8/stars/l31c/l31c.2/allStar-l31c.2.fits

### 2017-09-13 articles/chemical-tagging-gmm/experiments/slf-clustering-apogee-dr14/data/prepare.py
moving things

@@ -12,7 +12,6 @@ aspcap["APOGEE_ID"] = [each.strip() for each in aspcap["APOGEE_ID"]]
 assert len(cannon) == len(aspcap)
 assert all(aspcap["APOGEE_ID"] == cannon["APOGEE_ID"])
 
-
 # Transfer the following labels from ASPCAP table to The Cannon table:
 transfer_columns = (
     "APSTAR_ID",
@@ -23,12 +22,6 @@ transfer_columns = (
     "TELESCOPE",
     "LOCATION_ID",
     "FIELD",
-    "J",
-    "J_ERR",
-    "H",
-    "H_ERR",
-    "K",
-    "K_ERR",
     "RA",
     "DEC",
     "GLON",
@@ -58,13 +51,6 @@ transfer_columns = (
     "SYNTHVSCATTER",
     "SYNTHVERR",
     "SYNTHVERR_MED",
-    "RV_TEFF",
-    "RV_LOGG",
-    "RV_FEH",
-    "RV_ALPHA",
-    "RV_CARB",
-    "RV_CCFWHM",
-    "RV_AUTOFWHM",
     "SYNTHSCATTER",
     "STABLERV_CHI2",
     "STABLERV_RCHI2",
@@ -80,41 +66,12 @@ transfer_columns = (
     "ASPCAP_VERSION",
     "RESULTS_VERSION",
     "EXTRATARG",
-    "MIN_H",
-    "MAX_H",
-    "MIN_JK",
-    "MAX_JK",
     "VMICRO",
     "VMACRO",
     "VSINI",
     "ASPCAPFLAG",
     "ASPCAPFLAGS",
     "REDUCTION_ID",
-    "SRC_H",
-    "WASH_M",
-    "WASH_M_ERR",
-    "WASH_T2",
-    "WASH_T2_ERR",
-    "DDO51",
-    "DDO51_ERR",
-    "IRAC_3_6",
-    "IRAC_3_6_ERR",
-    "IRAC_4_5",
-    "IRAC_4_5_ERR",
-    "IRAC_5_8",
-    "IRAC_5_8_ERR",
-    "IRAC_8_0",
-    "IRAC_8_0_ERR",
-    "WISE_4_5",
-    "WISE_4_5_ERR",
-    "TARG_4_5",
-    "TARG_4_5_ERR",
-    "AK_TARG",
-    "AK_TARG_METHOD",
-    "AK_WISE",
[…]

### 2017-09-13 articles/gmm-search/ms.tex
kmeans++ and davids email on search

@@
 %% Copyright 2017 the authors. All rights reserved.
-
-\documentclass{aastex61}
-
-\usepackage{bm}
+\documentclass{elsarticle}
+\journal{Journal of Artificial Intelligence}
+\usepackage{bm,url}
 %\usepackage{boldsymbol}
@@
 
-\received{}
-\revised{}
-\accepted{}
-
-\submitjournal{TBD} % ApJ, MNRAS, PASA?
-
-\shorttitle{A search strategy for Gaussian mixtures using MML}
-\shortauthors{Casey et al.}
-
 \begin{document}
-
+	\begin{frontmatter}
 		\title{A search method for Gaussian mixtures using MML}
@@
 
-\keywords{}
-
+		\begin{keyword}
+			Search\sep mml
+		\end{keyword}
+	\end{frontmatter}
 \section{Introduction} 
@@
 
+\section{The k-means++ algorithm}
+
+The k-means++ algorithm extends the k-means method with an efficient way of choosing centres.
+Lets $D(x)$ denote the shortest distance from a data point $x$ to the closest centre that has already been selected.
+The k-means++ performs the following steps:
+
+\begin{enumerate}
+	\item Selects a centre $c_1$ uniformly at random from the $N$ data points.
+	\item Selects a new centre $c_i$ from the data points with probability $\frac{D(x)^2}{\sum_{x\in X}}D(x)^2$
+	\item Repeat Step 2. until k centres have been selected.
+	\item Proceed as with the standard k-means algorithm.
+\end{enumerate}
+
+k-means++ is O(log k)-Competitive.
+
+\section{Email from David}
+
+Re the case of two (or more) classes with the same mean:
+In 2D, a simple case in point is two classes with the same mean
+such that $sigma_{1, x} = sigma_{2, y} >> sigma_{2, x} = sigma_{1, y}$.
+Andy said that this and higher dimensional stuff can be picked up
+by dropping to 1 dimension along the eigenvector of largest eigenvalue.
+I like that.
+I wonder also about the merits of dropping to 2 dimensions (if we have
+at least 2 attributes), being the eigenvectors of the two largest eigenvalues.
+If it could be done quickly, one could transform to a circle, and then
+some test (e.g., Kolmogorov-Smirnov test) for whether data is uniform
+in radial distribution [0, 2 $\pi$).  If not uniform, then perhaps grounds for
+splitting.
+One could possibly check whether uniform radially directly from the
+original 2D projection without having to project it to a circle (i.e., it
+might be pretty easy to work out what the angle of each point would
+be post-transformation without having to do the transformation).  The
+2D projection is 1D in the sense that we only care about the angle
+around the circle.  And, again, Kolmogorov-Smirnov is a popular test
+for uniformity.
+Re looking at things in 1 dimension (not 2D followed by 1D [0, 2 pi) angle
+as above but rather projecting on eigenvector of largest eigenvalue), we
+discussed moments - e.g., \url{https://en.wikipedia.org/wiki/Normal_distribution} .
+So, 2nd, 4th, 6th and 8th moments of a true Gaussian are 1, 3, 15, 105.
+The heuristic I'm going to propose is independent (i.e., I haven't heard
+of it before), but it seems pretty easy, it mightn't be original, and I can
+conceive that Andy might already be using it.
+Let's suppose we have a 1D component.  We'll shift it (without loss of
+generality, w.l.o.g.) so that its mean is 0 and we'll shrink or expand it
+(again, w.l.o.g.) so that its s.d. ($\sigma$) is 1 (and variance v is 1).
+We consider now splitting it into K components with mean 0,
+mixing weights (or proportions) $p_1, ..., p_i, ..., p_K$ and s.d.s $sigma_i$ (and variances $v_i = (sigma_i)^2$).
+Let's set $K = 2, so p_2 = 1 - p_1$.
+Suppose the moments of our current component are 1, 3 $alpha_4, 15 alpha_6 and 105 alpha_8, ... $.
+We get some equations - if we have the correct number of equations,
+they're simultaneous equations.
+If we get too many equations (by using too many moments), then it
+becomes something of a fitting or regression problem to choose
+$p_1, ..., p_i, ..., p_K$ and s.d.s $sigma_i$, variances $v_i = (sigma_i)^2$.
+Our simultaneous equations are:
+$p_1 v_1       + (1 - p_1) v_2       =    1$
+$p_1 (v_1)^2  + (1 - p_1) (v_2)^2  =    3 alpha_4$
+$p_1 (v_1)^3  + (1 - p_1) (v_2)^3  =   15 alpha_6$
+$p_1 (v_1)^4  + (1 - p_1) (v_2)^4  =  105 alpha_8$
+(and possibly etc., as possibly needed).
+
+These might be messy to solve on the fly    but
+one could pre-process by solving many of these beforehand
+(actually, approximately solving these beforehand, as it's only
+a heuristic, anyway) at the very start of the program while
+loading in the data.
+(I remember that Chris Wallace got me to do this a similar
+pre-processing step when dealing with von Mises angular/circular data.)
+So, for $1000 x 1000 = 10^6$ different values of $alpha_4 and alpha_6$
+or for $100 x 100 x 100 = 10^6$ different values of $alpha_4, alpha_6 and alpha_8$
+one could pre-compute {\em approximate} (rough heuristic)
+1,000,000ish or so `solutions' to the above simultaneous equations.
+Then, when running the program with real actual values of $alpha_4$ and $alpha_6$ (and $alpha_8$),
+one could just grab something from nearby in this pre-computed look-up table
+as putative values for $p_1, v_1, v_2$ (and etc., if need be).
+
+I'm not sure whether you're already doing (something like) this.
+If so, am at least slightly but not totally surprised.
+If you're not doing this, it seems worth considering.
+As stated above, `solutions' only need to be rough.
+And, of course, one could change K from 2 to something bigger.
+
+This all said, there should still be some use of randomness in the search.
+
+I hope that the above is clear.
+I hope it's accurate.
+I hope that it or a simplified or more general variant might be useful.
+
 \begin{thebibliography}{}

### 2017-09-13 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/generate.py
OC generated data experiment

@@ -1,0 +1,100 @@
+
+""" Generate a fake catalog of open cluster stars. """
+
+
+# Assumptions:
+
+# (0) Every cluster has a unique chemical fingerprint.
+
+# (1) There is a underlying, single latent factor common to all abundances.
+
+# (2) The abundances in every cluster star will have some intrinsic variance.
+
+# (3) The abundances will have some uncertainty in their measurements, and
+#     uncertainty is homoscedastic.
+
+# (4) The number of stars in each cluster can be different.
+
+# (5) The cluster abundances are taken from some mean metallicity from the
+#     work of Chen et al. (2003): http://adsabs.harvard.edu/abs/2003AJ....125.1397C
+
+
+import numpy as np
+from astropy.table import Table
+
+np.random.seed(1)
+
+oc_mean_properties = Table.read("Chen-et-al-2003-AJ-125-1397-table1.fits")
+
+
+# Adjustable parameters #
+# --------------------- #
+elements_measured = ("Fe", "Ti", "Ca", "Si")
+
+def draw_number_of_stars_in_cluster(mean_cluster_abundance=None):
+    return np.random.randint(1, 250)
+
+means = np.mean(oc_mean_properties["__Fe_H_"]) \
+      + np.random.normal(0, 0.1, size=len(elements_measured))
+
+specific_sigmas = np.clip(
+    np.abs(np.random.normal(0, 0.03, size=len(elements_measured))),
+    0.01, 0.10)
+
+magnitude_of_factor_load = 0.1
+# --------------------- #
+
+
+data = []
+K = len(elements_measured)
+
+for i, open_cluster in enumerate(oc_mean_properties):
+
+    mean_cluster_abundance = open_cluster["__Fe_H_"]
+
+    # How many stars will we draw from this cluster.
+    N_cluster = draw_number_of_stars_in_cluster(mean_cluster_abundance)
+
+    # Determine the cluster fingerprint.
+    # factor_scores (v_n)
+    factor_scores = np.random.normal(0, 1, size=(N_cluster, 1))
+
+    # factor_loads (a_k)
+    factor_loads = np.random.uniform(
+        -abs(magnitude_of_factor_load), 
+        +abs(magnitude_of_factor_load),
+        size=(1, K))
+
+    # variates (r_{nk})
+    variates = np.random.normal(0, 1, size=(N_cluster, K))
+
+    stellar_abundances = means \
+                       + np.dot(factor_scores, factor_loads) \
[…]

### 2017-09-13 articles/chemical-tagging-gmm/experiments/slf-clustering-apogee-dr14/data/prepare.py
reduce file size of apogee catalog

@@ -74,9 +74,8 @@ transfer_columns = (
     "REDUCTION_ID",
     "PMRA",
     "PMDEC",
-    "PM_SRC",
-    "ALL_VISITS",
-    "VISITS")
+    "PM_SRC"
+)
 
 for column in transfer_columns:
     cannon[column] = aspcap[column]

### 2017-09-13 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/data/generate.py
fix data generation to save everything

@@ -19,10 +19,12 @@
 #     work of Chen et al. (2003): http://adsabs.harvard.edu/abs/2003AJ....125.1397C
 
 
+import pickle
 import numpy as np
 from astropy.table import Table
 
-np.random.seed(1)
+random_seed = 1
+np.random.seed(random_seed)
 
 oc_mean_properties = Table.read("Chen-et-al-2003-AJ-125-1397-table1.fits")
 
@@ -45,8 +47,14 @@ magnitude_of_factor_load = 0.1
 # --------------------- #
 
 
-data = []
+
 K = len(elements_measured)
+
+data = []
+all_variates = []
+all_factor_scores = []
+all_factor_loads = []
+
 
 for i, open_cluster in enumerate(oc_mean_properties):
 
@@ -82,6 +90,10 @@ for i, open_cluster in enumerate(oc_mean_properties):
 
     data.extend(rows)
 
+    all_variates.append(variates)
+    all_factor_loads.append(factor_loads)
+    all_factor_scores.append(factor_scores)
+
 
 # Generate a faux-catalog.
 names = ["star_id", "cluster_id"]
@@ -98,3 +110,17 @@ for column_name in ("star_id", "cluster_id"):
     stars[column_name] = np.array(stars[column_name], dtype=int)
 
 stars.write("catalog.fits", overwrite=True)
+
+
+# Load the data that we may want to compare with later.
+generated_parameters = dict(
+    random_seed=random_seed,
+    means=means,
+    variates=np.vstack(all_variates),
+    factor_loads=np.vstack(all_factor_loads),
+    factor_scores=np.vstack(all_factor_scores),
+    specific_sigmas=specific_sigmas
+)
+
+with open("generated-parameters.pkl", "wb") as fp:
+    pickle.dump(generated_parameters, fp, -1)

### 2017-09-13 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/experiment_v0.py

@@ -1,0 +1,111 @@
+
+"""
+Do chemical tagging on a fake data set of open cluster stars.
+"""
+
+import numpy as np
+import pickle
+from astropy.table import Table
+from sklearn import (decomposition, mixture)
+
+np.random.seed(42)
+
+catalog = Table.read("data/catalog.fits")
+with open("data/generated-parameters.pkl", "rb") as fp:
+    gp = pickle.load(fp)
+
+
+def converged_mixture_model(y, model_class, n_components, K_seq=3, **kwargs):
+
+    # Use BIC as our metric.
+    metric = lambda fitted_model: fitted_model.bic(y)
+
+    n_components = np.atleast_1d(n_components).astype(int)
+
+    converged = False
+    metrics = np.nan * np.ones(n_components.size, dtype=float)
+    
+    for i, K in enumerate(n_components):
+        try:
+            model = model_class(n_components=K, **kwargs)
+            fitted = model.fit(y)
+
+        except ValueError:
+            print("Failed on {} for K = {}".format(model_class, K))
+
+            # Don't run any more trials, as increasing K will certainly fail.
+            break
+
+        else:
+            metrics[i] = metric(fitted)
+
+            if K_seq is not None and K > K_seq - 1 \
+            and np.all(np.diff(metrics[-K_seq:]) > 0):
+                converged = True
+                break
+
+    K_best = 1 + np.argmin(metrics)
+
+    return (K_best, converged, metrics)
+
+
+
+
+
+X = np.vstack([catalog[name] for name in catalog.dtype.names[2:]]).T
+K_clusters_total = len(set(catalog["cluster_id"]))
+
+
+
+# Use a standard Gaussian mixture model with BIC penalisation.
+
+if False:
+
+    M = 10 # number of monte carlo realisations to do at each K
+    results = []
+
+    for K_true in range(1, K_clusters_total):
+
+        for m in range(M):
+            if m > 0 and K_true == K_clusters_total: break
+
+            # Pick which clusters we will include,
+            selected_cluster_ids = np.random.choice(np.arange(K_clusters_total),
+                size=K_true, replace=False)
+
+            match = np.in1d(catalog["cluster_id"], selected_cluster_ids)
+
+            K_trials = np.arange(1, np.max([2 * K_true, 10]))
+            K_best, converged, metrics = converged_mixture_model(X[match],
+                mixture.GaussianMixture, K_trials)
+
[…]

### 2017-09-13 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/data/generate.py
boom

@@ -47,14 +47,17 @@ magnitude_of_factor_load = 0.1
 # --------------------- #
 
 
-
 K = len(elements_measured)
 
-data = []
-all_variates = []
-all_factor_scores = []
-all_factor_loads = []
+# Generate the factor_loads (a_k)
+factor_loads = np.random.uniform(
+    -abs(magnitude_of_factor_load), 
+    +abs(magnitude_of_factor_load),
+    size=(1, K))
 
+data = []
+cluster_variates = []
+cluster_factor_scores = []
 
 for i, open_cluster in enumerate(oc_mean_properties):
 
@@ -66,12 +69,6 @@ for i, open_cluster in enumerate(oc_mean_properties):
     # Determine the cluster fingerprint.
     # factor_scores (v_n)
     factor_scores = np.random.normal(0, 1, size=(N_cluster, 1))
-
-    # factor_loads (a_k)
-    factor_loads = np.random.uniform(
-        -abs(magnitude_of_factor_load), 
-        +abs(magnitude_of_factor_load),
-        size=(1, K))
 
     # variates (r_{nk})
     variates = np.random.normal(0, 1, size=(N_cluster, K))
@@ -90,9 +87,8 @@ for i, open_cluster in enumerate(oc_mean_properties):
 
     data.extend(rows)
 
-    all_variates.append(variates)
-    all_factor_loads.append(factor_loads)
-    all_factor_scores.append(factor_scores)
+    cluster_variates.append(variates)
+    cluster_factor_scores.append(factor_scores)
 
 
 # Generate a faux-catalog.
@@ -116,9 +112,9 @@ stars.write("catalog.fits", overwrite=True)
 generated_parameters = dict(
     random_seed=random_seed,
     means=means,
-    variates=np.vstack(all_variates),
-    factor_loads=np.vstack(all_factor_loads),
-    factor_scores=np.vstack(all_factor_scores),
+    variates=np.vstack(cluster_variates),
+    factor_loads=factor_loads,
+    factor_scores=np.vstack(cluster_factor_scores),
     specific_sigmas=specific_sigmas
 )
 

### 2017-09-13 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/experiment_v0.py

@@ -101,10 +101,16 @@ for K_true in range(10, K_clusters_total):
 
         match = np.in1d(catalog["cluster_id"], selected_cluster_ids)
 
-
         model = decomposition.FactorAnalysis(n_components=1)
-
         model = model.fit(X[match])
+
+        # model.components_ == +gp["factor_loads"]
+        # OR
+        # model.components_ == -gp["factor_loads"]
+
+        # AND
+        # model.means_ == gp["means"]
+
 
         raise a
 

### 2017-09-14 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/data/generate.py
fix catalog and include clustering experiment on slf factor scores

@@ -32,6 +32,7 @@ oc_mean_properties = Table.read("Chen-et-al-2003-AJ-125-1397-table1.fits")
 # Adjustable parameters #
 # --------------------- #
 elements_measured = ("Fe", "Ti", "Ca", "Si")
+elements_measured = "ABCDEFGHIJKLMNO"
 
 def draw_number_of_stars_in_cluster(mean_cluster_abundance=None):
     return np.random.randint(1, 250)
@@ -40,8 +41,10 @@ means = np.mean(oc_mean_properties["__Fe_H_"]) \
       + np.random.normal(0, 0.1, size=len(elements_measured))
 
 specific_sigmas = np.clip(
-    np.abs(np.random.normal(0, 0.03, size=len(elements_measured))),
-    0.01, 0.10)
+    np.abs(np.random.normal(0, 0.01, size=len(elements_measured))),
+    0.005, 0.05)
+
+cluster_dispersion = 0.01
 
 magnitude_of_factor_load = 0.1
 # --------------------- #
@@ -59,6 +62,7 @@ data = []
 cluster_variates = []
 cluster_factor_scores = []
 
+
 for i, open_cluster in enumerate(oc_mean_properties):
 
     mean_cluster_abundance = open_cluster["__Fe_H_"]
@@ -68,7 +72,9 @@ for i, open_cluster in enumerate(oc_mean_properties):
 
     # Determine the cluster fingerprint.
     # factor_scores (v_n)
-    factor_scores = np.random.normal(0, 1, size=(N_cluster, 1))
+    factor_scores = np.random.normal(0, 1) * np.ones((N_cluster, 1))
+    #+ np.random.normal(1, cluster_dispersion, 
+    #  size=(N_cluster, 1))
 
     # variates (r_{nk})
     variates = np.random.normal(0, 1, size=(N_cluster, K))
@@ -78,6 +84,12 @@ for i, open_cluster in enumerate(oc_mean_properties):
                        + specific_sigmas * variates
 
     # TODO: add measurement noise.
+
+    """
+    _ = ax.plot(np.mean(means + np.dot(factor_scores, factor_loads), axis=0),
+     lw=2)
+    ax.plot(stellar_abundances.T, c=_[0].get_color(), zorder=-1, alpha=0.1)
+    """
 
     # Generate rows for the faux catalog.
     rows = np.vstack([
@@ -89,6 +101,7 @@ for i, open_cluster in enumerate(oc_mean_properties):
 
[…]

### 2017-09-14 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/experiment_v0.py

@@ -8,17 +8,22 @@ import pickle
 from astropy.table import Table
 from sklearn import (decomposition, mixture)
 
-np.random.seed(42)
+random_seed = 42
 
 catalog = Table.read("data/catalog.fits")
 with open("data/generated-parameters.pkl", "rb") as fp:
     gp = pickle.load(fp)
 
 
-def converged_mixture_model(y, model_class, n_components, K_seq=3, **kwargs):
+K_start = 20
 
-    # Use BIC as our metric.
-    metric = lambda fitted_model: fitted_model.bic(y)
+def converged_mixture_model(y, model_class, n_components, K_seq=3, 
+    metric_function=None, **kwargs):
+
+    if metric_function is None:
+        # Use BIC as our metric.
+        metric_function = lambda fitted_model: fitted_model.aic(y)
+
 
     n_components = np.atleast_1d(n_components).astype(int)
 
@@ -37,7 +42,7 @@ def converged_mixture_model(y, model_class, n_components, K_seq=3, **kwargs):
             break
 
         else:
-            metrics[i] = metric(fitted)
+            metrics[i] = metric_function(fitted)
 
             if K_seq is not None and K > K_seq - 1 \
             and np.all(np.diff(metrics[-K_seq:]) > 0):
@@ -55,16 +60,19 @@ def converged_mixture_model(y, model_class, n_components, K_seq=3, **kwargs):
 X = np.vstack([catalog[name] for name in catalog.dtype.names[2:]]).T
 K_clusters_total = len(set(catalog["cluster_id"]))
 
-
+K_clusters_total = K_start + 1
 
 # Use a standard Gaussian mixture model with BIC penalisation.
 
-if False:
+if True:
+    print("Standard GMM")
+
+    np.random.seed(random_seed)
 
     M = 10 # number of monte carlo realisations to do at each K
     results = []
 
-    for K_true in range(1, K_clusters_total):
+    for K_true in range(K_start, K_clusters_total):
 
         for m in range(M):
             if m > 0 and K_true == K_clusters_total: break
@@ -87,10 +95,70 @@ if False:
 # Use standard FactorAnalysis
 
 
+def compare_latent_factors(model, expected_parameters, rtol=1e-3, atol=1e-2):
+
+    # Check orientation.
+    signs = np.sign(model.components_/gp["factor_loads"])[0]
+
+    assert len(set(signs)) == 1
+
+
+    """
+
+    orientation = [-1, +1][signs[0] > 0]
+
[…]

### 2017-09-14 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/data/generate.py
do things

@@ -51,12 +51,13 @@ magnitude_of_factor_load = 0.1
 
 
 K = len(elements_measured)
+L = 1 # number of latent factors.
 
 # Generate the factor_loads (a_k)
 factor_loads = np.random.uniform(
     -abs(magnitude_of_factor_load), 
     +abs(magnitude_of_factor_load),
-    size=(1, K))
+    size=(L, K))
 
 data = []
 cluster_variates = []
@@ -72,7 +73,7 @@ for i, open_cluster in enumerate(oc_mean_properties):
 
     # Determine the cluster fingerprint.
     # factor_scores (v_n)
-    factor_scores = np.random.normal(0, 1) * np.ones((N_cluster, 1))
+    factor_scores = np.random.normal(0, 1.) * np.ones((N_cluster, 1))
     #+ np.random.normal(1, cluster_dispersion, 
     #  size=(N_cluster, 1))
 

### 2017-09-14 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/experiment_v0.py

@@ -8,51 +8,20 @@ import pickle
 from astropy.table import Table
 from sklearn import (decomposition, mixture)
 
+import utils
+from snob import mixture_slf as snob
+
 random_seed = 42
+
+K_start = 1
+M = 10 # number of monte carlo realisations to do at each K
+
 
 catalog = Table.read("data/catalog.fits")
 with open("data/generated-parameters.pkl", "rb") as fp:
     gp = pickle.load(fp)
 
-
-K_start = 20
-
-def converged_mixture_model(y, model_class, n_components, K_seq=3, 
-    metric_function=None, **kwargs):
-
-    if metric_function is None:
-        # Use BIC as our metric.
-        metric_function = lambda fitted_model: fitted_model.aic(y)
-
-
-    n_components = np.atleast_1d(n_components).astype(int)
-
-    converged = False
-    metrics = np.nan * np.ones(n_components.size, dtype=float)
     
-    for i, K in enumerate(n_components):
-        try:
-            model = model_class(n_components=K, **kwargs)
-            fitted = model.fit(y)
-
-        except ValueError:
-            print("Failed on {} for K = {}".format(model_class, K))
-
-            # Don't run any more trials, as increasing K will certainly fail.
-            break
-
-        else:
-            metrics[i] = metric_function(fitted)
-
-            if K_seq is not None and K > K_seq - 1 \
-            and np.all(np.diff(metrics[-K_seq:]) > 0):
-                converged = True
-                break
-
-    K_best = 1 + np.argmin(metrics)
-
-    return (K_best, converged, metrics)
-
 
 
 
@@ -60,17 +29,22 @@ def converged_mixture_model(y, model_class, n_components, K_seq=3,
 X = np.vstack([catalog[name] for name in catalog.dtype.names[2:]]).T
 K_clusters_total = len(set(catalog["cluster_id"]))
 
-K_clusters_total = K_start + 1
 
 # Use a standard Gaussian mixture model with BIC penalisation.
 
 if True:
-    print("Standard GMM")
+
+    gmm_kwds = dict()
 
     np.random.seed(random_seed)
 
-    M = 10 # number of monte carlo realisations to do at each K
     results = []
+    slf_results = []
+
+    running_delta_slf = 0
+    running_delta_gmm = 0
+
+    print("Standard GMM with kwds: {}".format(gmm_kwds))
 
     for K_true in range(K_start, K_clusters_total):
 
@@ -82,112 +56,89 @@ if True:
                 size=K_true, replace=False)
 
[…]

### 2017-09-14 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/utils.py

@@ -1,0 +1,111 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+from time import time
+
+
+
+def converged_mixture_model(y, model_class, n_components, K_seq=3, 
+    metric_function=None, break_when_breaking=True, **kwargs):
+    """
+    Run a series of Gaussian mixture models with an increasing number of K
+    components.
+    """
+
+    if metric_function is None:
+        # Use BIC as our metric.
+        metric_function = lambda fitted_model: fitted_model.aic(y)
+
+
+    n_components = np.atleast_1d(n_components).astype(int)
+
+    converged = False
+    metrics = np.nan * np.ones(n_components.size, dtype=float)
+    
+    t_init = time()
+
+    for i, K in enumerate(n_components):
+        try:
+            model = model_class(n_components=K, **kwargs)
+            fitted = model.fit(y)
+
+        except ValueError:
+            print("Failed on {} for K = {}".format(model_class, K))
+
+            # Don't run any more trials, as increasing K will certainly fail.
+            if break_when_breaking: break
+            else: continue
+
+        else:
+            metrics[i] = metric_function(fitted)
+
+
+            #print(i, K, K_seq, K > K_seq - 1, metrics[:i], np.diff(metrics[i - K_seq:i + 1]))
+
+            if K_seq is not None and K > K_seq \
+            and np.all(np.diff(metrics[i - K_seq:i + 1]) > 0):
+                converged = True
+                break
+
+    K_best = 1 + np.nanargmin(metrics)
+
+    meta = dict(time_taken=time() - t_init)
+
+    return (K_best, converged, metrics)
+
+
+
+
+def compare_latent_factors(model, expected_parameters, rtol=1e-3, atol=1e-2):
+
+    # Check orientation.
+    signs = np.sign(model.components_/gp["factor_loads"])[0]
+
+    assert len(set(signs)) == 1
+
+
+    """
+
+    orientation = [-1, +1][signs[0] > 0]
+
+    # TODO: this will fail for multiple latent factors -- vectorize!!
+    assert np.allclose(
+        model.components_[0], orientation * expected_parameters["factor_loads"],
+        rtol=rtol, atol=atol)
+
+    assert np.allclose(model.mean_, expected_parameters["means"],
+        rtol=rtol, atol=atol)
+
+    assert np.allclose(
+        model.transform(X[match]).flatten(),
[…]

### 2017-09-14 snob/mixture_slf.py

@@ -455,9 +455,6 @@ class SLFGMM(object):
                 np.dot(gamma, np.atleast_2d(np.dot(V, gamma).sum(axis=0)).T) + omega))
             B = np.dot(V, np.dot(gamma, Cinv))
 
-
-            raise a
-
             #V = self.weight * np.dot(residual, residual.T) \
             #  / np.sum(self.)
 

### 2017-09-20 articles/gmm-search/ms.tex
David's search

@@
 
-		\author{Aleti, Casey, Dowe, Lattanzio}
+		%\author{Aleti, Casey, Dowe, Lattanzio}
 
@@
 
+\section{MML}
+\
 There are $N$ data points each with $D$ dimensions, which are to be modelled
@@
 
-The full expression for the message length of a
+The full expression for the message length is
 given by,
@@
 
-\section{Email from David}
+\section{The Heuristic}
 
@@
 splitting.
+
 One could possibly check whether uniform radially directly from the
@@
 for uniformity.
+
 Re looking at things in 1 dimension (not 2D followed by 1D [0, 2 pi) angle
@@
 So, 2nd, 4th, 6th and 8th moments of a true Gaussian are 1, 3, 15, 105.
+
 The heuristic I'm going to propose is independent (i.e., I haven't heard
@@
 conceive that Andy might already be using it.
+
 Let's suppose we have a 1D component.  We'll shift it (without loss of
@@
 (again, w.l.o.g.) so that its s.d. ($\sigma$) is 1 (and variance v is 1).
+
 We consider now splitting it into K components with mean 0,
-mixing weights (or proportions) $p_1, ..., p_i, ..., p_K$ and s.d.s $sigma_i$ (and variances $v_i = (sigma_i)^2$).
+mixing weights (or proportions) $p_1, ..., p_i, ..., p_K$ and s.d.s $\sigma_i$ (and variances $v_i = (\sigma_i)^2$).
 Let's set $K = 2, so p_2 = 1 - p_1$.
-Suppose the moments of our current component are 1, 3 $alpha_4, 15 alpha_6 and 105 alpha_8, ... $.
+
+Suppose the moments of our current component are 1, $3 \alpha_4, 15 \alpha_6 and 105 \alpha_8, ... $.
 We get some equations - if we have the correct number of equations,
 they're simultaneous equations.
+
 If we get too many equations (by using too many moments), then it
 becomes something of a fitting or regression problem to choose
-$p_1, ..., p_i, ..., p_K$ and s.d.s $sigma_i$, variances $v_i = (sigma_i)^2$.
+$p_1, ..., p_i, ..., p_K$ and s.d.s $\sigma_i$, variances $v_i = (\sigma_i)^2$.
+
 Our simultaneous equations are:
 $p_1 v_1       + (1 - p_1) v_2       =    1$
-$p_1 (v_1)^2  + (1 - p_1) (v_2)^2  =    3 alpha_4$
-$p_1 (v_1)^3  + (1 - p_1) (v_2)^3  =   15 alpha_6$
-$p_1 (v_1)^4  + (1 - p_1) (v_2)^4  =  105 alpha_8$
+$p_1 (v_1)^2  + (1 - p_1) (v_2)^2  =    3 \alpha_4$
+$p_1 (v_1)^3  + (1 - p_1) (v_2)^3  =   15 \alpha_6$
+$p_1 (v_1)^4  + (1 - p_1) (v_2)^4  =  105 \alpha_8$
 (and possibly etc., as possibly needed).
@@
 loading in the data.
-(I remember that Chris Wallace got me to do this a similar
-pre-processing step when dealing with von Mises angular/circular data.)
-So, for $1000 x 1000 = 10^6$ different values of $alpha_4 and alpha_6$
-or for $100 x 100 x 100 = 10^6$ different values of $alpha_4, alpha_6 and alpha_8$
-one could pre-compute {\em approximate} (rough heuristic)
-1,000,000ish or so `solutions' to the above simultaneous equations.
-Then, when running the program with real actual values of $alpha_4$ and $alpha_6$ (and $alpha_8$),
+
+(I remember that Chris Wallace got me to do this a similar pre-processing step when dealing with von Mises angular/circular data.)
+So, for $1000 x 1000 = 10^6$ different values of $\alpha_4 and \alpha_6$
+or for $100 x 100 x 100 = 10^6$ different values of $\alpha_4, \alpha_6 and \alpha_8$
+one could pre-compute {\em approximate} (rough heuristic) 1,000,000ish or so `solutions' to the above simultaneous equations.
+
+Then, when running the program with real actual values of $\alpha_4$ and $\alpha_6$ (and $\alpha_8$),
 one could just grab something from nearby in this pre-computed look-up table
@@
 
-I'm not sure whether you're already doing (something like) this.
-If so, am at least slightly but not totally surprised.
-If you're not doing this, it seems worth considering.
-As stated above, `solutions' only need to be rough.
 And, of course, one could change K from 2 to something bigger.
 
-This all said, there should still be some use of randomness in the search.
-
-I hope that the above is clear.
-I hope it's accurate.
-I hope that it or a simplified or more general variant might be useful.
+There should still be some use of randomness in the search.
 

### 2017-09-21 snob/slf_mixture/__init__.py
refactoring slf gmms into ML and MML classes

@@ -1,0 +1,3 @@
+
+from .mml import MMLMixtureModel
+from .ml import MLMixtureModel

### 2017-09-21 snob/slf_mixture/base.py

@@ -1,0 +1,265 @@
+
+"""
+A base estimator to model data using a fixed number of multivariate Gaussian 
+distributions with a single multivariate latent factor.
+"""
+
+__all__ = ["BaseMixtureModel"]
+
+import logging
+import numpy as np
+import scipy
+
+logger = logging.getLogger("snob")
+
+
+class BaseMixtureModel(object):
+
+    r"""
+    Model data from a fixed number of multivariate Gaussian distributions with
+    a single (multivariate) latent factor.
+
+    The latent factor and mixture parameters are iteratively updated using
+    expectation-maximization, with minimum message length describing the cost
+    function.
+
+    :param num_components:
+        The number of multivariate Gaussian mixtures to model.
+
+    :param threshold: [optional]
+        The relative improvement in log probability required before stopping
+        an expectation-maximization step (default: ``1e-5``).
+
+    :param max_em_iterations: [optional]
+        The maximum number of iterations to run per expectation-maximization
+        loop (default: ``10000``).
+
+    :param initialization_method: [optional]
+        The method to use to initialize the mixture parameters. Available
+        options are: ``kmeans``, and ``random`` (default: ``kmeans``).
+    """
+
+    parameter_names = ("factor_loads", "factor_scores", "specific_variances",
+        "means", "weights")
+
+    def __init__(self, num_components, threshold=1e-5, max_em_iterations=10000,
+        initialization_method="kmeans++", **kwargs):
+
+        num_components = int(num_components)
+        if 1 > num_components:
+            raise ValueError("number of mixtures must be a positive integer")
+
+        if 0 >= threshold:
+            raise ValueError("threshold must be a positive value")
+
+        if 0 >= max_em_iterations:
+            raise ValueError("max_em_iterations must be a positive integer")
+
+        available = ("kmeans++", "random")
+        initialization_method = initialization_method.strip().lower()
+        if initialization_method not in available:
+            raise ValueError(
+                "Initialization method '{}' is invalid. Must be one of: {}"\
+                .format(initialization_method, ", ".join(available)))
+
+        self.num_components = num_components
+        self.threshold = threshold
+        self.max_em_iterations = max_em_iterations
+        self.initialization_method = initialization_method
+        self.meta = {}
+
+        self.set_parameters(**dict([(p, None) for p in self.parameter_names]))
+
+        return None
+
+
+    @property
+    def means(self):
+        r"""
[…]

### 2017-09-21 snob/slf_mixture/ml.py

@@ -1,0 +1,226 @@
+
+"""
+An estimator to model data using a fixed number of multivariate Gaussian 
+distributions with a single multivariate latent factor, using maximum
+likelihood as the objective function.
+"""
+
+__all__ = ["MLMixtureModel"]
+
+import logging
+import numpy as np
+import scipy
+from sklearn import (cluster, decomposition)
+
+from .base import BaseMixtureModel
+
+logger = logging.getLogger("snob")
+
+
+class MLMixtureModel(BaseMixtureModel):
+    
+
+    def initialize_parameters(self, y, **kwargs):
+
+        N, D = y.shape
+
+        # Initialize the factor load.
+        fa = decomposition.FactorAnalysis(n_components=1, **kwargs)
+        fa.fit(y)
+
+        factor_load = fa.components_
+        specific_variance = fa.noise_variance_
+
+        b = factor_load/np.sqrt(specific_variance)
+        b_sq = np.sum(b**2)
+        factor_score = np.dot(y, b.T) * (1 - D/(N - 1) * b_sq)/(1 + b_sq)
+
+        y_nlf = y - np.dot(factor_score, factor_load)
+
+        if self.initialization_method == "kmeans++":
+
+            K = self.num_components
+            row_norms = cluster.k_means_.row_norms(y_nlf, squared=True)
+            mean = cluster.k_means_._k_init(
+                y_nlf, K, row_norms, 
+                cluster.k_means_.check_random_state(None))
+
+            distance = np.sum((y_nlf[:, :, None] - mean.T)**2, axis=1).T
+
+            responsibility = np.zeros((K, N))
+            responsibility[np.argmin(distance, axis=0), np.arange(N)] = 1.0
+
+            weight = responsibility.sum(axis=1)/N
+
+        else:
+            raise NotImplementedError("random assignment not implemented yet"\
+                                      " for max likelihood SLF mixture model")
+
+        # Set the parameters.
+        return self.set_parameters(means=mean, weights=weight,
+            factor_loads=factor_load, factor_scores=factor_score, 
+            specific_variances=specific_variance)
+
+
+
+    def _aecm_step_1(self, data, responsibility):
+        r"""
+        Perform the first conditional maximization step of the ECM algorithm,
+        where we update the weights and the means.
+
+        :param data:
+            A :math:`N\times{}D` array of the observations :math:`y`,
+            where :math:`N` is the number of observations, and :math:`D` is
+            the number of dimensions per observation.
+
+        :param responsibility: 
[…]

### 2017-09-21 snob/slf_mixture/mml.py

@@ -1,0 +1,212 @@
+
+"""
+An estimator to model data using a fixed number of multivariate Gaussian 
+distributions with a single multivariate latent factor, using minimum message
+length as the objective function.
+"""
+
+__all__ = ["MMLMixtureModel"]
+
+import logging
+import numpy as np
+import scipy
+from sklearn import (cluster, decomposition)
+
+from .base import BaseMixtureModel
+
+logger = logging.getLogger("snob")
+
+
+class MMLMixtureModel(BaseMixtureModel):
+
+
+    def _initialize_parameters(self, data, **kwargs):
+        r"""
+        Initialize the model parameters, given the data.
+
+        :param data:
+            A :math:`N\times{}D` array of the observations :math:`y`,
+            where :math:`N` is the number of observations, and :math:`D` is
+            the number of dimensions per observation.
+        """
+
+        # Estimate the latent factor first.
+        N, K = data.shape
+
+        means = np.mean(data, axis=0)
+        w = data - means
+        correlation_matrix = np.corrcoef(w.T)
+
+        U, S, V = np.linalg.svd(correlation_matrix)
+
+
+        if kwargs.get("randomly_initialize_factor_loads", False):
+            b = np.random.uniform(size=(K, )) - 0.5
+        else:
+
+            # Set the initial values for b as the principal eigenvector, scaled by
+            # the square-root of the principal eigenvalue
+            b = V[0] * S[0]**0.5
+
+
+
+        specific_sigmas = np.sqrt(np.diag(np.dot(w.T, w)) \
+                    / ((b*b + 1.0) * (N - 1)))
+        factor_loads = specific_sigmas * b
+        y = (data - means)/specific_sigmas
+
+        raise a
+
+        b_sq = np.sum(b**2)
+        factor_scores = np.dot(y, b) * (1 - K/(N - 1) * b_sq)/(1 + b_sq)
+        specific_variances = specific_sigmas**2
+        specific_variances = specific_variances.reshape((1, -1))
+
+        # Subtract off the factor loads and factor scores, then estimate the
+        # cluster properties.
+        factor_scores = np.atleast_2d(factor_scores).reshape((-1, 1))
+        factor_loads = np.atleast_2d(factor_loads).reshape((1, -1))
+
+        y_nlf = data - np.dot(factor_scores, factor_loads)
+
+        if self.initialization_method == "kmeans++":
+
+            raise a
+            # Do k-means on the data without the latent factor.
[…]

### 2017-09-21 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/toy_experiment.py
simplified toy model

@@ -1,0 +1,38 @@
+
+"""
+Do chemical tagging on toy data to check the SLF convergence.
+"""
+
+import numpy as np
+import pickle
+from astropy.table import Table
+import matplotlib.pyplot as pltmodel
+
+import utils
+from snob.slf_mixture import MLMixtureModel
+
+random_seed = 42
+
+K_start = 1
+M = 10 # number of monte carlo realisations to do at each K
+
+
+USE_TOY_DATA = True
+
+prefix = "toy_" if USE_TOY_DATA else ""
+catalog = Table.read("data/{}catalog.fits".format(prefix))
+with open("data/{}generated-parameters.pkl".format(prefix), "rb") as fp:
+    gp = pickle.load(fp)
+
+
+
+
+X = np.vstack([catalog[name] for name in catalog.dtype.names[2:]]).T
+K_clusters_total = len(set(catalog["cluster_id"]))
+
+#keep = catalog["cluster_id"] < 10
+#X = X[keep]
+
+model = MLMixtureModel(num_components=2)
+
+model.fit(X)

### 2017-09-21 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/data/toy_generate.py
MML/ML estmiators for SLF toy model

@@ -1,0 +1,140 @@
+
+""" Generate data for a toy model. """
+
+
+# Assumptions:
+
+# (0) Every cluster has a unique chemical fingerprint.
+
+# (1) There is a underlying, single latent factor common to all abundances.
+
+# (2) The abundances in every cluster star will have some intrinsic variance.
+
+# (3) The abundances will have some uncertainty in their measurements, and
+#     uncertainty is homoscedastic.
+
+# (4) The number of stars in each cluster can be different.
+
+# (5) The cluster abundances are taken from some mean metallicity from the
+#     work of Chen et al. (2003): http://adsabs.harvard.edu/abs/2003AJ....125.1397C
+
+
+import pickle
+import numpy as np
+from astropy.table import Table
+
+random_seed = 1
+np.random.seed(random_seed)
+
+oc_mean_properties = Table.read("Chen-et-al-2003-AJ-125-1397-table1.fits")
+
+
+PREFIX = "toy"
+
+# Adjustable parameters #
+# --------------------- #
+elements_measured = ("Fe", "Ti", "Ca",)
+
+def draw_number_of_stars_in_cluster(mean_cluster_abundance=None):
+    return np.random.randint(50, 100)
+
+means = np.mean(oc_mean_properties["__Fe_H_"]) \
+      + np.random.normal(0, 0.1, size=len(elements_measured))
+
+specific_sigmas = np.clip(
+    np.abs(np.random.normal(0, 0.01, size=len(elements_measured))),
+    0.005, 0.05)
+
+cluster_dispersion = 0.01
+
+magnitude_of_factor_load = 0.5
+# --------------------- #
+
+
+K = len(elements_measured)
+L = 1 # number of latent factors.
+
+# Generate the factor_loads (a_k)
+factor_loads = np.random.uniform(
+    -abs(magnitude_of_factor_load), 
+    +abs(magnitude_of_factor_load),
+    size=(L, K))
+
+data = []
+cluster_variates = []
+cluster_factor_scores = []
+
+# Just take two clusters.
+open_clusters = np.random.choice(oc_mean_properties, size=2, replace=False)
+
+
+for i, open_cluster in enumerate(open_clusters):
+
+    mean_cluster_abundance = open_cluster["__Fe_H_"]
+
+    # How many stars will we draw from this cluster.
+    N_cluster = draw_number_of_stars_in_cluster(mean_cluster_abundance)
+
+    # Determine the cluster fingerprint.
[…]

### 2017-09-26 articles/gmm-search/ms.tex
motivation for mml

@@
 \journal{Journal of Artificial Intelligence}
-\usepackage{bm,url}
+\usepackage{bm,url,amsmath}
 %\usepackage{boldsymbol}
@@
 \newcommand{\gaia}{\project{Gaia}}
-
+\newcommand{\Cspace}{$\mathcal{C}$-space }
+\newcommand{\Cspaceno}{$\mathcal{C}$-space}
 % Common terms
@@
 
-		%\author{Aleti, Casey, Dowe, Lattanzio}
+%% Group authors per affiliation:
+\author{Elsevier\fnref{myfootnote}}
+\address{Radarweg 29, Amsterdam}
+\fntext[myfootnote]{Since 1880.}
+
+%% or include affiliations in footnotes:
+\author[mymainaddress,mysecondaryaddress]{Elsevier Inc}
+\ead[url]{www.elsevier.com}
+
+\author[mysecondaryaddress]{Global Customer Service\corref{mycorrespondingauthor}}
+\cortext[mycorrespondingauthor]{Corresponding author}
+\ead{support@elsevier.com}
+
+\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
+\address[mysecondaryaddress]{360 Park Avenue South, New York}
 
@@
 \section{MML}
-\
+A complex model is not an optimal one, unless its complexity is
+justifiable by the added explanatory power. 
+There is an established connection between simplicity and truth, as beautifully 
+elucidated by the Occam's Razor principle. MML is a statistical technique that formalizes and quantifies this principle.
+
+The basic idea is to consider the transmission of a set of data and a model that describes it. 
+Clearly one can make increasingly complex models to obtain better fits -- think of epicycles in the solar system. Suppose we try to best represent N data points. 
+Traditional curve fitting methods 
+would give a polynomial of degree N-1, which is perhaps the most precise, 
+but it is also 
+the most complex, and the one that most models the noise, 
+by spuriously {\it over-fitting non-existing patterns.\/}
+Over-fitting can result in a model that may be far from the truth.
+MML will only consider a more complex model if the encoding of the complex model and the data is more efficient 
+than the less complex model, i.e. it finds the minimum message length needed to transmit the data and the model.   It is this \textit{true} model that MML aims to find.
+
+It differs from PCA in that it does find groups (or {\it classes\/} in the notation of MML).
+For each class  it can provide a ``latent factor'' which is akin to the first PC in a PCA study.
+Let us be a little more quantitative.
+
+The MML principle states that the best explanation of the data is the one that leads to the shortest message~\cite{Wallace05}, 
+where a \textit{message} takes into account both the complexity of the model, and its explanatory power. 
+In essence, MML infers the most concise way of encoding the model, $\theta$, and the data, $x$, given the model, $x|\theta$. 
+A complex model will produce a longer first part of the message than a simple model, 
+since more quantities must be stated. On the other hand, the length of the second part of the 
+message decreases with more accurate and complex models, since less data has to be described fully.
+The \textit{message} must encode two parts: the model, and the data. 
+The encoding of the message is based on Shannon's information theory.
+According to Shannon, the information we gain from an event $e$ is $h(e) = - \log_2 p(e)$, where $p(e)$ is the probability of that event.
+The information content is biggest for the improbable outcomes, and smallest for outcomes that we are almost certain about.
+An outcome that has a probability close to one has close to zero information content, since we don't learn anything new from it, whereas the rarer events have much higher information content.
+
+Consider the simplified illustrative example of rolling an unbiased four-sided die (d4), 
+where each of the four outcomes %$(1,2,3,4,5,6)$
+has equal probability $p_i=0.25$. The information content in each datum is $- \log_2 0.25 = 2$. 
+In total $\sum_{i=1}^4-\log_2 (p_i) = 2 + 2 + 2 + 2 = 8$ bits are required to describe the four outcomes. 
+The model description could be $\{00,01,10,11\}$:~this is the first part of the message.
+%We might choose $000$ to represent 1, $001$ to represent 2, $\ldots$  and $100$ to represent 6.
+Here, we need no delimiter as each datum takes 3 bits.
+
+Now assume that the die is biased, with probabilities
+$p(1)=0.5, p(2)=0.25, p(3)=0.125$, and $p(4)=0.125$. 
+In this case, the number of bits required to represent each alternative is different.
+One bit is required for 1, since $-\log_2 0.5 = 1$, and the codeword could be the single binary value 1.
+Similarly, 2 bits are required for the outcome 2, which can be represented with codeword $01$.
+It follows that possible codewords for the remaining alternatives, all using the bit ``1'' to indicate the end of each datum, are $001$ for 3, $0001$ for 4, $00001$ for 5, and $000001$ for six.
+The length of the model required to describe 
+the outcomes of the biased die would be 
+$\sum_{i=1}^4-\log_2 (p_i) = 1 + 2 + 3 + 3 = 9$ bits, and the codewords 
+would be $\{0,10,110,111\}$.
+The second description is 1 bit longer than the first, 
+but, since the most frequent events have a shorter codeword,
+the second part of the message (the data given the model) is shorter. The overall message length of the second description is shorter than the first, when applied to the biased die.
+
+To illustrate this point let's encode 1000 outcomes of a biased die, 
+where 500 outcomes are 1, 250 are 2, 125 are 3, and 125 are 4.
+With the first model each outcome is encoded with two bits,
+so the length of the second part of the message is $1000\times 2=2000$.
+The total message length is the sum of the model length 
+and the length of the data given the model, or 2008 bits. 
+But if we encode the data with the second model,
+the length of the data given the model is 
+$ 500+(250\times2)+(125\times3)+(125\times3)=1750$.
+The total length is $1750+9=1759$. Thus the second model, 
+although it is more complex (takes more bits to describe), creates a shorter encoded message, 
+and hence is a better description than the shorter model.
+
+Formally, the length of the first part of the message is
+$- \log_2h(\theta)$, and that
+of the second part is
+$- \log_2f(x|\theta)$,
+where $h(\theta)$ is the prior probability of model $\theta$,
+and $f(x|\theta)$ is the likelihood of data $x$ assuming model $\theta$.
+The total message length is the sum of the two parts,
+i.e. $- \log_2h(\theta) - \log_2f(x|\theta)$,
+which is minimised precisely when $h(\theta)f(x|\theta)$ is maximised.
+The multiplication of the prior and the likelihood is proportional to the posterior {\em probability} $g(\theta|x)$.
+Thus the MML estimate $h(\theta)f(x|\theta)$ is equal to the posterior mode.
+This is similar in spirit to the (Bayesian) Maximum A Posteriori (MAP) posterior mode, but with the
+benefit of statistical invariance under re-parameterisation \cite{Dowe07GO}.
+
+\section{MML for clustering with a single latent factor model}
+MML is used to identify groups (also known as ``classes'') of stars with the  same composition, and determine the number of independent dimensions in \Cspace.
+It is important to note that MML can answer these two questions simultaneously.
+Without MML one would use some clustering
+technique to find the groups of similar stars, and then use a dimensionality reduction method  like PCA to find the underlying patterns of nucleosynthesis within these groups.
+MML performs these two tasks simultaneously.
+The latent factors within MML are encoded into the optimisation when finding the groups.
+As a result, MML is able to assess whether adding an extra group
+hinders the fit of the latent factor model.
+In current (non-MML) methods these two tasks are not performed jointly, which may lead to groups that are a perfect fit, but with a biased latent factor model.
+MML will be including the underlying nuclear physics when finding the groups within the data
+(of course, we can instruct it not to do this if we prefer). This is a significant advantage over other techniques that have been investigated so far.
+
+One very substantial difference between MML's latent factors and PCA is the following: if we increase the noise in 
+one of the variables, the relevant PC will move in the direction of the noise, ultimately with increasing eigenvalue. 
+But for single latent factor analysis, the latent factor 
+%load vector 
+won't change, rather the noise %($\sigma$) 
+in the relevant variable will increase. This is a very desirable characteristic, helping to isolate
+the underlying physics from random effects, and is an indication of the robustness of latent factors
+%models towards 
+to errors in measurements.
+
+For each group determined by MML, the latent factor is a vector that best encapsulates the location 
+and distribution of this group in \Cspaceno.
+Mixture modelling enables the partitioning of the data into overlapping groups. 
+For each group, a single latent factor model is created. 
+In MML we  model this by constructing a two-part message, with the first part describing the mixture 
+model (or hypothesis, $\theta$) and the second part representing the encoded data ($x$) given the hypothesis. 
+The assumed single latent factor model is $x_{n k} = \mu_k + \nu_n a_k + \sigma_k r_{n k}$, where $\mu_k$ is the 
+estimated mean for each attribute or dimension $k$ ($k=1, \ldots, K$);~$\vec{a} = (a_1,  ...,  a_K)$ is the factor load vector, giving the 
+contribution of each variable to the underlying common factor, $\vec{\nu} = (\nu_1, ..., \nu_N)$ is the 
+factor score vector, representing how much each data item in turn has of the factor;
+%- by convention, factor scores are normalised to be of length order 1; 
+$\sigma_k$ is the standard deviation in 
+attribute $k$, and the variates $r_{n k}$ are independently and identically distributed  
+in $N(0, 1)$. In essence, the model regards each observation $x_{n k}$ as determined by the 
+mean $\mu_k$, depending on an unobserved variable $\nu_n$ and related to a specific 
+variability with standard deviation $\sigma_k$. The MML \textit{message} contains the following components:
+
+%In the stuff below I have removed much of the maths. It is commented out rather than deleted so we can add it back if we want :-)
+
+\begin{itemize}
+	{\setlength\itemindent{5mm}\item[\textbf{1a)}] A statement of the number of groups, or classes, $C$.}
+	%Each class is assumed to have the same prior probability, equal to $2^{-K}$. The length of this part is equal to the negative logarithm of the prior, i.e. $K$.
+
+	{\setlength\itemindent{5mm}\item[\textbf{1b)}] The relative abundances (or mixing proportions) of the amount(s) of data in each group. 
+		The length of this 
+		%part is equal to the negative log likelihood of the multiplication of the prior probabilities 
+		part is equal (within a small constant) to the negative log of the multiplication of the prior probabilities
+		of each group ($h(\vec{p})$) divided by the square root of the Fisher information ($\sqrt{F(\vec{p_k})}$). 
+		The Fisher information matrix contains the expected second-order partial derivatives of the log-likelihood function, and quantifies how sharply the likelihood function peaks, and as a result how precisely the parameters should be stated.}
+	%In the case of uniform prior probabilities $h(\vec{p}) = (K-1)!$, and the Fisher information $F(\vec{p}) = N^{K-1} / (\prod_{k=1}^{K} p_k)$.
+It follows that the length of this part is equal to $- \log( h(\vec{p}) \times (1 / \sqrt{{\kappa_K}^K F(\vec{p})}) ) = - \log((K-1)!) + (K/2) \log(\kappa_K) + ((K-1)/2) \log(N) - (1/2) \sum_{k=1}^{K} \log(p_k)$, where $\kappa_K$ is a lattice constant (varying between $1/12$ and $1/(2 \pi e)$).
+
+	{\setlength\itemindent{5mm}\item[\textbf{1c)}] The distributional parameters $\mu_k$ and $\sigma_k$ 
+		(one for each attribute or dimension  $k$) and the factor
+		load (and score) vectors for each group, again rounded off as above using the Fisher information.}
+	%The explanation length of this part is $(N-1)\sum_k log \sigma_k + \frac{1}{2}[K log (N \sum_n \nu_n^2 - (\sum_n\nu_n)^2) + (N+K-1) log(1+\sum_k(a_k/\sigma_k)^2)] + \frac{1}{2} [\sum_n \nu_n^2 + \sum_n \sum_k (x_{nk} \mu_k - v_na_k)^2/\sigma_k^2]$.
+This part measures the information required to encode the latent factor model for each class.
+The more classes, the longer the message length is.
+
+	{\setlength\itemindent{5mm}\item[\textbf{2)}] The data given the model, whose cost is the well-known statistical log-likelihood. }
+	%In the case of a normal distribution, this is equal to $-log p_k(x) (1/(\sigma_k \sqrt(2\pi) e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}})$.
+\end{itemize}
+
+\vspace{-1mm}
+
+Parts 1a and 1b encode the parameters from the clustering process, part 1c  
+estimates the complexity of the latent factor model, and part 2 encodes the likelihood 
+%(the data given the hypothesis), 
+which represents the goodness-of-fit. 
+The estimation of the multinomial parameters from part 1b and the distributional parameters 
+from part 1c is done by quantifying the trade-off between stating the parameters imprecisely, 
+so as to have higher prior probability in the relevant uncertainty region and keep the first 
+part of the message short, and the goodness of fit to the data, which is typically better 
+if we state the parameter estimates more precisely. Quantitatively, this is done by 
+minimising the sum of the lengths of the four components (1a, 1b, 1c and 2 above).
+Single latent factor analysis attempts to reduce the data to one dimension. 
+If the factor is very weak,  MML may decide that the cost 
+of describing the factor in part 1c is greater than the saving that would occur 
+in part 2. In such a case MML will 
+posit that there is no factor; i.e. including this factor would be over-fitting 
+the data, taking us away from the optimum model.
+
+Similarly, if MML finds that 
+the data for a specific element do not affect the selection %add anything to the ability to 
+of groups, then it will {\it ignore that element.\/} This could tell us something 
+important - such as that there is a large random component for that element 
+or that there are significant observational 
+errors ruining the usefulness of that measurement. 
+These are significant insights into the data that are, importantly, quantified by the MML method.
+
+\section{Message Length}
 There are $N$ data points each with $D$ dimensions, which are to be modelled
@@
 The k-means++ algorithm extends the k-means method with an efficient way of choosing centres.
-Lets $D(x)$ denote the shortest distance from a data point $x$ to the closest centre that has already been selected.
-The k-means++ performs the following steps:
+Let $D(x)$ denote the shortest distance from a data point $x$ to the closest centre that has already been selected.
+The k-means++ has the following steps:
 
@@
 In 2D, a simple case in point is two classes with the same mean
-such that $sigma_{1, x} = sigma_{2, y} >> sigma_{2, x} = sigma_{1, y}$.
+such that $\sigma_{1, x} = \sigma_{2, y} >> \sigma_{2, x} = \sigma_{1, y}$.
 Andy said that this and higher dimensional stuff can be picked up
 by dropping to 1 dimension along the eigenvector of largest eigenvalue.
-I like that.
-I wonder also about the merits of dropping to 2 dimensions (if we have
-at least 2 attributes), being the eigenvectors of the two largest eigenvalues.
-If it could be done quickly, one could transform to a circle, and then
-some test (e.g., Kolmogorov-Smirnov test) for whether data is uniform
-in radial distribution [0, 2 $\pi$).  If not uniform, then perhaps grounds for
-splitting.
 
-One could possibly check whether uniform radially directly from the
-original 2D projection without having to project it to a circle (i.e., it
-might be pretty easy to work out what the angle of each point would
-be post-transformation without having to do the transformation).  The
-2D projection is 1D in the sense that we only care about the angle
-around the circle.  And, again, Kolmogorov-Smirnov is a popular test
-for uniformity.
+I wonder also about the merits of dropping to 2 dimensions (if we have at least 2 attributes), being the eigenvectors of the two largest eigenvalues.
+
+One could possibly check whether uniform radially directly from the original 2D projection without having to project it to a circle (i.e., it might be pretty easy to work out what the angle of each point would be post-transformation without having to do the transformation).
+The 2D projection is 1D in the sense that we only care about the angle around the circle.
+And then some test (e.g., Kolmogorov-Smirnov test) for whether data is uniform in radial distribution.
+If not uniform, then perhaps grounds for splitting.
 
@@
 discussed moments - e.g., \url{https://en.wikipedia.org/wiki/Normal_distribution} .
-So, 2nd, 4th, 6th and 8th moments of a true Gaussian are 1, 3, 15, 105.
-
-The heuristic I'm going to propose is independent (i.e., I haven't heard
-of it before), but it seems pretty easy, it mightn't be original, and I can
-conceive that Andy might already be using it.
+%So, 2nd, 4th, 6th and 8th moments of a true Gaussian are 1, 3, 15, 105.
 
@@
 We consider now splitting it into K components with mean 0,
-mixing weights (or proportions) $p_1, ..., p_i, ..., p_K$ and s.d.s $\sigma_i$ (and variances $v_i = (\sigma_i)^2$).
-Let's set $K = 2, so p_2 = 1 - p_1$.
+mixing weights (or proportions) $w_1, ..., w_i, ..., w_K$ and s.d.s $\sigma_i$ (and variances $v_i = (\sigma_i)^2$).
+Let's set $K = 2$, so $w_2 = 1 - w_1$.
 
-Suppose the moments of our current component are 1, $3 \alpha_4, 15 \alpha_6 and 105 \alpha_8, ... $.
+Suppose the moments of our current component are 1, $3 \alpha_4, 15 \alpha_6$ and $105 \alpha_8, ... $.
 We get some equations - if we have the correct number of equations,
@@
 becomes something of a fitting or regression problem to choose
-$p_1, ..., p_i, ..., p_K$ and s.d.s $\sigma_i$, variances $v_i = (\sigma_i)^2$.
+$w_1, ..., w_i, ..., w_K$ and s.d.s $\sigma_i$, variances $v_i = (\sigma_i)^2$.
 
 Our simultaneous equations are:
-$p_1 v_1       + (1 - p_1) v_2       =    1$
-$p_1 (v_1)^2  + (1 - p_1) (v_2)^2  =    3 \alpha_4$
-$p_1 (v_1)^3  + (1 - p_1) (v_2)^3  =   15 \alpha_6$
-$p_1 (v_1)^4  + (1 - p_1) (v_2)^4  =  105 \alpha_8$
-(and possibly etc., as possibly needed).
 
-These might be messy to solve on the fly  but
-one could pre-process by solving many of these beforehand
-(actually, approximately solving these beforehand, as it's only
-a heuristic, anyway) at the very start of the program while
-loading in the data.
+\begin{align*}
+w_1 v_1       + (1 - w_1) v_2       &=    1\\
+w_1 (v_1)^2  + (1 - w_1) (v_2)^2  &=    3 \alpha_4\\
+w_1 (v_1)^3  + (1 - w_1) (v_2)^3  &=   15 \alpha_6\\
+w_1 (v_1)^4  + (1 - w_1) (v_2)^4  &=  105 \alpha_8
+\end{align*}
 
-(I remember that Chris Wallace got me to do this a similar pre-processing step when dealing with von Mises angular/circular data.)
-So, for $1000 x 1000 = 10^6$ different values of $\alpha_4 and \alpha_6$
-or for $100 x 100 x 100 = 10^6$ different values of $\alpha_4, \alpha_6 and \alpha_8$
-one could pre-compute {\em approximate} (rough heuristic) 1,000,000ish or so `solutions' to the above simultaneous equations.
+These might be messy to solve on the fly  but one could pre-process by solving many of these beforehand (actually, approximately solving these beforehand, as it's only
+a heuristic, anyway) at the very start of the program while loading in the data.
+
+So, for $1000 \times 1000 = 10^6$ different values of $\alpha_4$ and $\alpha_6$
+or for $100 \times 100 \times 100 = 10^6$ different values of $\alpha_4, \alpha_6$ and $\alpha_8$
+one could pre-compute {\em approximate} 1,000,000ish `solutions' to the above simultaneous equations.
 
@@
 one could just grab something from nearby in this pre-computed look-up table
-as putative values for $p_1, v_1, v_2$ (and etc., if need be).
+as putative values for $w_1, v_1, v_2$ (and etc., if need be).
 
@@
 There should still be some use of randomness in the search.
-
-\begin{thebibliography}{}
-
-% Note: I include the DOI as a comment on the same line after every \bibitem
-%       entry. I do this in case I ever need to extract full bibstems for the
-%       bibliography, which can be done using this curl command and the DOI:
-
-%       curl -LH "Accept: application/x-bibtex" http://dx.doi.org/10.5555/12345678
-
-%       If no DOI is available then I list the ADS identifier.
-\end{thebibliography}
+\bibliographystyle{elsarticle-num}
+\bibliography{mmlbibliography}
 

### 2017-09-26 articles/gmm-search/ms.tex
motivation for mml

@@
 
-One very substantial difference between MML's latent factors and PCA is the following: if we increase the noise in 
-one of the variables, the relevant PC will move in the direction of the noise, ultimately with increasing eigenvalue. 
-But for single latent factor analysis, the latent factor 
-%load vector 
-won't change, rather the noise %($\sigma$) 
-in the relevant variable will increase. This is a very desirable characteristic, helping to isolate
-the underlying physics from random effects, and is an indication of the robustness of latent factors
-%models towards 
-to errors in measurements.
+%One very substantial difference between MML's latent factors and PCA is the following: if we increase the noise in one of the variables, the relevant PC will move in the direction of the noise, ultimately with increasing eigenvalue.
+But for single latent factor analysis, the latent factor won't change, rather the noise %($\sigma$)
+%in the relevant variable will increase.
+This is a very desirable characteristic, helping to isolate the underlying physics from random effects, and is an indication of the robustness of latent factors to errors in measurements.
 
@@
 \begin{itemize}
-	{\setlength\itemindent{5mm}\item[\textbf{1a)}] A statement of the number of groups, or classes, $C$.}
-	%Each class is assumed to have the same prior probability, equal to $2^{-K}$. The length of this part is equal to the negative logarithm of the prior, i.e. $K$.
-
-	{\setlength\itemindent{5mm}\item[\textbf{1b)}] The relative abundances (or mixing proportions) of the amount(s) of data in each group. 
+\item[\textbf{1a)}] A statement of the number of groups, or classes, $C$.
+Each class is assumed to have the same prior probability, equal to $2^{-K}$.
+The length of this part is equal to the negative logarithm of the prior, i.e.
+$K$.
+\item[\textbf{1b)}] The relative abundances (or mixing proportions) of the amount(s) of data in each group. 
 		The length of this 
@@
 		of each group ($h(\vec{p})$) divided by the square root of the Fisher information ($\sqrt{F(\vec{p_k})}$). 
-		The Fisher information matrix contains the expected second-order partial derivatives of the log-likelihood function, and quantifies how sharply the likelihood function peaks, and as a result how precisely the parameters should be stated.}
-	%In the case of uniform prior probabilities $h(\vec{p}) = (K-1)!$, and the Fisher information $F(\vec{p}) = N^{K-1} / (\prod_{k=1}^{K} p_k)$.
+		The Fisher information matrix contains the expected second-order partial derivatives of the log-likelihood function, and quantifies how sharply the likelihood function peaks, and as a result how precisely the parameters should be stated.
+
+	In the case of uniform prior probabilities $h(\vec{p}) = (K-1)!$, and the Fisher information $F(\vec{p}) = N^{K-1} / (\prod_{k=1}^{K} p_k)$.
 It follows that the length of this part is equal to $- \log( h(\vec{p}) \times (1 / \sqrt{{\kappa_K}^K F(\vec{p})}) ) = - \log((K-1)!) + (K/2) \log(\kappa_K) + ((K-1)/2) \log(N) - (1/2) \sum_{k=1}^{K} \log(p_k)$, where $\kappa_K$ is a lattice constant (varying between $1/12$ and $1/(2 \pi e)$).
 
-	{\setlength\itemindent{5mm}\item[\textbf{1c)}] The distributional parameters $\mu_k$ and $\sigma_k$ 
+\item[\textbf{1c)}] The distributional parameters $\mu_k$ and $\sigma_k$ 
 		(one for each attribute or dimension  $k$) and the factor
-		load (and score) vectors for each group, again rounded off as above using the Fisher information.}
-	%The explanation length of this part is $(N-1)\sum_k log \sigma_k + \frac{1}{2}[K log (N \sum_n \nu_n^2 - (\sum_n\nu_n)^2) + (N+K-1) log(1+\sum_k(a_k/\sigma_k)^2)] + \frac{1}{2} [\sum_n \nu_n^2 + \sum_n \sum_k (x_{nk} \mu_k - v_na_k)^2/\sigma_k^2]$.
+		load (and score) vectors for each group, again rounded off as above using the Fisher information.
+
+	The explanation length of this part is $(N-1)\sum_k log \sigma_k + \frac{1}{2}[K log (N \sum_n \nu_n^2 - (\sum_n\nu_n)^2) + (N+K-1) log(1+\sum_k(a_k/\sigma_k)^2)] + \frac{1}{2} [\sum_n \nu_n^2 + \sum_n \sum_k (x_{nk} \mu_k - v_na_k)^2/\sigma_k^2]$.
 This part measures the information required to encode the latent factor model for each class.
@@
 
-	{\setlength\itemindent{5mm}\item[\textbf{2)}] The data given the model, whose cost is the well-known statistical log-likelihood. }
-	%In the case of a normal distribution, this is equal to $-log p_k(x) (1/(\sigma_k \sqrt(2\pi) e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}})$.
+\item[\textbf{2)}] The data given the model, whose cost is the well-known statistical log-likelihood. 
+	In the case of a normal distribution, this is equal to
+
+	$$-\log p_k(x) (1/(\sigma_k \sqrt(2\pi) e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}).$$
 \end{itemize}
@@
 \begin{eqnarray}
-  \log{Q^{(new)}} - \log{Q^{(old)}} &=& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
-                                    &&+ \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} - \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)} \quad .\\
+  \log{Q^{(new)}} &- \log{Q^{(old)}} = \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
+                                    &+ \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} - \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)}.
 \end{eqnarray}
@@
 \begin{eqnarray}
-\Delta{}I_{K+1} - I_K &\approx& \log{2} % \Delta{}I_k
+\Delta{}I_{K+1} &-& I_K \approx \log{2} % \Delta{}I_k
     + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w

### 2017-10-02 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/data/toy_generate.py
debugging convergence problems in SLF+GMM

@@ -23,7 +23,7 @@ import pickle
 import numpy as np
 from astropy.table import Table
 
-random_seed = 1
+random_seed = 100
 np.random.seed(random_seed)
 
 oc_mean_properties = Table.read("Chen-et-al-2003-AJ-125-1397-table1.fits")
@@ -33,7 +33,7 @@ PREFIX = "toy"
 
 # Adjustable parameters #
 # --------------------- #
-elements_measured = ("Fe", "Ti", "Ca",)
+elements_measured = ("Fe", "Ti", )
 
 def draw_number_of_stars_in_cluster(mean_cluster_abundance=None):
     return np.random.randint(50, 100)
@@ -47,7 +47,7 @@ specific_sigmas = np.clip(
 
 cluster_dispersion = 0.01
 
-magnitude_of_factor_load = 0.5
+magnitude_of_factor_load = 1
 # --------------------- #
 
 
@@ -73,7 +73,9 @@ for i, open_cluster in enumerate(open_clusters):
     mean_cluster_abundance = open_cluster["__Fe_H_"]
 
     # How many stars will we draw from this cluster.
-    N_cluster = draw_number_of_stars_in_cluster(mean_cluster_abundance)
+    #N_cluster = draw_number_of_stars_in_cluster(mean_cluster_abundance)
+    N_cluster = {0: 30, 1: 60}[i]
+    
 
     # Determine the cluster fingerprint.
     # factor_scores (v_n)

### 2017-10-02 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/toy_experiment.py

@@ -38,6 +38,7 @@ K_clusters_total = len(set(catalog["cluster_id"]))
 mml_mod = MMLMixtureModel(num_components=2)
 mml_mod.fit(X)
 
+raise a
 
 ml_mod = MLMixtureModel(num_components=2)
 ml_mod.fit(X)

### 2017-10-02 snob/slf_mixture/base.py

@@ -39,8 +39,8 @@ class BaseMixtureModel(object):
         options are: ``kmeans``, and ``random`` (default: ``kmeans``).
     """
 
-    parameter_names = ("factor_loads", "factor_scores", "specific_variances",
-        "means", "weights")
+    parameter_names = ("factor_loads", "cluster_factor_scores",
+        "specific_variances", "means", "weights")
 
     def __init__(self, num_components, threshold=1e-5, max_em_iterations=10000,
         initialization_method="kmeans++", **kwargs):
@@ -106,6 +106,14 @@ class BaseMixtureModel(object):
 
 
     @property
+    def cluster_factor_scores(self):
+        r"""
+        Return the mean factor scores for each :math:`k`-th cluster,
+        :math:`v_k`.
+        """
+        return self._cluster_factor_scores
+
+    @property
     def specific_variances(self):
         r"""
         Return the specific variances, :math:`\sigma_k^2`.
@@ -122,6 +130,8 @@ class BaseMixtureModel(object):
             if parameter_name in self.parameter_names:
                 value = value if value is None else np.atleast_2d(value)
                 setattr(self, "_{}".format(parameter_name), value)
+                logger.debug("set_parameters: {} {}".format(parameter_name, value))
+
             else:
                 raise ValueError("unknown parameter '{}'".format(parameter_name))
 
@@ -152,6 +162,7 @@ class BaseMixtureModel(object):
 
         # Calculate the inital expectation.
         responsibility, log_likelihood = self._expectation(data) 
+
         results = [log_likelihood.sum()]
 
         for iteration in range(self.max_em_iterations):
@@ -163,18 +174,21 @@ class BaseMixtureModel(object):
 
             # Compute the e-step of cycle 2
             responsibility, _ = self._expectation(data)
+            assert _.sum() > log_likelihood.sum()
 
             # Do the M-step of cycle 2
             self._aecm_step_2(data, responsibility)
 
             # Re-compute the expectation,
             responsibility, log_likelihood = self._expectation(data)
+            #assert log_likelihood_.sum() > log_likelihood.sum()
+            #log_likelihood = log_likelihood_
             results.append(log_likelihood.sum())
 
             # Check for convergence.
             change = np.abs((results[-1] - results[-2])/results[-2])
             
-            logger.debug("E-M step {}: {} {}".format(
+            logger.debug("fit: {} {} {}".format(
[…]

### 2017-10-02 snob/slf_mixture/mml.py

@@ -37,6 +37,8 @@ class MMLMixtureModel(BaseMixtureModel):
         # single mean for the entire sample, not the cluster means!
         # TODO REVISIT TERMINOLOGY
         means = np.mean(data, axis=0)
+        means = np.array([-0.25640511, -0.04716052])
+        logger.warn("DANGER WILL ROBINSON")
         w = data - means
 
         U, S, V = np.linalg.svd(np.corrcoef(w.T))
@@ -75,16 +77,25 @@ class MMLMixtureModel(BaseMixtureModel):
         effective_membership = np.sum(responsibility, axis=0)
         weights = (effective_membership + 0.5)/(N + D/2.0)
 
+        # We need to initialise the factor scores to be of K clusters.
+        # TODO: Revisit this whether this should be effective_membership > 1
+        cluster_factor_scores = np.dot(factor_scores.T, responsibility) \
+                              / effective_membership
+        
+        # TODO: Do we need to pass cluster_factor_scores around, or factor_scores?
+        #       Or responsibility matrix (ew)
+
         # Set the parameters.
         return self.set_parameters(means=means, weights=weights,
-            factor_loads=factor_loads, factor_scores=factor_scores, 
+            factor_loads=factor_loads, 
+            cluster_factor_scores=cluster_factor_scores, 
             specific_variances=specific_variances)
 
 
     def _aecm_step_1(self, data, responsibility):
         r"""
         Perform the first conditional maximization step of the ECM algorithm,
-        where we update the weights and the means.
+        where we update the weights and the cluster factor scores.
 
         :param data:
             A :math:`N\times{}D` array of the observations :math:`y`,
@@ -96,33 +107,57 @@ class MMLMixtureModel(BaseMixtureModel):
             partially assigned to each :math:`K` component.
         """
 
-        # Update the estimate of the latent factor, after subtracting the
-        # means from each cluster.
         K, N = responsibility.shape
         N, D = data.shape
 
+        # Update the weights by minimising the message length.
         effective_membership = np.sum(responsibility, axis=1)
+        weights = (effective_membership + 0.5)/(N + K/2.0)
+
+        # Update the cluster factor scores.
+        data_ = (data - self.means)/self.factor_loads
+        #factor_scores = np.sum(responsibility.T * data_, axis=1).reshape((-1, 1))
+        
[…]

### 2017-10-02 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/data/toy_generate.py
ml obviously doesn't work. need corrected iterative scheme by taking mml derivatives [wip]

@@ -23,7 +23,7 @@ import pickle
 import numpy as np
 from astropy.table import Table
 
-random_seed = 100
+random_seed = 123
 np.random.seed(random_seed)
 
 oc_mean_properties = Table.read("Chen-et-al-2003-AJ-125-1397-table1.fits")
@@ -40,6 +40,7 @@ def draw_number_of_stars_in_cluster(mean_cluster_abundance=None):
 
 means = np.mean(oc_mean_properties["__Fe_H_"]) \
       + np.random.normal(0, 0.1, size=len(elements_measured))
+means = np.zeros(2)
 
 specific_sigmas = np.clip(
     np.abs(np.random.normal(0, 0.01, size=len(elements_measured))),
@@ -47,7 +48,7 @@ specific_sigmas = np.clip(
 
 cluster_dispersion = 0.01
 
-magnitude_of_factor_load = 1
+magnitude_of_factor_load = 0.25
 # --------------------- #
 
 

### 2017-10-02 articles/chemical-tagging-gmm/experiments/slf-clustering-fake-ocs/toy_experiment.py

@@ -35,7 +35,7 @@ K_clusters_total = len(set(catalog["cluster_id"]))
 
 
 
-mml_mod = MMLMixtureModel(num_components=2)
+mml_mod = MMLMixtureModel(num_components=2, initialization_method="random")
 mml_mod.fit(X)
 
 raise a

### 2017-10-02 snob/mixture.py

@@ -219,7 +219,7 @@ class GaussianMixture(object):
             If there is a shape mis-match between the input arrays.
         """
 
-        if None in (means, covariances, weights):
+        if means is None or covariances is None or weights is None:
             for parameter_name in self.parameter_names:
                 setattr(self, "_{}".format(parameter_name), None)
                 return False
@@ -264,6 +264,7 @@ class GaussianMixture(object):
         # Only initialize if we don't have parameters already.
         self._soft_initialize(y, **kwargs)
         self.converged = False
+
         prev_log_prob_norm = -np.inf
 
         for iteration in range(self.max_iter):
@@ -271,6 +272,10 @@ class GaussianMixture(object):
             # Expectation.
             log_prob_norm, log_responsibility = self._expectation(y)
 
+
+            if iteration > 0:
+                raise a
+                
             # Maximization.
             self._maximization(y, np.exp(log_responsibility))
 
@@ -318,7 +323,6 @@ class GaussianMixture(object):
             # Ignore underflow errors.
             log_resp = weighted_log_prob - log_prob_norm[:, np.newaxis]
 
-        raise a
         return (log_prob_norm, log_resp)
 
 

### 2017-10-02 snob/slf_mixture/base.py

@@ -164,6 +164,7 @@ class BaseMixtureModel(object):
         responsibility, log_likelihood = self._expectation(data) 
 
         results = [log_likelihood.sum()]
+        logger.debug("fit: initial ll: {}".format(results[0]))
 
         for iteration in range(self.max_em_iterations):
 
@@ -174,15 +175,16 @@ class BaseMixtureModel(object):
 
             # Compute the e-step of cycle 2
             responsibility, _ = self._expectation(data)
-            assert _.sum() > log_likelihood.sum()
+            logger.debug("fit intermediate ll: {}".format(_.sum()))
+
 
             # Do the M-step of cycle 2
             self._aecm_step_2(data, responsibility)
 
             # Re-compute the expectation,
-            responsibility, log_likelihood = self._expectation(data)
-            #assert log_likelihood_.sum() > log_likelihood.sum()
-            #log_likelihood = log_likelihood_
+            responsibility, log_likelihood_ = self._expectation(data)
+            assert log_likelihood_.sum() > log_likelihood.sum()
+            log_likelihood = log_likelihood_
             results.append(log_likelihood.sum())
 
             # Check for convergence.
@@ -218,8 +220,6 @@ class BaseMixtureModel(object):
         with np.errstate(under="ignore"):
             # Ignore underflow errors.
             log_resp = weighted_log_prob - log_likelihood[:, np.newaxis]
-
-        raise a
 
         responsibility = np.exp(log_resp).T
         
@@ -276,12 +276,21 @@ def _estimate_log_latent_factor_prob(data, factor_loads, cluster_factor_scores,
     # L = number of latent factors
     # K = number of clusters.
 
+    #import matplotlib.pyplot as plt
+    #fig, ax = plt.subplots()
+    #ax.scatter(data.T[0], data.T[1], facecolor="#666666", zorder=-1)
+    #colors = "br"
+
+    # TODO consider multiplication of inv_specific_variances instead
+
     log_prob = np.empty((N, K))
     for k, factor_scores in enumerate(cluster_factor_scores.T):
         squared_diff = (data - mean - np.dot(factor_scores, factor_loads))**2
         log_prob[:, k] = np.sum(squared_diff / specific_variances, axis=1)
 
-    raise a
+        #foo = mean + np.dot(factor_scores, factor_loads)
+        #ax.scatter(foo.T[0], foo.T[1], facecolor=colors[k], alpha=0.5)
+
 
[…]

### 2017-10-02 snob/slf_mixture/mml.py

@@ -37,32 +37,22 @@ class MMLMixtureModel(BaseMixtureModel):
         # single mean for the entire sample, not the cluster means!
         # TODO REVISIT TERMINOLOGY
         means = np.mean(data, axis=0)
-        means = np.array([-0.25640511, -0.04716052])
-        logger.warn("DANGER WILL ROBINSON")
+        
         w = data - means
 
-        U, S, V = np.linalg.svd(np.corrcoef(w.T))
+        # Set the initial factor scores by their location along the principal
+        # eigenvetor, scaled by the square root
+        U, S, V = np.linalg.svd(w.T)
+        factor_scores = (V[0] * S[0]**0.5).reshape((N, -1))
+        factor_loads = np.mean(w/factor_scores, axis=0).reshape((-1, D))
+        specific_variances = np.var(
+            w - np.dot(factor_scores, factor_loads), axis=0)
 
-        # Set the initial values for b as the principal eigenvector, scaled by
-        # the square-root of the principal eigenvalue
-        b = V[0] * S[0]**0.5
-
-        # b = factor_load/np.sqrt(specific_variances)
-        b_sq = np.sum(b**2)
-        factor_scores = (np.dot(data, b.T) \
-                      * (1 - D/(N - 1) * b_sq)/(1 + b_sq)).reshape((-1, 1))
-
-        specific_variances = np.diag(np.dot(w.T, w)) / ((b*b + 1.0) * (N - 1))
-        factor_loads = (np.sqrt(specific_variances) * b).reshape((1, -1))
-
-        #y_nlf = data - np.dot(factor_scores, factor_loads)
-
-        # We want to do the clustering on the factor scores.
-        
+        # Do initial clustering on the factor scores.
         responsibility = np.zeros((N, self.num_components))
-            
+           
         if self.initialization_method == "kmeans++":
-            # Assign cluster memberships to the transformed data.
+            # Assign cluster memberships using k-means++.
             kmeans = cluster.KMeans(
                 n_clusters=self.num_components, n_init=1, **kwargs)
             assignments = kmeans.fit(factor_scores).labels_
@@ -81,9 +71,22 @@ class MMLMixtureModel(BaseMixtureModel):
         # TODO: Revisit this whether this should be effective_membership > 1
         cluster_factor_scores = np.dot(factor_scores.T, responsibility) \
                               / effective_membership
-        
[…]

### 2017-10-02 .ci/travis.sh
get travis to build all pdfs (hopefully)

@@ -13,7 +13,7 @@ if [ -n "$GITHUB_API_KEY" ]; then
   cd $TRAVIS_BUILD_DIR
   git checkout --orphan $TRAVIS_BRANCH-pdf
   git rm -rf .
-  git add -f articles/*/ms.pdf
+  git add -f articles/chemical-tagging-gmm/ms.pdf articles/gmm-search/ms.pdf
   git -c user.name='travis' -c user.email='travis' commit -m "Compiling PDFs"
   git push -q -f https://$GITHUB_USER:$GITHUB_API_KEY@github.com/$TRAVIS_REPO_SLUG $TRAVIS_BRANCH-pdf
 fi

### 2017-10-02 .ci/build-papers.sh
Change Travis to use tectonic instead of texlive

@@ -1,0 +1,25 @@
+#!/bin/bash -x
+
+if [[ $BUILD_PAPERS == false ]]; then
+  return
+fi
+
+# Build the papers
+declare -a folders=("chemical-tagging-gmm" "gmm-search")
+for folder in "${folders[@]}"
+do
+   cd $TRAVIS_BUILD_DIR
+   cd "articles/$folder"
+   make
+done
+
+
+# Push to GitHub
+if [ -n "$GITHUB_API_KEY" ]; then
+  cd $TRAVIS_BUILD_DIR
+  git checkout --orphan $TRAVIS_BRANCH-pdf
+  git rm -rf .
+  git add -f paper/ms.pdf
+  git -c user.name='travis' -c user.email='travis' commit -m "building papers"
+  git push -q -f https://$GITHUB_USER:$GITHUB_API_KEY@github.com/$TRAVIS_REPO_SLUG $TRAVIS_BRANCH-pdf
+fi

### 2017-10-02 .ci/setup-texlive.sh

@@ -1,12 +1,0 @@
-#!/bin/bash -x
-
-# Download and extract texlive installer
-wget http://mirror.ctan.org/systems/texlive/tlnet/install-tl-unx.tar.gz -O install-tl.tar.gz
-mkdir -p install-tl
-tar -xf install-tl.tar.gz -C install-tl --strip-components 1
-
-# Install only the necessary packages
-mkdir -p texlive
-TEXLIVE_INSTALL_PREFIX=`pwd`/texlive install-tl/install-tl -profile .ci/texlive.profile
-export PATH=`pwd`/texlive/bin/x86_64-linux:$PATH
-tlmgr install lineno revtex textcase epsf xcolor ulem microtype multirow helvetic

### 2017-10-02 .ci/travis.sh

@@ -1,19 +1,39 @@
 #!/bin/bash -x
 
-# Build the paper(s)
-cd $TRAVIS_BUILD_DIR/articles/chemical-tagging-gmm
-make
-#cd $TRAVIS_BUILD_DIR/articles/nips-2017
-#make
-cd $TRAVIS_BUILD_DIR/articles/gmm-search
-make
-
-# Push to GitHub
-if [ -n "$GITHUB_API_KEY" ]; then
-  cd $TRAVIS_BUILD_DIR
-  git checkout --orphan $TRAVIS_BRANCH-pdf
-  git rm -rf .
-  git add -f articles/*/ms.pdf
-  git -c user.name='travis' -c user.email='travis' commit -m "Compiling PDFs"
-  git push -q -f https://$GITHUB_USER:$GITHUB_API_KEY@github.com/$TRAVIS_REPO_SLUG $TRAVIS_BRANCH-pdf
+# If building the paper, do that here
+if [[ $TEST_LANG == paper ]]
+then
+  if git diff --name-only $TRAVIS_COMMIT_RANGE | grep 'articles/'
+  then
+    echo "Building the paper..."
+    export BUILD_PAPERS=true
+    source "$( dirname "${BASH_SOURCE[0]}" )"/setup-tectonic.sh
+    return
+  fi
+  export BUILD_PAPERS=false
+  return
 fi
+
+# If we make it here then we're testing Python
+
+# http://conda.pydata.org/docs/travis.html#the-travis-yml-file
+if [[ "$TRAVIS_OS_NAME" == "osx" ]]; then
+  wget https://repo.continuum.io/miniconda/Miniconda3-latest-MacOSX-x86_64.sh -O miniconda.sh;
+else
+  wget https://repo.continuum.io/miniconda/Miniconda3-latest-Linux-x86_64.sh -O miniconda.sh;
+fi
+bash miniconda.sh -b -p $HOME/miniconda
+export PATH="$HOME/miniconda/bin:$PATH"
+
+# Conda Python
+hash -r
+conda config --set always_yes yes --set changeps1 no
+conda update -q conda
+conda info -a
+conda create --yes -n test python=$PYTHON_VERSION
+source activate test
+conda install -c conda-forge numpy=$NUMPY_VERSION setuptools pytest pytest-cov pip
+pip install coveralls
+
+# Install code.
+python setup.py develop

### 2017-10-02 .ci/build-papers.sh
fix bug in travis script

@@ -19,7 +19,10 @@ if [ -n "$GITHUB_API_KEY" ]; then
   cd $TRAVIS_BUILD_DIR
   git checkout --orphan $TRAVIS_BRANCH-pdf
   git rm -rf .
-  git add -f paper/ms.pdf
+  for folder in "${folders[@]}"
+  do
+      git add -f "articles/$folder/ms.pdf"
+  done
   git -c user.name='travis' -c user.email='travis' commit -m "building papers"
   git push -q -f https://$GITHUB_USER:$GITHUB_API_KEY@github.com/$TRAVIS_REPO_SLUG $TRAVIS_BRANCH-pdf
 fi

### 2017-10-02 .ci/build-papers.sh
bug in travis build

@@ -10,7 +10,7 @@ for folder in "${folders[@]}"
 do
    cd $TRAVIS_BUILD_DIR
    cd "articles/$folder"
-   make
+   tectonic ms.tex --print
 done
 
 

### 2017-10-02 .ci/build-papers.sh
have travis control vc.tex for us if using tectonic

@@ -10,6 +10,16 @@ for folder in "${folders[@]}"
 do
    cd $TRAVIS_BUILD_DIR
    cd "articles/$folder"
+
+   # Create vc.tex file with \githash, \gitdate, and \gitauthor
+   echo '\newcommand{\giturl}{\url{%' >> vc.tex
+   git config --get remote.origin.url | sed "s/\.git$$//" | sed "s/^ssh:\/\///g" | sed "s/^git\@github\.com:/https:\/\/www\.github\.com\//" >> vc.tex
+   echo '}}' >> vc.tex
+   git log -1 --date=short --format="format:\\newcommand{\\githash}{%h}\\newcommand{\\gitdate}{%ad}\\newcommand{\\gitauthor}{%an}" >> vc.tex
+   # Safely replace newline character if it exists
+   cat vc.tex | sed "s/^ewcommand/\\\newcommand/" > vc_safe.tex
+   mv vc_safe.tex vc.tex
+
    tectonic ms.tex --print
 done
 

### 2017-10-10 articles/gmm-search/ms.tex
Expanding the MML objective function in search paper [wip]

@@
 %\usepackage{boldsymbol}
+\usepackage{color} % for \todo command
 \usepackage{verbatim} % for \begin{comment}'ing out sections.
@@
 	\newcommand{\giturl}{UNKNOWN}}
+
+\newcommand{\todo}[1]{\textcolor{red}{#1}}
 
@@
 \def\likelihood{\mathcal{L}}
+\newcommand{\fisher}[1]{\mathcal{F}\left(#1\right)}
+\newcommand{\detfisher}[1]{\left|\fisher{#1}\right|}
+\newcommand{\prior}[1]{p\left(#1\right)}
 
@@
 \section{MML}
+% \section{Minimum Message Length (MML)}
 A complex model is not an optimal one, unless its complexity is
@@
 \section{Message Length}
-There are $N$ data points each with $D$ dimensions, which are to be modelled
-by $K$ mixture of $D$-dimensional gaussian distributions, each with a relative
-weighting $\weight_k$ such that $\sum_{k=1}^{K}\weight_k = 1$. 
+
+The MML principle requires that we encode everything required to reconstruct 
+the message, including our prior beliefs on the model parameters. Consider 
+that we have $N$ data points each with $D$ dimensions which are to be 
+modelled by a finite number of Gaussian mixtures. Therefore the message for 
+a finite number of Gaussian mixtures must encode:
+
+\begin{enumerate}
+  \item The number of Gaussian mixtures, $K$.
+  \item The relative weights $\weights$ of the $K - 1$ Gaussian mixtures. Only
+        $K - 1$ weights must be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
+  \item The component parameters $\vecmean$ and $\veccov$ for all $K$ Gaussian
+        mixtures.
+  \item The data, given the model parameters.
+  \item The lattice quantisation constant $\kappa(Q)$ for the number of model
+        parameters $Q$.
+\end{enumerate}
+
+We provide the message length of each component in turn before outlining our
+full objective function. We assume a prior on the number of mixtures $K$ of
+$\prior{K} \propto 2^{-K}$ \todo{\cite{someone}} such that the length of the 
+optimal lossless message to encode $K$ is 
+\begin{equation}
+	I(K) = -\log{\prior{K}} = K\log{2} + \textrm{constant}.
+\end{equation}
+
+We assume a uniform prior on the individual weights $\weight_{k}$. The weights
+$\weights$ are drawn from a multinomial distribution such that they can be 
+optimally encoded in a message with length\todo{\cite{someone}}:
+\begin{equation}
+  I(\weights) 
+    = \frac{K - 1}{2}\log{N} 
+    - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k 
+    - \log{\Gamma{\left(K\right)}} \quad .
+\end{equation}
+
+In order to properly encode the means $\vecmean$ and covariance matrices
+$\veccov$ for all $K$ mixtures, we must encode both our prior belief on 
+those parameters and the determinant of the expected Fisher information 
+matrix. For the $k$-th mixture the joint message length is
+\begin{eqnarray}
+  I(\vecmean_k,\veccov_k) &=& -\log{\left(\frac{\prior{{\vecmean_k,\veccov_k}}}{\sqrt{\detfisher{{\vecmean_k,\veccov_k}}}}\right)} \nonumber \\ 
+  I(\vecmean_k,\veccov_k) &=& -\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\log{\detfisher{{\vecmean_k,\veccov_k}}}
+\end{eqnarray}
+
+\noindent{}and for all mixtures:
+\begin{equation}
+  I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \quad .
+  \label{eq:I_component_params}
+\end{equation}
+
+We use an improper uniform prior on the mean $\vecmean$ for all mixtures such
+that 
+	$\mathcal{U}(\vecmean) = (-\infty, +\infty)$.
+However,  the MML principle \todo{demands} proper priors. In practice we make 
+this prior proper by assuming the bounds on $\prior{\vecmean}$ are very large, but 
+we do so without actually specifying the value of the bounds because they add 
+only a constant value to the total message length\todo{\cite{someone}}.
+We adopt a conjugate inverted Wishart prior for the covariance matrix of an
+individual mixture \cite[e.g., Section 5.2.3. of ][]{Schafer_1997}:
+\begin{equation}
+  \prior{{\vecmean_k, \veccov_k}} \propto |\veccov_k|^{\frac{D+1}{2}} \quad .
+  \label{eq:covariance-prior}
+\end{equation}
+
+% Casey: Reword following paragraph because it is clumsy.
+Following Eq. \ref{eq:I_component_params}, we require the logarithm of the
+determinant of the expected Fisher information matrix for $\vecmean_k$ and
+$\veccov_k$ in order to encode the full message length. We approximate the
+determinant of the Fisher information $\detfisher{{\vecmean_k, \veccov_k}}$ 
+as the product of $\detfisher{\vecmean_k}$ and $\detfisher{\veccov_k}$ 
+\cite{Oliver_1996,Roberts_1998}. In order to derive $\detfisher{\vecmean_k}$
+we take the second derivative of the log-likelihood function 
+$\likelihood(\data|\vecmean,\veccov)$, which yields:
+\begin{equation}
+  \detfisher{\vecmean_k} = (N\weight_k)^{D}|\veccov_k|^{-1} \quad .
+\end{equation}
+We make use of an analytical expression for $\detfisher{\veccov_k}$
+\cite{Dwyer_1967,Magnus_1988}:
+\begin{equation}
+  \detfisher{\veccov_k} = (N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \quad .
+\end{equation}
+Combining these expressions gives our approximation for the determinant of
+the expected Fisher information $\detfisher{{\vecmean_k,\veccov_k}}$:
+\begin{eqnarray}
+  \detfisher{{\vecmean_k,\veccov_k}} & \approx & \detfisher{\vecmean_k} \cdot \detfisher{\veccov_k} \nonumber \\
+  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^{D}|\veccov_k|^{-1}(N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
+  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
+  \label{eq:mixture-component-det-fisher}
+\end{eqnarray}
+
+Substituting Eqs. \ref{eq:covariance-prior} and \ref{eq:mixture-component-det-fisher} 
+into Eq. \ref{eq:I_component_params} gives the message length to encode the component
+parameters $\vecmean$ and $\veccov$ for all $K$ mixtures:
+
+\begin{eqnarray}
+  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \nonumber \\
+  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{|\veccov_k|^{\frac{D + 1}{2}}}
+	+ \frac{1}{2}\sum_{k=1}^{K}\log{\left[(N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right]} \nonumber \\
+  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
+	-\frac{KD}{2}\log{2} \nonumber \\
+  && -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
+  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{2D+3}{2}\sum_{k=1}^{K}\log{|\veccov_k|} - \frac{KD}{2}\log{2} 
+\end{eqnarray}
+
+If we assume that the data have homoskedastic noise properties, then the 
+precision of each measurement $\mathcal{E}$ relates the probability of a
+datum $p(\datum_n)$ can be related to the probability density of
+the datum, given the model 
+$p(\datum_n) = \mathcal{E}^{D}p(y_i|\mathcal{M})$.
+In this work we adopt $\mathcal{E} = 0.01$, but we note that the value of
+$\mathcal{E}$ has no effect on our inferences.  The message length of encoding
+a datum given the model is,
+\begin{equation}
+  I(\data_n) = -\log{p(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
+\end{equation}
+
+\noindent{}and for the entire data:
+\begin{eqnarray}
+  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .
+\end{eqnarray}
+
+The final part of the message is to encode the lattice quantisation constant,
+which arises from the approximation to the (so-called) strict MML, where
+parameters are quantised into intervals in high dimensional space in order to
+be losslessly encoded.  We use an approximation of the lattice constant
+\citep[see Sections 5.1.12 and 3.3.4 of ][]{Wallace_2005} such that,
+\begin{equation}
+  \log\kappa(Q) = \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1 \quad ,
+\end{equation}
+
+\noindent{}where $Q$ is the total number of free parameters in the model. Each
+Gaussian component in the mixture requires $D$ parameters to specify $\vecmean$,
+and a covariance matrix with off-diagonal terms requires $\frac{1}{2}D(D+1)$ 
+parameters. Because only $K - 1$ weights need to be encoded, the total number of free
+parameters is
+\begin{equation}
+	Q = K\left[\frac{D(D+3)}{2} + 1\right] - 1 \quad .
+	\label{eq:number-of-parameters}
+\end{equation}
+
+%  Here
+
+ Each $k$-th mixture is a $D$-dimensional Gaussian that contributes a relative weight $\weight_k$ such that $\sum_{k=1}^{K}\weight_k = 1$.
+We will assume that the data have homoscedastic noise properties.
+
 The data have the same error value, $y_{err}$, in all $D$ dimensions, for all $N$ observations.
 
-The full expression for the message length is
-given by,
+The message length for the data is given by:
+
+Here we introduce the individual components that contribute to the message length.
+
+Combining these expressions, the complete message length for $K$ mixtures is given by
 
@@
 \section{The Heuristic}
+% \section{The Heuristic - Method of Moments}
 

### 2017-10-11 ms.tex
Update on Overleaf.

--- 2017-10-10 articles/gmm-search/ms.tex
@@
 
+\section{Related Work in Inferring the Number of Mixture Components}
+
+Akaike information criterion (AIC) ~\cite{akaike1974} chooses a model that minimizes the Kullback-Leibler distance between the model and the truth, defined as:
+
+AIC = -2 (ln ( likelihood )) + 2 K
+
+where likelihood is the probability of the data given a model and K is the number of free parameters in the model.
+
+BIC
+
 \section{The k-means++ algorithm}
@@
 
-And, of course, one could change K from 2 to something bigger.
-
-There should still be some use of randomness in the search.
 \bibliographystyle{elsarticle-num}
